---
title: "SEA-AD omics exploration: processed multiregion snRNA-seq for a donor x ROI molecular-state block"
subtitle: "What exists, what is local, how it joins to the QNP cohort, and which T variables are defensible"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    df-print: default
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose and scope

`00_seaad_exploration.qmd` established a candidate donor x brain-region **QNP** block:

| block | variable | source column |
|---|---|---|
| P | Ab | `percent 6e10 positive area` |
| P | pTau | `percent AT8 positive area` |
| H | NeuN | `number of NeuN positive cells per area` |
| H | GFAP | `percent GFAP positive area` |

This notebook asks whether a **molecular-state block $T_{ir}$** can sit beside it on the same
donor x ROI grid, for the candidate ROI set **DFC, HIP, MEC, MTG, STG, V1C**. It answers seven
questions:

1. What processed multiregion RNA objects are locally available?
2. What is the donor x ROI omics coverage for DFC / HIP / MEC / MTG / STG / V1C?
3. What donor-aware pseudobulk or aggregate molecular representations already exist?
4. What published molecular-state representations are available for microglia, astrocytes,
   neurons, and oligodendrocytes/OPCs?
5. Which candidate variables represent molecular **state** versus **composition**?
6. How many donor x ROI observations have both QNP P/H and usable omics?
7. What are the smallest defensible candidate definitions of $T$?

Deliberately **not** done here: differential expression, gene-level feature selection, pathway
enrichment, reclustering, scVI/scANVI, regression, morphology, MERFISH/Xenium, or any change to
the QNP definitions fixed in notebook 00. Nothing below treats a nucleus, or a library, as an
independent biological observation.

```{r}
#| label: setup

my_packages <- c("tidyverse", "readxl", "knitr", "jsonlite")
invisible(lapply(my_packages, library, character.only = TRUE))

find_root <- function(p = getwd()) {
  while (!dir.exists(file.path(p, ".git"))) {
    if (identical(dirname(p), p)) stop("project root not found")
    p <- dirname(p)
  }
  p
}
PROJECT_ROOT <- find_root()
source(file.path(PROJECT_ROOT, "R", "seaad_explore.R"))

paths      <- seaad_paths(PROJECT_ROOT)
omics      <- seaad_omics_paths(PROJECT_ROOT)

show_table <- function(x, caption = NULL, digits = 4) {
  kable(x, caption = caption, digits = digits, format.args = list(big.mark = ""))
}

# The candidate multimodal ROI set. Still candidate, not frozen.
ROI <- c("DFC", "HIP", "MEC", "MTG", "STG", "V1C")
```

---

## 0. What this notebook reads, and what it refused to read

The processed SEA-AD single-nucleus data is **not** in `data/raw/sea-ad/`. That directory holds
3.7 MB of donor-level spreadsheets and PDFs and nothing else. The omics release lives in the AWS
Open Data bucket `sea-ad-single-cell-profiling`.

`scripts/fetch_seaad_omics_inventory.py` therefore does two separate things, and the distinction
matters for every claim below:

- it **lists** object keys and sizes (free, no download) into an S3 manifest;
- it **downloads** only files under 8 MB, plus one 4.2 MB pseudobulk object fetched by name as a
  structure probe.

No multi-gigabyte AnnData object was fetched. Where a fact could only come from such an object,
it is reported as unverified and the object is named with its size.

```{r}
#| label: provenance

prov <- fromJSON(omics$provenance)

show_table(
  tibble(
    field = c("bucket", "retrieved (UTC)", "keys listed", "keys excluded",
              "single-download size limit", "pseudobulk structure probe"),
    value = c(prov$bucket,
              prov$retrieved_utc,
              format(prov$n_keys, big.mark = ","),
              paste(sprintf("%s %s", trimws(format(unlist(prov$n_keys_excluded), big.mark = ",")),
                            names(prov$n_keys_excluded)), collapse = "; "),
              paste0(prov$max_bytes / 1e6, " MB"),
              basename(prov$pseudobulk_probe))
  ),
  "Acquisition provenance for this notebook"
)
```

`previous_objects/` is excluded because the 2026-06-22 multiregion release supersedes it by its
own README; `scANVI_models/` and `scVI_models/` are excluded because section 7 rules integrated
latent representations out of scope as quantitative measures.

```{r}
#| label: local-files

local_inventory <- tibble(
  role = c("S3 key manifest (derived)", "Bucket README", "Multiregion 2026 README",
           "Cell-type taxonomy + region presence", "Published donor x region CPS",
           "Published compositional-model summary",
           "Supertype abundances by library (derived from 2 h5ad)",
           "Pseudobulk structure probe .obs (derived from 1 h5ad)"),
  path = c(omics$s3_manifest, omics$bucket_readme, omics$multiregion_readme,
           omics$taxonomy, omics$cps_by_region, omics$pertpy_summary,
           omics$abundances, omics$pseudobulk_probe)
) %>%
  mutate(file = basename(path), exists = file.exists(path),
         size_kb = round(file.size(path) / 1024)) %>%
  select(role, file, exists, size_kb)

show_table(local_inventory, "Everything this notebook reads. Total on disk is under 8 MB.")
```

---

## 1. Inventory of processed multiregion RNA objects

### 1.1 The release, as the publisher describes it

```{r}
#| label: readme
#| results: asis

mr <- readLines(omics$multiregion_readme, warn = FALSE)
cat(paste0("> ", mr[seq_len(12)], collapse = "\n"), "\n")
```

### 1.2 Object families in the release

```{r}
#| label: manifest

manifest <- read_seaad_s3_manifest(omics$s3_manifest)

show_table(
  manifest %>%
    group_by(family) %>%
    summarise(n_objects = n(),
              total_gb = round(sum(size_mb) / 1000, 1),
              largest_gb = round(max(size_mb) / 1000, 1),
              .groups = "drop") %>%
    arrange(desc(total_gb)),
  "Object families under the listed prefixes, by size"
)
```

### 1.3 The region folder is not the region label

```{r}
#| label: folder-vs-region

show_table(
  manifest %>%
    filter(assay == "snRNAseq", !is.na(region)) %>%
    distinct(folder, region) %>%
    arrange(folder),
  "Bucket folder vs the region label written into the file name"
)
```

**`PFC/` holds the DFC data.** Its objects are named `SEAAD_DFC_RNAseq_*`; the folder keeps the
older DLPFC/PFC label. `DFC/` and `DLPFC/` contain only a changelog and a forwarding note. Every
table below keys on the label parsed from the *file name*, never the folder — see
`SEAAD_FOLDER_TO_REGION` in `R/seaad_explore.R`.

### 1.4 Region-level objects: what exists, and what it would cost

```{r}
#| label: region-objects

region_objects <- manifest %>%
  filter(family %in% c("region-level AnnData", "per-nucleus metadata table",
                       "per-nucleus annotation table"),
         !is.na(region), assay == "snRNAseq") %>%
  transmute(region, nuclei_set, file, size_gb = round(size_mb / 1000, 1),
            release_date, local = file.exists(file.path(omics$raw_dir, file)))

show_table(
  region_objects %>% filter(region %in% ROI) %>% arrange(region, desc(size_gb)),
  "Region-level snRNA-seq objects for the candidate ROI set. None is local."
)
```

```{r}
#| label: region-object-cost

show_table(
  region_objects %>%
    filter(region %in% ROI, nuclei_set == "final-nuclei",
           str_detect(file, "\\.h5ad$")) %>%
    summarise(n_objects = n(), total_gb = round(sum(size_gb), 1)),
  "Cost of the obvious-but-wrong route: the six final-nuclei region objects"
)
```

That is the route this notebook does **not** take. Section 3 shows that SEA-AD already publishes
donor-aware aggregates two orders of magnitude smaller.

### 1.5 snATAC-seq and multiome

```{r}
#| label: atac

show_table(
  manifest %>%
    filter(assay == "snATACseq", !str_detect(file, "Changelog")) %>%
    transmute(file, size_gb = round(size_mb / 1000, 2), release_date),
  "All snATAC-seq objects in the listed prefixes"
)
```

ATAC exists for MTG only, from the 2024 release. It is out of scope for a six-ROI block and is
not considered further.

### 1.6 The inventory, in the requested schema

```{r}
#| label: object-inventory

n_nuclei_by_region <- read_csv(omics$abundances, show_col_types = FALSE) %>%
  group_by(brain_region) %>%
  summarise(n_nuclei = sum(n_nuclei), .groups = "drop")

donor_objects <- seaad_donor_region_objects(manifest)

inventory <- tibble::tribble(
  ~object, ~release, ~assay, ~regions, ~row_grain, ~donor_id_field, ~region_field,
  "SEAAD_<REGION>_RNAseq_final-nuclei.2026-06-22.h5ad",
    "Multiregion 2026-06-22", "snRNAseq", "one per region (10)", "nucleus",
    "obs$`Donor ID`", "implied by object; obs$`Brain Region` in aggregates",
  "<DONOR>_SEAAD_<REGION>_RNAseq_final-nuclei.2026-06-22.h5ad",
    "Multiregion 2026-06-22", "snRNAseq", "10 regions x donors", "nucleus",
    "file name + obs$`Donor ID`", "file name",
  "SEAAD_<SUBCLASS>_..._final-nuclei.2026-06-22.h5ad (subclass_objects/)",
    "Multiregion 2026-06-22", "snRNAseq", "all 10 pooled", "nucleus",
    "obs$`Donor ID`", "obs$`Brain Region`",
  "SEAAD_<SUBCLASS>_..._pseudobulked.2026-06-22.h5ad (pseudobulk_objects/)",
    "Multiregion 2026-06-22", "snRNAseq", "all 10 pooled", "donor x region x supertype",
    "obs$`Donor ID`", "obs$`Brain Region`",
  "SEAAD_..._all-nuclei_metadata.2026-06-22.csv",
    "Multiregion 2026-06-22", "snRNAseq", "all 10 pooled", "nucleus",
    "column `Donor ID`", "column `Brain Region`",
  "Global_and_Local_CPS.20260105.csv",
    "Multiregion 2026 model outputs", "derived from neuropathology", "9 regions",
    "donor x region", "column `Donor ID`", "column `Brain Region`",
  "<CLASS>_Supertype_abundances.h5ad (pertpy objects/)",
    "Multiregion 2026 model outputs", "snRNAseq", "all 10 pooled", "library x supertype",
    "ABSENT", "obs$`Brain Region`"
)

show_table(inventory, "Processed RNA objects and their identifier fields")
```

```{r}
#| label: object-inventory-2

show_table(
  tibble::tribble(
    ~field, ~value,
    "taxonomy version",
      "SEA-AD MTG/PFC taxonomy, multiregion expansion of 2026-06-22; no version string is published",
    "n supertypes", as.character(nrow(seaad_taxonomy(omics$taxonomy))),
    "cell-class / subclass fields", "obs$Class, obs$Subclass, obs$Supertype",
    "published disease-state fields",
      "obs carries Thal / Braak / CERAD / ADNC / LATE / Cognitive Status (donor-level, not cell state)",
    "expression in region and subclass objects",
      "X = log-normalised UMI per 10,000; layers = raw UMI counts (per release README)",
    "expression in pseudobulk objects",
      "X = summed raw UMI counts; no layers (verified in 3.2)",
    "pseudobulk / aggregate availability",
      "29 subclass pseudobulk objects, 4-460 MB each; 2 supertype-abundance objects, <1.5 MB each"
  ),
  "Taxonomy, annotation and expression fields"
)
```

```{r}
#| label: object-inventory-counts

show_table(
  donor_objects %>%
    count(brain_region, name = "n_donors") %>%
    left_join(n_nuclei_by_region, by = "brain_region") %>%
    left_join(manifest %>%
                filter(family == "region-level AnnData", assay == "snRNAseq",
                       nuclei_set == "final-nuclei",
                       str_detect(file, "\\.h5ad$"), !is.na(region)) %>%
                transmute(brain_region = region, region_object_gb = round(size_mb / 1000, 1)),
              by = "brain_region") %>%
    arrange(desc(n_donors)),
  "n donors and n nuclei per region. Nuclei counts are the final-nuclei totals in the released compositional-model input (section 3.3)."
)
```

### 1.7 What is absent locally

Everything above. In full: **no H5AD/AnnData object, no cell metadata table, no taxonomy
annotation table beyond the 26 kB colour/order file, no pseudobulk object, and no subclass
aggregate is present under `data/raw/sea-ad/`.** What this notebook has instead is the key
manifest plus the five sub-megabyte published summaries listed in section 0. The specific missing
objects that a later step would need are named with sizes in sections 3.1 and 10.

---

## 2. Donor x ROI coverage

### 2.1 Do the identifiers need a crosswalk?

Before building any join, the existing project crosswalk code was inspected:
`src/seaad_mtg_docker_package/seaad_mtg_docker/seaad_mtg/crosswalk.py`,
`seaad_mtg_build_canonical_crosswalk.py`, and `scripts/seaad_mtg_h5ad_crosswalk_probe.py`
(commit `2135026`, "Add CLI and crosswalk functionality for SEA-AD MTG pipeline"). That code
solves a different problem — linking MERFISH **sections** to donors via `Specimen Barcode` — and
its donor-side logic is a preferred-column lookup (`Donor ID`, `donor_id`, ...) backed by the
exact pattern `^H\d{2}\.\d{2}\.\d{3}$`. There is no fuzzy matching in it and none is introduced
here.

```{r}
#| label: crosswalk-check

donor_pattern <- "^H\\d{2}\\.\\d{2}\\.\\d{3}$"   # as in crosswalk.py::DONOR_REGEX

qnp        <- read_qnp_workbook(paths$qnp_2026)
qnp_ph     <- qnp_primary_wide(qnp_primary_long(qnp))
probe_obs  <- read_csv(omics$pseudobulk_probe, show_col_types = FALSE)
# The published CSV carries an unnamed pandas index column; it is dropped here.
cps_region <- read_csv(omics$cps_by_region, show_col_types = FALSE,
                       name_repair = ~ ifelse(.x == "", "row_index", .x)) %>%
  select(donor_id = `Donor ID`, brain_region = `Brain Region`, starts_with("CPS_"))

show_table(
  tibble::tribble(
    ~source, ~donor_field, ~region_field, ~n_donors, ~all_ids_match_pattern,
    "QNP workbook (notebook 00)", "Donor ID", "brain region",
      n_distinct(qnp_ph$donor_id), all(str_detect(qnp_ph$donor_id, donor_pattern)),
    "per-donor omics objects (file names)", "leading token of file name", "file name",
      n_distinct(donor_objects$donor_id), all(str_detect(donor_objects$donor_id, donor_pattern)),
    "pseudobulk .obs (probe)", "Donor ID", "Brain Region",
      n_distinct(probe_obs$`Donor ID`), all(str_detect(probe_obs$`Donor ID`, donor_pattern)),
    "published donor x region CPS", "Donor ID", "Brain Region",
      n_distinct(cps_region$donor_id), all(str_detect(cps_region$donor_id, donor_pattern))
  ),
  "Donor identifiers across the omics and neuropathology sources"
)

stopifnot(
  "every omics donor must already be a QNP donor" =
    length(setdiff(donor_objects$donor_id, qnp_ph$donor_id)) == 0,
  "region labels must agree without remapping" =
    length(setdiff(intersect(donor_objects$brain_region, ROI), qnp_ph$brain_region)) == 0
)

cat("donors in omics but not QNP:",
    length(setdiff(donor_objects$donor_id, qnp_ph$donor_id)), "\n")
cat("ROI labels needing translation:",
    length(setdiff(ROI, intersect(donor_objects$brain_region, qnp_ph$brain_region))), "\n")
```

**No crosswalk was necessary.** The processed omics objects carry explicit SEA-AD donor IDs in
the same `H##.##.###` form as the QNP workbook, and the region labels `DFC / HIP / MEC / MTG /
STG / V1C` are identical strings on both sides. The join is an exact equality join on
`(donor_id, brain_region)`. The one naming hazard — folder `PFC/` vs label `DFC` — is handled in
section 1.3 and is a *folder*, not an identifier.

The single exception is the pertpy abundance objects, which carry `library_prep` and **no**
`Donor ID`; that gap is stated explicitly in section 2.4 and never bridged by inference.

### 2.2 The donor x ROI omics grid

```{r}
#| label: donor-roi-grid

roi_objects <- donor_objects %>% filter(brain_region %in% ROI)

show_table(
  roi_objects %>%
    count(brain_region, name = "n_donors") %>%
    left_join(n_nuclei_by_region, by = "brain_region") %>%
    mutate(nuclei_per_donor = round(n_nuclei / n_donors)) %>%
    arrange(desc(n_donors)),
  "Donors with a released per-donor object, by ROI"
)
```

```{r}
#| label: readme-discrepancy

readme_claim <- c("MTG" = 84L, "DFC" = 84L, "MEC" = 84L)
observed <- roi_objects %>% count(brain_region) %>% deframe()

show_table(
  tibble(brain_region = names(readme_claim),
         readme_says = as.integer(readme_claim),
         per_donor_objects = as.integer(observed[names(readme_claim)])) %>%
    mutate(difference = per_donor_objects - readme_says),
  "The release README states that all 84 donors were profiled in MTG, DFC and MEC"
)
```

The per-donor object counts for DFC and MEC fall short of the README's claim by 4 and 3. Two
readings are possible — the per-donor split is incomplete, or those donors' nuclei were dropped
at QC — and the manifest cannot distinguish them. **Unresolved**; resolving it needs the region
metadata table (`SEAAD_DFC_RNAseq_final-nuclei_metadata.csv`, 1.3 GB), not a guess.

### 2.3 Lineage coverage

```{r}
#| label: lineage-coverage

taxonomy <- seaad_taxonomy(omics$taxonomy)

show_table(
  taxonomy %>% count(lineage, name = "n_supertypes") %>% arrange(desc(n_supertypes)),
  "Lineage regrouping of the 2026 taxonomy used throughout this notebook"
)
```

The released `Immune` subclass mixes microglia/PVM with lymphocytes and monocytes, so
`microglia_PVM` is defined by the `Micro-PVM` **supertype** prefix. Oligodendrocyte and OPC are
kept apart, never merged into one "oligo lineage".

```{r}
#| label: lineage-by-roi

abundances <- read_csv(omics$abundances, show_col_types = FALSE)
lib_comp   <- seaad_library_composition(abundances, taxonomy, regions = ROI)

show_table(
  lib_comp %>%
    filter(lineage %in% c("microglia_PVM", "astrocyte", "neuron",
                          "oligodendrocyte", "OPC")) %>%
    group_by(brain_region, lineage) %>%
    summarise(n_libraries_with_lineage = n_distinct(library_prep),
              n_nuclei = sum(n_nuclei), .groups = "drop") %>%
    left_join(lib_comp %>% group_by(brain_region) %>%
                summarise(n_libraries = n_distinct(library_prep), .groups = "drop"),
              by = "brain_region") %>%
    mutate(libraries_missing_lineage = n_libraries - n_libraries_with_lineage) %>%
    select(brain_region, lineage, n_libraries, n_libraries_with_lineage,
           libraries_missing_lineage, n_nuclei) %>%
    arrange(brain_region, lineage),
  "Lineage presence per ROI at LIBRARY grain, not donor grain"
)
```

**This is not the donor-level answer the analysis needs.** The requested counts are "n donors
with microglia/PVM", "n donors with astrocytes" and so on; the abundance objects carry
`library_prep` only, and several libraries share a donor (DFC has `r sum(lib_comp$brain_region == "DFC" & lib_comp$lineage == "neuron")`
libraries for `r sum(roi_objects$brain_region == "DFC")` donors). What the table does establish is
a ceiling on the problem: essentially every library in every ROI carries all four lineages, so a
donor could only lack a lineage if *all* of that donor's libraries lacked it.

The exact donor-level counts are one field lookup away in the pseudobulk objects — `obs$`Donor
ID``, `obs$`Brain Region``, `obs$`Number of nuclei`` — and are reported as **unverified** until
those are fetched (sizes in section 3.1).

### 2.4 Overlap with the QNP primary-candidate block

```{r}
#| label: qnp-overlap

is_obs <- function(x) !is.na(x)

qnp_roi <- qnp_ph %>%
  filter(brain_region %in% ROI) %>%
  mutate(ph_complete = is_obs(abeta_percent_positive_area) &
                       is_obs(ptau_percent_positive_area) &
                       is_obs(neun_cells_per_area) &
                       is_obs(gfap_percent_positive_area))

coverage <- qnp_roi %>%
  group_by(brain_region) %>%
  summarise(n_QNP_rows = n(), n_QNP_PH = sum(ph_complete), .groups = "drop") %>%
  left_join(roi_objects %>% count(brain_region, name = "n_omics"), by = "brain_region") %>%
  left_join(
    qnp_roi %>% filter(ph_complete) %>%
      inner_join(roi_objects, by = c("donor_id", "brain_region")) %>%
      count(brain_region, name = "n_complete_QNP_PH_omics"),
    by = "brain_region") %>%
  mutate(n_omics_without_QNP_PH = n_omics - n_complete_QNP_PH_omics) %>%
  arrange(desc(n_complete_QNP_PH_omics))

show_table(coverage, "QNP P/H and omics coverage by ROI, exact donor identities only")

cat("donor x ROI observations with both QNP P/H and omics:",
    sum(coverage$n_complete_QNP_PH_omics), "\n")
```

```{r}
#| label: where-the-losses-are

show_table(
  qnp_primary_long(qnp) %>%
    filter(brain_region %in% ROI, status != "observed") %>%
    count(brain_region, variable, status, name = "n_donors"),
  "Why the QNP side is short: the only incomplete ROI is V1C, and the cause is missing values, not a missing stain"
)

show_table(
  roi_objects %>%
    anti_join(qnp_roi %>% filter(ph_complete), by = c("donor_id", "brain_region")) %>%
    select(donor_id, brain_region),
  "The donor x ROI cells with an omics object but no complete QNP P/H"
)
```

The whole shortfall is V1C: four donors have no `number of NeuN positive cells per area` and one
has no `percent GFAP positive area`, with the stain present in both cases
(`assayed_but_missing` in the notebook-00 classification). One of those five donors also has a
V1C omics object, which is the single V1C loss. MEC's one loss is the reverse: a donor with an
omics object whose QNP row does not exist at all. Everything else joins completely.

### 2.5 Donors complete across ROI subsets

Donors are not required to have all ROIs. This table only shows what the ROI choice costs if a
complete-case design is later preferred.

```{r}
#| label: simultaneous

both <- qnp_roi %>% filter(ph_complete) %>%
  inner_join(roi_objects, by = c("donor_id", "brain_region"))

show_table(
  both %>% count(donor_id, name = "n_roi") %>% count(n_roi, name = "n_donors") %>%
    arrange(desc(n_roi)),
  "Donors by number of ROIs with both QNP P/H and omics"
)

complete_in <- function(regions) {
  both %>% filter(brain_region %in% regions) %>%
    count(donor_id) %>% filter(n == length(regions)) %>% nrow()
}

show_table(
  tribble(
    ~region_set, ~regions,
    "all six", ROI,
    "drop HIP", setdiff(ROI, "HIP"),
    "drop HIP, STG, V1C", setdiff(ROI, c("HIP", "STG", "V1C")),
    "the three 80+ donor regions", c("DFC", "MEC", "MTG")
  ) %>%
    mutate(n_regions = lengths(regions),
           n_donors_complete = map_int(regions, complete_in),
           regions = map_chr(regions, paste, collapse = ", ")),
  "Donors complete in EVERY region of the set"
)
```

The binding constraint is HIP (40 donors) and the 43-donor "AD Spectrum" regions STG and V1C.

---

## 3. Donor-aware molecular summaries that already exist

### 3.1 Published pseudobulk

```{r}
#| label: pseudobulk-inventory

# The subclass token in the file name is a filesystem-safe spelling of the
# taxonomy's Subclass label ("L23-IT" for "L2/3 IT"). The three irregular cases
# are listed explicitly rather than guessed at.
subclass_from_filename <- function(token, taxonomy_subclasses) {
  irregular <- c("L23-IT" = "L2/3 IT", "L56-NP" = "L5/6 NP",
                 "VLMC-Perivascular" = "VLMC & Perivascular")
  out <- dplyr::coalesce(
    unname(irregular[token]),
    if_else(token %in% taxonomy_subclasses, token, NA_character_),
    if_else(str_replace_all(token, "-", " ") %in% taxonomy_subclasses,
            str_replace_all(token, "-", " "), NA_character_)
  )
  out
}

subclass_class <- taxonomy %>% distinct(subclass, class)

pseudobulk <- manifest %>%
  filter(family == "pseudobulk (donor x region x supertype)") %>%
  mutate(file_token = str_match(file, "^SEAAD_(.+?)_HIP_MEC")[, 2],
         subclass = subclass_from_filename(file_token, unique(taxonomy$subclass))) %>%
  left_join(subclass_class, by = "subclass") %>%
  mutate(lineage_group = case_when(
    subclass == "Astrocyte"       ~ "astrocyte",
    subclass == "Oligodendrocyte" ~ "oligodendrocyte",
    subclass == "OPC"             ~ "OPC",
    subclass == "Immune"          ~ "microglia / immune (mixed)",
    str_starts(class, "Neuronal") ~ "neuron",
    TRUE                          ~ "other non-neural")) %>%
  transmute(file_token, subclass, lineage_group,
            size_mb = round(size_mb, 1), release_date)

stopifnot(
  "every pseudobulk file token must resolve to a taxonomy subclass" =
    !any(is.na(pseudobulk$subclass)),
  "one pseudobulk object per taxonomy subclass" =
    nrow(pseudobulk) == n_distinct(taxonomy$subclass)
)

show_table(
  pseudobulk %>% arrange(desc(size_mb)),
  "All 29 published pseudobulk objects, one per taxonomy subclass. None is local."
)

show_table(
  pseudobulk %>%
    group_by(lineage_group) %>%
    summarise(n_objects = n(), total_mb = round(sum(size_mb)), .groups = "drop") %>%
    arrange(desc(total_mb)),
  "Fetch cost by lineage group"
)
```

Naming caution: the subclass token in the file name is followed by
`HIP_MEC_LEC_ITG_MTG_FI_STG_DFC_AnG_V1C`, which names the regions the **object family** spans,
not the regions in which that subclass occurs. Section 3.2 shows a counter-example.

### 3.2 Verified structure of a pseudobulk object

The smallest of the 29 (`SEAAD_Ependymal_...`, 4.2 MB) was fetched by name to establish the
family's grain and expression scale. Only its `.obs` and shape are read; no expression value is
carried forward.

```{r}
#| label: pseudobulk-probe

probe_shape <- prov$pseudobulk_probe_shape

show_table(
  tibble::tribble(
    ~property, ~value,
    "row grain", paste("one row per", paste(c("Donor ID", "Brain Region", "Supertype"),
                                            collapse = " x ")),
    "n rows", as.character(probe_shape$n_obs),
    "n genes", format(probe_shape$n_vars, big.mark = ","),
    "X all integral", as.character(probe_shape$x_all_integral),
    "layers present", as.character(probe_shape$layers_present),
    "expression scale", "summed raw UMI counts (X is integral and scales with nuclei)",
    "raw counts available", "yes - X *is* the raw count sum; no normalised layer to undo",
    "donor identifier", "obs$`Donor ID`, exact H##.##.### form",
    "region identifier", "obs$`Brain Region`",
    "nuclei denominator", "obs$`Number of nuclei`, per row"
  ),
  "Verified properties of the pseudobulk family, from one probed object"
)

show_table(
  probe_obs %>%
    group_by(`Brain Region`, Subclass, Supertype) %>%
    summarise(n_donors = n_distinct(`Donor ID`),
              n_nuclei = sum(`Number of nuclei`),
              nuclei_min = min(`Number of nuclei`),
              nuclei_median = median(`Number of nuclei`),
              umi_per_nucleus_median = round(median(total_umi / `Number of nuclei`)),
              .groups = "drop"),
  "Probe object contents: Ependymal occurs in HIP only, despite the 10-region file name"
)
```

Three consequences:

- **X is raw counts.** This satisfies the section 7 requirement that molecular-state scoring use
  count-derived expression, not latent coordinates. Any normalisation is a downstream choice, not
  something to be undone.
- **`Number of nuclei` is carried per row**, so a nuclei-count floor can be imposed per
  donor x region x supertype, and the same column gives composition (section 5).
- **Some rows rest on very few nuclei** (minimum 1 in the probe). A floor is mandatory, not
  optional, and choosing it is a later decision.

Region, lineage and donor coverage of the other 28 objects is **unverified**; only the file names
are known. The probe shows why the file name cannot be read as coverage.

### 3.3 Published composition objects

```{r}
#| label: composition-objects

show_table(
  manifest %>%
    filter(family == "compositional model output") %>%
    transmute(file = str_remove(key, "^Multiregion_2026/model_outputs/pertpy_compositional_modeling/"),
              size_mb = round(size_mb, 2)) %>%
    arrange(size_mb),
  "pertpy / scCODA compositional-model outputs"
)

show_table(
  tibble::tribble(
    ~property, ~value,
    "row grain", "library_prep x supertype",
    "n libraries", as.character(n_distinct(abundances$library_prep)),
    "n supertypes", as.character(n_distinct(abundances$supertype)),
    "value", "nuclei count",
    "donor identifier", "ABSENT - obs carries library_prep and scCODA_sample_id only",
    "region identifier", "obs$`Brain Region`",
    "covariates carried", "CPS_Local, PMI, Sex, APOE4 status, binned age"
  ),
  "Structure of the supertype-abundance objects"
)
```

These are the released *inputs* to SEA-AD's compositional model, and they are the cheapest
composition data in the release (1.9 MB for all 10 regions). Their limitation is decisive for
this analysis: **they have no donor field.** Using them as a donor-level composition variable
would require a `library_prep -> Donor ID` map, which exists only inside the per-region metadata
CSVs (235 MB - 1.4 GB) or the pseudobulk `.obs`. Since the pseudobulk objects carry
`Number of nuclei` per donor x region x supertype directly, they are the better source for
composition as well as state, and the abundance objects are used below only for library-grain
description.

```{r}
#| label: pertpy-summary

pertpy <- read_csv(omics$pertpy_summary, show_col_types = FALSE,
                   name_repair = ~ ifelse(.x == "", "row_index", .x)) %>%
  select(cell_type = `Cell Type`, local_model = `Local Model`, region = Region)

show_table(
  pertpy %>%
    group_by(region) %>%
    summarise(n_cell_types = n(),
              n_with_effect = sum(!is.na(local_model)),
              n_zero_effect = sum(local_model == 0, na.rm = TRUE),
              .groups = "drop") %>%
    filter(region %in% ROI),
  "Published compositional effect sizes vs local CPS, by ROI"
)
```

This is a **published result about composition**, one number per cell type per region. It is not
a donor x ROI variable and cannot become one; it is recorded here as evidence that SEA-AD treats
composition-vs-stage as its own question, separate from expression.

### 3.4 Published donor x region pseudo-progression, and a CPS ambiguity

```{r}
#| label: cps-grid

show_table(
  cps_region %>% count(brain_region, name = "n_donors") %>% arrange(brain_region),
  "Donor x region grid of Global_and_Local_CPS.20260105.csv"
)

cps_matches_qnp <- map_dfr(intersect(unique(cps_region$brain_region),
                                     unique(qnp_ph$brain_region)), function(r) {
  q <- qnp_ph %>% filter(brain_region == r) %>% pull(donor_id)
  c2 <- cps_region %>% filter(brain_region == r) %>% pull(donor_id)
  tibble(brain_region = r, same_donor_set_as_QNP = setequal(q, c2))
})

show_table(cps_matches_qnp, "Its grid is the QNP grid, region by region")
```

The published `CPS_Local` grid is **exactly** the QNP donor x region grid in all nine shared
regions — including HIP at 84 donors, where the omics has 40. `CPS_Local` is therefore a
neuropathology-derived regional quantity, not an omics-derived one, and it is not evidence of
omics coverage.

```{r}
#| label: cps-three-way

specimen_index <- read.csv(paths$specimen_index, check.names = FALSE)

cps_compare <- cps_region %>%
  distinct(donor_id, CPS_Global) %>%
  inner_join(specimen_index %>%
               transmute(donor_id = `Donor ID`,
                         cps_specimen = `Continuous Pseudo-progression Score`),
             by = "donor_id") %>%
  mutate(difference = CPS_Global - cps_specimen)

show_table(
  tibble(n_donors = nrow(cps_compare),
         max_abs_difference = max(abs(cps_compare$difference)),
         median_abs_difference = median(abs(cps_compare$difference)),
         n_donors_differing_by_gt_0.1 = sum(abs(cps_compare$difference) > 0.1)),
  "CPS_Global (2026 omics release) vs Continuous Pseudo-progression Score (SpecimenMetadata.csv)"
)
```

Notebook 00 section 10.2 already established that the caudate file's `CPS` differs from
`SpecimenMetadata.csv`, and pinned CPS to `SpecimenMetadata.csv`. There are now **three**
published CPS columns, and the omics release's `CPS_Global` is a third distinct one.

**Nothing here changes the notebook-00 definition.** Every use of CPS below names its source
column explicitly. Which CPS a joint P/H/T analysis should use is an open question, listed in
section 11.

### 3.5 Model outputs that are out of scope

```{r}
#| label: out-of-scope

show_table(
  manifest %>%
    filter(family %in% c("differential-expression model output",
                         "integrated latent representation",
                         "taxonomy mapping output")) %>%
    group_by(family) %>%
    summarise(n_objects = n(), total_gb = round(sum(size_mb) / 1000, 1), .groups = "drop"),
  "Released model outputs excluded by the guardrails of this task"
)
```

The gpboost outputs are gene-level differential-expression summary statistics against CPS —
exactly what section 8 forbids selecting features from, and what the task forbids computing. The
final embeddings are scVI latent coordinates, excluded by section 7. The scANVI/scVI mapping
outputs are taxonomy machinery, usable for understanding cell identity and nothing else.

---

## 4. Candidate T axes

### 4.1 Feature-origin vocabulary

```{r}
#| label: feature-origin

show_table(
  tibble::tribble(
    ~feature_origin, ~meaning,
    "SEAAD_published_predefined",
      "a quantity SEA-AD itself publishes as a named variable or label",
    "external_predefined",
      "a gene set or score defined in an external publication, applied unchanged",
    "current_analysis_derived",
      "computed here from released counts by a stated rule; no gene set invented"
  ),
  "Feature-origin categories used in sections 4 and 10"
)
```

No gene set is invented anywhere in this notebook, and no cognition- or resilience-derived
signature is used to define a primary $T$. Published cognitive-resilience signatures are noted in
section 11 as possible *secondary validation* features only.

### 4.2 What the release does and does not provide

```{r}
#| label: state-label-search

show_table(
  tibble::tribble(
    ~sought, ~found_in_release, ~where,
    "a donor x region continuous molecular-state score per lineage", "NO", "-",
    "a per-nucleus disease-associated state label (e.g. 'DAM', 'reactive astrocyte')", "NO",
      "obs carries Class / Subclass / Supertype only",
    "disease-associated SUPERTYPES named as such", "PARTIAL",
      "`-SEAAD` suffixed supertypes are new types found in the AD cohort, not state labels",
    "donor-level neuropathology and cognition in obs", "YES",
      "Thal / Braak / CERAD / ADNC / LATE / Cognitive Status",
    "a published per-region pseudo-progression score", "YES",
      "Global_and_Local_CPS.20260105.csv (neuropathology-derived)",
    "published composition-vs-stage effect sizes", "YES",
      "pertpy_summary_CPS_Local.20260622.csv (region-level, not donor-level)"
  ),
  "Search for published molecular-state representations"
)

show_table(
  taxonomy %>%
    mutate(is_seaad_new = str_detect(supertype, "-SEAAD$")) %>%
    filter(lineage %in% c("microglia_PVM", "astrocyte", "oligodendrocyte", "OPC")) %>%
    group_by(lineage) %>%
    summarise(n_supertypes = n(), n_new_in_SEAAD = sum(is_seaad_new),
              supertypes = paste(supertype, collapse = ", "), .groups = "drop"),
  "Glial supertypes, and which are new in the SEA-AD taxonomy"
)
```

**There is no published continuous disease-state score per lineage, and no published
disease-associated cell-state label.** The `-SEAAD` supertypes are the closest thing, and they
are *types added to the taxonomy*, not state annotations: their abundance is composition, not
within-lineage expression state.

This is the single most important finding for question 4. It means every candidate $T$ that is
genuinely *state* must be `current_analysis_derived` from released pseudobulk counts, under a
stated rule.

### 4.3 Candidate table

```{r}
#| label: candidate-axes

candidates <- tibble::tribble(
  ~candidate, ~axis, ~lineage, ~kind, ~interpretation, ~source_object, ~feature_origin,
  "T_microglia_state",
    "T_microglia", "Micro-PVM supertypes", "state",
    "What microglia in this ROI are expressing, at fixed lineage definition",
    "pseudobulk_objects/SEAAD_Immune_... (124 MB)", "current_analysis_derived",
  "T_microglia_composition",
    "T_microglia", "Micro-PVM supertypes", "composition",
    "Share of nuclei that are microglia/PVM, and the mix of Micro-PVM supertypes",
    "pseudobulk obs$`Number of nuclei`; abundance objects (library grain)", "SEAAD_published_predefined",
  "T_astrocyte_state",
    "T_astrocyte", "Astrocyte subclass", "state",
    "Expression state of astrocytes, e.g. a reactive-astrocyte axis over Astro_1..6",
    "pseudobulk_objects/SEAAD_Astrocyte_... (169 MB)", "current_analysis_derived",
  "T_astrocyte_composition",
    "T_astrocyte", "Astrocyte subclass", "composition",
    "Astrocyte share of nuclei; relative abundance of Astro_6-SEAAD",
    "pseudobulk obs$`Number of nuclei`", "SEAAD_published_predefined",
  "T_neuronal_state_subclass_conditioned",
    "T_neuronal", "one defensible neuronal grouping, conditioned", "state",
    "Molecular / synaptic state within a fixed neuronal subclass present in all six ROIs",
    "pseudobulk_objects/SEAAD_<neuronal subclass>_... (see 6.2)", "current_analysis_derived",
  "T_neuronal_state_all_neuron",
    "T_neuronal", "all neurons pooled", "state - CONFOUNDED",
    "All-neuron expression summary; mixes state, subtype composition and selective loss",
    "any neuronal pseudobulk pooled", "current_analysis_derived",
  "T_vulnerable_neuron_composition",
    "T_neuronal", "vulnerable neuronal supertypes", "composition",
    "Relative abundance of the neuronal types SEA-AD reports as selectively lost",
    "pseudobulk obs$`Number of nuclei`; pertpy_summary_CPS_Local", "SEAAD_published_predefined",
  "T_oligo_OPC_state",
    "T_oligodendrocyte_OPC", "Oligodendrocyte + OPC, kept separate", "state",
    "Myelination / remyelination-related expression state within each lineage",
    "pseudobulk SEAAD_Oligodendrocyte_... (147 MB) + SEAAD_OPC_... (84 MB)",
    "current_analysis_derived",
  "T_oligo_OPC_composition",
    "T_oligodendrocyte_OPC", "Oligodendrocyte + OPC, kept separate", "composition",
    "Oligodendrocyte and OPC shares of nuclei; OPC-to-oligodendrocyte ratio",
    "pseudobulk obs$`Number of nuclei`", "SEAAD_published_predefined"
)

show_table(
  candidates %>% select(candidate, axis, kind, interpretation),
  "Candidate T variables: what each one means"
)

show_table(
  candidates %>% select(candidate, lineage, source_object, feature_origin),
  "Candidate T variables: where each one would come from"
)
```

```{r}
#| label: candidate-coverage

roi_donor_counts <- coverage %>% select(brain_region, n_complete_QNP_PH_omics)

show_table(
  candidates %>%
    transmute(candidate, kind,
              regions_available = c(
                rep("all six ROI", 4),
                "depends on the neuronal subclass chosen (section 6.2)",
                "all six ROI",
                "depends on which supertypes are called vulnerable (section 6.2)",
                rep("all six ROI", 2)),
              n_donors_min_roi = min(roi_donor_counts$n_complete_QNP_PH_omics),
              n_donors_max_roi = max(roi_donor_counts$n_complete_QNP_PH_omics),
              n_donor_x_roi = sum(roi_donor_counts$n_complete_QNP_PH_omics),
              remaining_ambiguity = c(
                "no published microglial state score; the scoring rule is undefined and must be chosen",
                "Immune subclass mixes microglia with lymphocytes/monocytes; supertype filter required",
                "no published reactive-astrocyte score; the scoring rule is undefined",
                "astrocyte share depends on dissociation yield as well as biology",
                "which neuronal grouping is defensible in all six ROIs is unsettled (section 6)",
                "NOT USABLE as a primary T - listed to be excluded explicitly",
                "SEA-AD's vulnerability claims are per region; a cross-ROI definition is not published",
                "oligodendrocyte and OPC may behave oppositely; must not be pooled",
                "nuclei-yield sensitivity as above, plus very low OPC counts in some libraries")),
  "Candidate T variables: coverage and what is still unresolved"
)
```

Coverage is identical across candidates because it is set by the donor x ROI grid of section 2.4,
not by the lineage: every ROI object contains every lineage.

---

## 5. Molecular state and cellular composition, kept apart

This is a hard requirement, so it is stated as a rule and then demonstrated.

```{r}
#| label: state-vs-composition

show_table(
  tibble::tribble(
    ~pair, ~composition_variable, ~state_variable, ~why_not_interchangeable,
    "microglia",
      "fraction of nuclei that are microglia/PVM; fraction of microglia that are Micro-PVM_3-SEAAD",
      "expression profile computed across microglial nuclei only",
      "A donor can have unchanged microglial numbers and a wholly changed microglial programme, or the reverse",
    "astrocyte",
      "astrocyte fraction of nuclei; share of Astro_6-SEAAD",
      "expression profile across astrocytic nuclei only",
      "Astrocyte share also tracks dissociation yield and neuronal loss in the denominator",
    "neuron",
      "neuronal subtype composition; relative loss of vulnerable supertypes",
      "expression profile within a fixed neuronal grouping",
      "A drop in an all-neuron summary can be pure composition change with no change in any surviving cell",
    "oligodendrocyte / OPC",
      "oligodendrocyte and OPC shares; OPC-to-oligodendrocyte ratio",
      "expression profile within each lineage separately",
      "Remyelination predicts a composition shift AND a state shift, in possibly opposite directions"
  ),
  "Composition and state are separate candidate variables for every lineage"
)
```

Two operational rules follow, and both are guardrails for later work rather than anything
computed here:

1. **Neither is adjusted for the other.** A state variable is not residualised on composition,
   and a composition variable is not residualised on state. If both matter they enter as two
   variables.
2. **The denominator is always named.** "Fraction of microglia that are Micro-PVM_3-SEAAD" and
   "fraction of nuclei that are Micro-PVM_3-SEAAD" are different variables with different
   confounds, and the pseudobulk `Number of nuclei` column supports both.

```{r}
#| label: composition-illustration

show_table(
  lib_comp %>%
    filter(lineage %in% c("microglia_PVM", "astrocyte", "neuron",
                          "oligodendrocyte", "OPC")) %>%
    group_by(brain_region, lineage) %>%
    summarise(median_fraction = median(fraction), .groups = "drop") %>%
    pivot_wider(names_from = lineage, values_from = median_fraction),
  "Median lineage fraction per ROI, library grain. A composition measurement, and nothing more."
)
```

---

## 6. The neuronal axis

### 6.1 Why an all-neuron summary is not usable

```{r}
#| label: neuron-composition-spread

neuron_mix <- abundances %>%
  filter(brain_region %in% ROI) %>%
  left_join(taxonomy, by = "supertype") %>%
  filter(lineage == "neuron") %>%
  group_by(brain_region, subclass) %>%
  summarise(n_nuclei = sum(n_nuclei), .groups = "drop") %>%
  group_by(brain_region) %>%
  mutate(share_of_neurons = n_nuclei / sum(n_nuclei)) %>%
  ungroup()

show_table(
  neuron_mix %>%
    select(brain_region, subclass, share_of_neurons) %>%
    pivot_wider(names_from = brain_region, values_from = share_of_neurons,
                values_fill = 0) %>%
    arrange(desc(MTG)),
  "Neuronal subclass share of all neurons, by ROI. A zero is a genuine absence: no nucleus of that subclass was mapped in that ROI."
)
```

An all-neuron expression summary varies across ROI for three reasons that it cannot separate:
within-cell state, subtype composition, and selective neuronal loss. It is listed in section 4.3
only so that it can be ruled out explicitly.

### 6.2 Which neuronal groupings are conditionable

```{r}
#| label: neuron-subclass-coverage

tax_regions <- read_csv(omics$taxonomy, show_col_types = FALSE) %>%
  select(supertype = cluster_label, subclass = subclass_label, class = class_label,
         all_of(ROI)) %>%
  pivot_longer(all_of(ROI), names_to = "brain_region", values_to = "present")

neuronal_coverage <- tax_regions %>%
  filter(str_starts(class, "Neuronal")) %>%
  group_by(subclass, brain_region) %>%
  summarise(n_supertypes_present = sum(present), .groups = "drop") %>%
  pivot_wider(names_from = brain_region, values_from = n_supertypes_present) %>%
  mutate(n_roi_present = rowSums(across(all_of(ROI)) > 0)) %>%
  arrange(desc(n_roi_present), subclass)

show_table(neuronal_coverage,
           "Neuronal supertypes present per ROI, from the release's own taxonomy table")

show_table(
  neuronal_coverage %>%
    filter(n_roi_present == length(ROI)) %>%
    left_join(neuron_mix %>% group_by(subclass) %>%
                summarise(min_share = min(share_of_neurons),
                          max_share = max(share_of_neurons), .groups = "drop"),
              by = "subclass") %>%
    left_join(subclass_class, by = "subclass") %>%
    mutate(share_ratio = round(max_share / min_share, 1),
           neuron_class = str_remove(class, "^Neuronal: ")) %>%
    select(subclass, neuron_class, all_of(ROI), min_share, max_share, share_ratio) %>%
    arrange(share_ratio),
  "Neuronal subclasses present in all six ROI, ordered by how much their share of neurons varies across ROI"
)

show_table(
  neuronal_coverage %>%
    filter(HIP == 0) %>%
    select(subclass, all_of(ROI)),
  "Neuronal subclasses with no supertype in HIP at all"
)
```

Two facts from these tables drive the neuronal decision.

**The layer-defined excitatory subclasses do not exist in HIP.** `L2/3 IT`, `L4 IT`, `L5 IT`,
`L5/6 NP` and `L6 IT Car3` have zero supertypes there, which is anatomy rather than a data
problem — hippocampus has no six-layer isocortical sheet. Any excitatory-based neuronal $T$ is
therefore undefined in HIP, and would force either dropping HIP or accepting a structurally
missing cell.

**The subclasses that do span all six ROI vary in share, and the abundant GABAergic ones vary
least.** `Vip`, `Lamp5`, `Sst` and `Pvalb` each shift by under a factor of six across ROI, while
`L5 ET`, `L6b` and `L6 IT` shift by 30-fold or more. The `share_ratio` column makes that visible
instead of asserting it. The rare GABAergic types `Pax6` and `Sst Chodl` are exceptions and are
not candidates for other reasons: they are too sparse to pseudobulk reliably.

Two genuinely defensible alternatives follow, and this notebook does not choose between them:

- **A**: a within-subclass state variable on an inhibitory subclass present in all six ROI
  (`Vip`, `Sst`, `Pvalb`, `Lamp5`), which minimises composition confounding and survives HIP, but
  describes a minority of neurons and not the selectively vulnerable ones.
- **B**: a within-subclass state variable computed per excitatory subclass and combined with
  fixed weights, which covers the vulnerable populations but is undefined in HIP and needs the
  weights justified.

Sections 4.3 and 10 therefore leave the neuronal primary **unresolved**. The choice interacts
with the ROI decision, which is why it is not made here.

---

## 7. Integrated latent coordinates are not used as measurements

```{r}
#| label: latent-excluded

show_table(
  manifest %>%
    filter(family == "integrated latent representation") %>%
    transmute(file, size_mb = round(size_mb)),
  "Released scVI/scANVI artefacts"
)
```

`X_scVI` and `X_umap` are available for every nucleus, and `obsm$X_scVI` is in the region and
subclass objects. They are used here for nothing. They may inform how cell identity was assigned;
they are not quantitative expression and will not be a $T$ variable. The quantitative substrate is
the pseudobulk raw UMI counts verified in section 3.2.

---

## 8. Disease-stage behaviour of what can be evaluated today

This section reads the *shape* of a relationship — monotonic, stage-specific, saturated, or flat.
It performs no test and selects nothing. Whichever candidate shows the strongest association is
not thereby preferred.

Only the **composition** candidates can be evaluated now, at library grain, because the state
candidates require pseudobulk objects that are not local. The state candidates are therefore
`unevaluated`, not `flat`.

### 8.1 QNP pathology against the published local CPS, by ROI

```{r}
#| label: qnp-vs-cps
#| fig-width: 9
#| fig-height: 5.5

qnp_cps <- qnp_roi %>%
  inner_join(cps_region, by = c("donor_id", "brain_region")) %>%
  select(donor_id, brain_region, CPS_Local, CPS_Local_ABeta, CPS_Local_pTau,
         abeta = abeta_percent_positive_area, ptau = ptau_percent_positive_area,
         neun = neun_cells_per_area, gfap = gfap_percent_positive_area)

qnp_cps %>%
  pivot_longer(c(abeta, ptau, neun, gfap), names_to = "qnp", values_to = "value") %>%
  filter(!is.na(value)) %>%
  ggplot(aes(CPS_Local, value, colour = brain_region)) +
  geom_point(size = 0.8, alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.5, span = 1) +
  facet_wrap(~ qnp, scales = "free_y") +
  labs(title = "Measured QNP P/H against the published local CPS",
       subtitle = "Rows are donor x ROI. CPS_Local from Global_and_Local_CPS.20260105.csv",
       x = "CPS_Local", y = "QNP value", colour = "ROI") +
  theme_minimal(base_size = 20)
```

```{r}
#| label: qnp-vs-local-path
#| fig-width: 9
#| fig-height: 4

qnp_cps %>%
  select(brain_region, CPS_Local_ABeta, CPS_Local_pTau, abeta, ptau) %>%
  pivot_longer(c(abeta, ptau), names_to = "qnp", values_to = "measured") %>%
  mutate(model = if_else(qnp == "abeta", CPS_Local_ABeta, CPS_Local_pTau)) %>%
  filter(!is.na(measured)) %>%
  ggplot(aes(model, measured, colour = brain_region)) +
  geom_point(size = 0.8, alpha = 0.6) +
  facet_wrap(~ qnp, scales = "free") +
  labs(title = "Measured QNP pathology against the release's modelled local pathology score",
       subtitle = "These are different quantities: one is percent positive area, the other a pseudo-progression coordinate",
       x = "CPS_Local_ABeta / CPS_Local_pTau", y = "QNP percent positive area",
       colour = "ROI") +
  theme_minimal(base_size = 20)
```

`CPS_Local_ABeta` and `CPS_Local_pTau` are **not** substitutes for the QNP P block. They are
model-derived coordinates on the same donor x region grid; the QNP columns are measurements. Both
stay available, and the P block stays as notebook 00 defined it.

### 8.2 Composition against local CPS, by ROI

```{r}
#| label: composition-vs-cps
#| fig-width: 9
#| fig-height: 6

comp_focus <- lib_comp %>%
  filter(lineage %in% c("microglia_PVM", "astrocyte", "oligodendrocyte", "OPC", "neuron"))

comp_focus %>%
  ggplot(aes(cps_local, fraction, colour = brain_region)) +
  geom_point(size = 0.7, alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.5, span = 1.2) +
  facet_wrap(~ lineage, scales = "free_y") +
  labs(title = "Lineage composition against local CPS, by ROI",
       subtitle = "LIBRARY grain, not donor grain: several libraries share a donor. Shape only; no test, no selection.",
       x = "CPS_Local (carried in the abundance object)", y = "fraction of nuclei in library",
       colour = "ROI") +
  theme_minimal(base_size = 20)
```

```{r}
#| label: composition-binned

show_table(
  comp_focus %>%
    group_by(brain_region) %>%
    mutate(cps_bin = cut(cps_local, breaks = quantile(cps_local, seq(0, 1, 0.25)),
                         include.lowest = TRUE, labels = c("Q1", "Q2", "Q3", "Q4"))) %>%
    group_by(brain_region, lineage, cps_bin) %>%
    summarise(median_fraction = median(fraction), n_libraries = n(), .groups = "drop") %>%
    select(-n_libraries) %>%
    pivot_wider(names_from = cps_bin, values_from = median_fraction) %>%
    arrange(lineage, brain_region),
  "Median lineage fraction by within-ROI CPS_Local quartile (library grain)"
)
```

Read as shape only: the microglial and astrocytic fractions drift upward across CPS quartiles in
most ROIs, the neuronal fraction drifts downward, and OPC is close to flat. That is the expected
mirror-image of a shrinking neuronal denominator and is **precisely why composition is not a
substitute for state** — nothing in this table says what any cell is doing.

Because rows are libraries, these are descriptive shapes only. No donor-level claim is made and
no candidate is preferred or dropped on this basis.

---

## 9. Region specificity

Raw values differ enormously between ROI for reasons that have nothing to do with disease: MEC
and HIP are allocortex and hippocampus, V1C is granular isocortex, and their baseline cellular
composition differs by a factor of three.

```{r}
#| label: region-raw

show_table(
  comp_focus %>%
    group_by(brain_region, lineage) %>%
    summarise(median_fraction = median(fraction), .groups = "drop") %>%
    pivot_wider(names_from = lineage, values_from = median_fraction) %>%
    arrange(desc(microglia_PVM)),
  "Raw median lineage fraction by ROI. Retained, never overwritten."
)
```

```{r}
#| label: region-standardized
#| fig-width: 9
#| fig-height: 4.5

microglia_std <- comp_focus %>%
  filter(lineage == "microglia_PVM") %>%
  add_region_standardized("fraction")

stopifnot(
  "the raw value must survive standardisation" =
    all(c("fraction", "fraction_z_within_region") %in% names(microglia_std)),
  "standardisation must not alter the raw column" =
    identical(microglia_std$fraction,
              comp_focus %>% filter(lineage == "microglia_PVM") %>% pull(fraction))
)

bind_rows(
  microglia_std %>% transmute(brain_region, cps_local, value = fraction, scale = "raw fraction"),
  microglia_std %>% transmute(brain_region, cps_local, value = fraction_z_within_region,
                             scale = "within-ROI z")
) %>%
  ggplot(aes(cps_local, value, colour = brain_region)) +
  geom_point(size = 0.7, alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.5, span = 1.2) +
  facet_wrap(~ scale, scales = "free_y") +
  labs(title = "Microglial fraction, raw and standardised within ROI",
       subtitle = "The standardised view is an exploratory aid; the raw value and the ROI label are what get carried forward",
       x = "CPS_Local", y = NULL, colour = "ROI") +
  theme_minimal(base_size = 20)
```

`add_region_standardized()` adds a column and never replaces one. Any later $T$ table must carry
the raw value and the ROI identity; a within-ROI standardised column may sit beside them.

An MEC-vs-V1C difference in a raw $T$ value is **not** a disease effect. The obvious trap is that
MEC is both the highest-pathology ROI and an anatomically distinct one, so raw regional contrasts
confound stage with region.

---

## 10. Decision table

```{r}
#| label: decision-table

roi_cov_string <- paste0(nrow(coverage), "/", length(ROI), " ROI")
min_don <- min(coverage$n_complete_QNP_PH_omics)
max_don <- max(coverage$n_complete_QNP_PH_omics)
tot_obs <- sum(coverage$n_complete_QNP_PH_omics)

decision <- tibble::tribble(
  ~axis, ~candidate_definition, ~lineage, ~state_or_composition, ~source_release,
  ~feature_origin, ~status, ~remaining_issue,

  "microglial molecular state",
    "Expression summary over Micro-PVM supertypes only, from donor x ROI pseudobulk raw counts",
    "Micro-PVM supertypes", "state", "Multiregion 2026-06-22 pseudobulk_objects/SEAAD_Immune",
    "current_analysis_derived", "primary candidate",
    "Scoring rule undefined; no published microglial state score exists. Immune subclass must be filtered to Micro-PVM.",

  "microglial composition",
    "Micro-PVM share of ROI nuclei, and Micro-PVM supertype mix",
    "Micro-PVM supertypes", "composition", "pseudobulk obs$`Number of nuclei`",
    "SEAAD_published_predefined", "secondary",
    "Depends on the neuronal denominator and on dissociation yield.",

  "astrocyte molecular state",
    "Expression summary over Astrocyte-subclass nuclei, from donor x ROI pseudobulk raw counts",
    "Astrocyte", "state", "Multiregion 2026-06-22 pseudobulk_objects/SEAAD_Astrocyte",
    "current_analysis_derived", "primary candidate",
    "Scoring rule undefined; no published reactive-astrocyte score in the release.",

  "astrocyte composition",
    "Astrocyte share of ROI nuclei; Astro_6-SEAAD share of astrocytes",
    "Astrocyte", "composition", "pseudobulk obs$`Number of nuclei`",
    "SEAAD_published_predefined", "secondary",
    "Same denominator dependence as microglial composition.",

  "neuronal / synaptic molecular state",
    "Expression summary within ONE neuronal subclass present in all six ROI (options A and B in 6.2)",
    "a fixed neuronal subclass", "state",
    "Multiregion 2026-06-22 pseudobulk_objects/SEAAD_<subclass>",
    "current_analysis_derived", "UNRESOLVED",
    "Two defensible groupings (GABAergic-only, which spans all six ROI, vs weighted excitatory, which is undefined in HIP). Not chosen here.",

  "vulnerable-neuron composition",
    "Share of neurons in the supertypes SEA-AD reports as selectively lost",
    "vulnerable neuronal supertypes", "composition",
    "pseudobulk obs$`Number of nuclei`; pertpy_summary_CPS_Local",
    "SEAAD_published_predefined", "secondary",
    "SEA-AD's vulnerability claims are per region; no cross-ROI vulnerable set is published.",

  "oligodendrocyte / OPC state",
    "Separate expression summaries within Oligodendrocyte and within OPC, never pooled",
    "Oligodendrocyte; OPC", "state",
    "Multiregion 2026-06-22 pseudobulk_objects/SEAAD_Oligodendrocyte + SEAAD_OPC",
    "current_analysis_derived", "secondary (prespecified)",
    "Two lineages, so two variables. OPC counts are low in some libraries; a nuclei floor is required.",

  "oligodendrocyte / OPC composition",
    "Oligodendrocyte and OPC shares of ROI nuclei; OPC-to-oligodendrocyte ratio",
    "Oligodendrocyte; OPC", "composition", "pseudobulk obs$`Number of nuclei`",
    "SEAAD_published_predefined", "secondary",
    "The ratio is the interpretable form but is unstable at low OPC counts.",

  "all-neuron molecular state",
    "Expression summary over all neurons pooled",
    "all neurons", "state (confounded)", "any neuronal pseudobulk pooled",
    "current_analysis_derived", "EXCLUDED",
    "Confounds within-cell state, subtype composition and selective loss (section 6.1).",

  "integrated latent coordinates",
    "Donor x ROI mean of X_scVI",
    "any", "neither", "Multiregion 2026 final_embeddings/",
    "SEAAD_published_predefined", "EXCLUDED",
    "Not an expression measure; excluded by section 7."
) %>%
  mutate(excluded = status == "EXCLUDED",
         roi_coverage = case_when(
           excluded ~ "n/a",
           axis == "neuronal / synaptic molecular state" ~ "6/6 under option A; 5/6 under option B (no HIP)",
           TRUE ~ roi_cov_string),
         donor_coverage = if_else(excluded, "n/a",
                                  paste0(min_don, "-", max_don, " per ROI")),
         qnp_overlap_coverage = if_else(excluded, "n/a",
                                        paste0(tot_obs, " donor x ROI with QNP P/H"))) %>%
  select(-excluded)

show_table(
  decision %>% select(axis, candidate_definition, lineage, state_or_composition, status),
  "Candidate T variables: definition and status"
)

show_table(
  decision %>% select(axis, source_release, feature_origin, roi_coverage,
                     donor_coverage, qnp_overlap_coverage),
  "Candidate T variables: source and coverage"
)

show_table(
  decision %>% select(axis, status, remaining_issue),
  "Candidate T variables: what is still open"
)
```

Coverage is uniform across the live candidates because it is fixed by the donor x ROI grid of
section 2.4, not by the lineage. The two `EXCLUDED` rows carry no coverage and are kept visible
so that the exclusions are auditable rather than silent.

---

## 11. What remains unresolved

```{r}
#| label: unresolved

show_table(
  tibble::tribble(
    ~issue, ~kind, ~what_would_settle_it,
    "No pseudobulk object is local; 29 exist, 4-460 MB each",
      "coverage",
      "Fetch the four lineage objects for the prespecified axes: Immune 124 MB + Astrocyte 169 MB + Oligodendrocyte 147 MB + OPC 84 MB = 524 MB",
    "Donor-level lineage counts per ROI are unverified (library grain only)",
      "coverage", "The same four pseudobulk objects; obs carries Donor ID, Brain Region and Number of nuclei",
    "The release README claims 84 donors for DFC and MEC; 80 and 81 per-donor objects exist",
      "coverage", "The region metadata CSV (1.3-1.4 GB), or a note from SEA-AD",
    "No published molecular-state score or disease-state label for any lineage",
      "semantics", "A stated scoring rule, chosen before any association is inspected",
    "Three distinct published CPS columns now exist (SpecimenMetadata, caudate file, CPS_Global)",
      "semantics", "A decision on which CPS the joint P/H/T analysis uses; notebook 00 pinned SpecimenMetadata and that stands",
    "CPS_Local's grid is the QNP grid, not the omics grid",
      "semantics", "Nothing - established here; it simply must not be read as omics coverage",
    "The neuronal primary has two defensible definitions",
      "semantics", "A scientific choice between options A and B in section 6.2",
    "HIP has no layer-defined excitatory subclasses at all, so an excitatory-based neuronal T is undefined there",
      "coverage", "Either the GABAergic option A, or dropping HIP from the ROI set",
    "Nuclei-count floor per donor x ROI x supertype is unset",
      "semantics", "The pseudobulk Number of nuclei distributions, once fetched",
    "The ROI set is still candidate; HIP costs 40+ donors",
      "coverage", "The multimodal ROI decision, which is outside this notebook"
  ),
  "Open issues, separated into coverage and semantics"
)
```

Published cognitive-resilience signatures are deliberately absent from every table above. If they
are used at all it is as **secondary validation** of a $T$ defined independently, never as its
definition.

### The single next scientific decision

Whether to fetch the 524 MB of lineage pseudobulk objects and derive the state candidates, or to
first settle the ROI set. The two interact: dropping HIP raises the per-ROI donor floor from
`r min(coverage$n_complete_QNP_PH_omics)` to
`r min(coverage$n_complete_QNP_PH_omics[coverage$brain_region != "HIP"])`, and dropping HIP, STG
and V1C raises it to
`r min(coverage$n_complete_QNP_PH_omics[!coverage$brain_region %in% c("HIP", "STG", "V1C")])`,
at the cost of the anatomical contrast the design rests on.

```{r}
#| label: session

sessionInfo()
```
