---
title: "SEA-AD molecular-state exploration: from prepared glial pseudobulk to a provisional donor x region x supertype state block"
subtitle: "Which predefined signatures are defensible, whether their genes are there, and whether repeated library measurements can be combined"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    df-print: default
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose and scope

`01_seaad_omics_exploration.qmd` established that the SEA-AD multiregion release publishes
**no** molecular-state score and **no** disease-state cell label for any lineage, so every
genuine state variable has to be computed from released pseudobulk counts under a stated rule.
Acquisition and preparation have since been completed: four glial lineages (Immune, Astrocyte,
Oligodendrocyte, OPC) are local as raw-count H5AD objects on the released row grain.

This notebook asks how to turn those counts into **reliable, predefined molecular-state
measurements**. It answers nine questions:

1. Which externally predefined and SEA-AD-published glial state signatures are defensible candidates?
2. Are their genes adequately represented in the released pseudobulk data?
3. Are row-level state scores reproducible across repeated sample/library measurements?
4. Does assay method materially affect those scores?
5. Is there evidence for a minimum support criterion?
6. What aggregation rule should convert repeated measurements into donor x region x supertype estimates?
7. How strongly do state scores vary by supertype and region?
8. What provisional state variables are ready for later donor x ROI modelling?
9. Which uncertainties remain before defining final $T$ variables?

Deliberately **not** done here: any join to morphology or cognition; any feature selection against
an eventual outcome; any final $P/H/T \rightarrow$ morphology model; any ROI freeze; any neuronal
download; any further acquisition or upstream refactoring. Nothing below selects a signature,
a scoring parameter or an aggregation policy by its association with pathology, morphology,
cognition or cognitive resilience.

```{r}
#| label: setup

my_packages <- c("tidyverse", "knitr", "jsonlite", "readxl")
invisible(lapply(my_packages, library, character.only = TRUE))

find_root <- function(p = getwd()) {
  while (!dir.exists(file.path(p, ".git"))) {
    if (identical(dirname(p), p)) stop("project root not found")
    p <- dirname(p)
  }
  p
}
PROJECT_ROOT <- find_root()
source(file.path(PROJECT_ROOT, "R", "seaad_explore.R"))
source(file.path(PROJECT_ROOT, "R", "seaad_state.R"))

paths  <- seaad_paths(PROJECT_ROOT)
omics  <- seaad_omics_paths(PROJECT_ROOT)
state  <- seaad_state_paths(PROJECT_ROOT)

show_table <- function(x, caption = NULL, digits = 3) {
  kable(x, caption = caption, digits = digits, format.args = list(big.mark = ""))
}

theme_set(theme_minimal(base_size = 10))

# The candidate multimodal ROI set carried forward from notebook 01. Still a
# candidate: nothing here freezes it, and every table below reports all ten
# released regions so the freeze stays an open decision.
ROI <- c("DFC", "HIP", "MEC", "MTG", "STG", "V1C")

# Domains in a fixed display order, glial state axes first.
DOMAIN_LEVELS <- c("Micro/PVM", "Astrocyte", "Oligodendrocyte", "OPC", "Lymphocyte", "Monocyte")
```

```{r}
#| label: load-derivatives

registry   <- read_csv(state$registry_csv, show_col_types = FALSE)
registry_j <- fromJSON(state$registry_json, simplifyVector = FALSE)
sig_genes  <- read_csv(state$signature_genes, show_col_types = FALSE)
coverage   <- read_csv(state$gene_coverage, show_col_types = FALSE)
contrib    <- read_csv(state$gene_contributions, show_col_types = FALSE)
normtab    <- read_row_normalization(state$row_normalization)
scores     <- read_state_scores(state$row_scores)
pooled     <- read_state_scores(state$pooled_scores)
prep_index <- read_csv(state$prepared_index, show_col_types = FALSE)
taxonomy   <- read_csv(state$prepared_taxonomy, show_col_types = FALSE)
prep_prov  <- fromJSON(state$preparation_prov, simplifyVector = FALSE)
score_prov <- fromJSON(state$scoring_prov, simplifyVector = FALSE)

scores  <- scores  %>% mutate(domain = factor(domain, levels = DOMAIN_LEVELS))
normtab <- normtab %>% mutate(domain = factor(domain, levels = DOMAIN_LEVELS))

# Everything downstream distinguishes candidate state axes from the two control
# families, so a control can never be mistaken for a T variable.
role_of <- registry %>% select(signature_name, role, feature_origin, domain_reg = domain)
scores  <- scores %>% left_join(role_of, by = "signature_name")
STATE_SIGS <- registry %>% filter(role == "candidate_state", resolution_status == "resolved") %>% pull(signature_name)
NULL_SIGS  <- registry %>% filter(role == "null_control") %>% pull(signature_name)
ID_SIGS    <- registry %>% filter(role == "identity_control") %>% pull(signature_name)
```

---

## 0. What this notebook reads, and what it refuses to read

```{r}
#| label: inputs

inputs <- tibble::tribble(
  ~role, ~path,
  "prepared raw-count H5AD (4 lineages)", file.path(state$prepared_dir, "multiregion_2026"),
  "prepared row index",                   state$prepared_index,
  "released gene catalogue",              state$gene_catalog,
  "released supertype taxonomy",          state$prepared_taxonomy,
  "preparation provenance",               state$preparation_prov,
  "signature registry (derived)",         state$registry_csv,
  "signature genes (derived)",            state$signature_genes,
  "row-level normalization (derived)",    state$row_normalization,
  "row-level state scores (derived)",     state$row_scores,
  "pooled sensitivity scores (derived)",  state$pooled_scores,
  "published donor x region CPS",         omics$cps_by_region,
  "QNP workbook (P block only, section 12)", paths$qnp_2026
) %>%
  mutate(exists = file.exists(path) | dir.exists(path),
         size_mb = round(ifelse(file.exists(path), file.size(path) / 1e6, NA_real_), 2),
         file = basename(path)) %>%
  select(role, file, exists, size_mb)

show_table(inputs, "Everything this notebook reads. No new object was downloaded.")
```

Refused, deliberately:

- the **~3.2 GB neuronal pseudobulk set** (section 13);
- any SEA-AD model output that would have to be fetched (the `effect_size_table.csv` behind the
  published pseudoprogression gene dynamics, section 2);
- morphology, cognition and cognitive-resilience tables, which are out of scope for every
  decision in this notebook.

---

## 1. The measurement hierarchy

This is the single assumption that everything else rests on, so it is stated before any number.

```{r}
#| label: hierarchy

show_table(
  tibble::tribble(
    ~level, ~meaning, ~consequence,
    "donor",
      "biological replication unit",
      "the only level at which independent replication exists",
    "donor x region",
      "eventual regional analysis unit",
      "the grid the T block must eventually live on",
    "supertype",
      "molecular / cellular stratum",
      "a stratum to condition on, never a state measurement in itself",
    "sample / library-prep row",
      "repeated molecular measurement within a donor x region x supertype",
      "repeated measurements, not replicate donors; they inform reliability, not sample size"
  ),
  "The measurement hierarchy used throughout"
)
```

Two rules follow, and both are enforced structurally rather than by convention:

- **Repeated library/sample rows are never treated as independent donors.** Every group key below
  is `donor_id x brain_region x supertype`, and every reliability statistic is computed *within*
  that key.
- **Repeated rows are never summed before molecular-state scoring.** Normalization and scoring both
  happen at the released row grain. A pooled-count arm exists in section 9, is written to separate
  files, and is labelled a sensitivity analysis.

```{r}
#| label: hierarchy-observed

grain <- normtab %>%
  summarise(
    n_prepared_rows = n(),
    n_donors = n_distinct(donor_id),
    n_regions = n_distinct(brain_region),
    n_supertypes = n_distinct(supertype),
    n_sample_names = n_distinct(sample_name),
    n_library_preps = n_distinct(library_prep),
    n_donor_region_supertype = n_distinct(paste(donor_id, brain_region, supertype)),
    n_assay_methods = n_distinct(assay_method)
  )

show_table(grain, "The released grain, counted from the prepared rows")

repeats <- normtab %>%
  count(lineage, donor_id, brain_region, supertype, name = "n_rows") %>%
  count(lineage, n_rows, name = "n_groups")

show_table(
  repeats %>% pivot_wider(names_from = n_rows, values_from = n_groups, values_fill = 0) %>%
    rename_with(~ paste0("rows_", .x), .cols = -lineage),
  "How many donor x region x supertype groups carry how many sample/library rows"
)
```

```{r}
#| label: hierarchy-note

cat(prep_prov$aggregation_interpretation, "\n\n", prep_prov$sample_name_semantics, sep = "")
```

The preparation stage left repeated keys uncollapsed because `sample_name` semantics are
undocumented. That is exactly the right state for this notebook: those repeats are the only
internal information the release gives about **measurement error**, and sections 6 and 7 use them
as such.

---

## 2. Candidate molecular-state signatures

### 2.1 Feature-origin vocabulary

```{r}
#| label: feature-origin

show_table(
  tibble::tribble(
    ~feature_origin, ~meaning,
    "external_predefined",
      "a gene set defined outside this project and outside SEA-AD, applied unchanged",
    "SEAAD_published_predefined",
      "a quantity SEA-AD itself publishes as a named variable, label or gene list",
    "null_control_current_data",
      "a size-matched random gene set drawn from this data; a control, never a T variable"
  ),
  "Feature-origin categories"
)
```

### 2.2 How gene membership was resolved, and when it was refused

Gene membership for the resolvable candidates is copied verbatim from **MSigDB**, whose GMT files
are openly downloadable, versioned and hashed. Where a published definition could not be recovered
without guessing, the candidate is recorded with **no genes** and marked unresolved. No gene is
ever invented, completed from a paper's prose, or substituted from a similar set.

```{r}
#| label: registry-provenance

show_table(
  tibble(
    field = c("registry built (UTC)", "MSigDB release", "collections used", "rules"),
    value = c(
      registry_j$provenance$created_utc,
      registry_j$provenance$msigdb_release,
      paste(map_chr(registry_j$provenance$sources, "collection"), collapse = ", "),
      paste(unlist(registry_j$provenance$rules), collapse = " | ")
    )
  ),
  "Provenance of the resolved gene definitions"
)

show_table(
  map_dfr(registry_j$provenance$sources, ~ tibble(
    collection = .x$collection, n_sets = .x$n_sets,
    size_mb = round(.x$size_bytes / 1e6, 2), sha256 = substr(.x$sha256, 1, 16)
  )),
  "MSigDB source files, hashed"
)
```

### 2.3 The resolved candidates

```{r}
#| label: registry-resolved

show_table(
  registry %>%
    filter(resolution_status == "resolved", role == "candidate_state") %>%
    select(signature_name, biological_axis, lineage, domain, n_genes, directionality),
  "Resolved candidate state signatures: what each one measures"
)

show_table(
  registry %>%
    filter(resolution_status == "resolved", role == "candidate_state") %>%
    select(signature_name, source_publication, gene_definition_source, feature_origin),
  "Resolved candidate state signatures: where each definition came from"
)

show_table(
  registry %>%
    filter(role %in% c("identity_control", "null_control")) %>%
    select(signature_name, role, biological_axis, domain, feature_origin, n_genes),
  "Controls. Neither family is a candidate T variable."
)
```

The two **identity controls** exist because a gene set can score high simply by tracking how much
of the lineage's own identity programme a library recovered. If a state score behaves like the
identity control, it is measuring recovery, not state. The **null controls** are size-matched
random draws from each domain's detectably expressed genes and give the floor that a real signature
has to beat on reliability and on dynamic range.

### 2.4 The unresolved candidates, and why

```{r}
#| label: registry-unresolved

show_table(
  registry %>%
    filter(resolution_status == "unresolved") %>%
    select(signature_name, biological_axis, lineage, feature_origin, source_publication),
  "Candidates carried but NOT scored: no exact gene definition was recovered"
)

show_table(
  registry %>%
    filter(resolution_status == "unresolved") %>%
    select(signature_name, gene_definition_source, unresolved_reason),
  "Why each unresolved candidate is unresolved"
)
```

Three of these matter more than the others:

- **Sun et al. 2023 human microglial states (MG4 lipid-processing, MG8 inflammatory II).** These
  are the only *human, AD-cohort-derived, state-level* microglial definitions on the list. Their
  membership lives in a supplementary spreadsheet; the article text names two or three exemplar
  genes per state, which is not a signature. Resolving them is the highest-value remaining
  literature action.
- **Liddelow et al. 2017 PAN / A1 / A2 astrocyte cassettes.** The cassette membership appears only
  inside Extended Data Fig. 12a as an image; the article's tables list the whole qPCR panel, not
  the partition. Escartin et al. 2021 additionally advise against treating A1/A2 as a settled
  polarity, so the ontology activation axis in section 2.3 is the more defensible primary anyway.
- **The SEA-AD-published option.** SEA-AD publishes a per-gene, per-supertype **effect-size table**
  along its continuous pseudo-progression score, not a named gene set. Turning it into a signature
  would require choosing a threshold; it is MTG-only; and it is defined against CPS, which is also
  the QC axis in section 12, so a score built from it would be partly circular there. It is
  therefore recorded as an explicitly labelled secondary alternative and **not** retrieved, since
  this task performs no further acquisition.

The practical consequence is stated plainly: **there is currently no scored signature that is
simultaneously human, AD-derived and state-level.** Every scored candidate is an ontology- or
Hallmark-defined *biological axis*. That is a real limitation of what follows, not a presentational
one.

---

## 3. Gene-set availability in the released pseudobulk

Availability is computed twice, by two independent routes, and the two are reconciled: once in R
from `pseudobulk_gene_catalog.csv`, and once in Python by the extractor that actually built the
scores. Matching is **exact** on the released symbol. Nothing is fuzzy-matched, case-folded or
alias-resolved.

```{r}
#| label: gene-availability

catalog <- read_csv(state$gene_catalog, show_col_types = FALSE)

avail <- signature_availability(
  sig_genes %>% filter(signature_name %in% c(STATE_SIGS, ID_SIGS)),
  catalog,
  detect_fraction = score_prov$detectability$fraction
) %>%
  inner_join(registry %>% select(signature_name, lineage, role), by = c("signature_name", "lineage"))

show_table(
  avail %>%
    arrange(lineage, signature_name) %>%
    select(lineage, signature_name, n_defined, n_present, n_detectably_expressed, fraction_present),
  "Gene availability by the signature's own lineage (exact symbol match, R route)"
)
```

```{r}
#| label: gene-availability-missing

show_table(
  avail %>%
    filter(nchar(missing_genes) > 0) %>%
    arrange(lineage, signature_name) %>%
    select(lineage, signature_name, n_defined, n_present, missing_genes),
  "Exactly which defined symbols are absent from the released feature set"
)
```

```{r}
#| label: gene-availability-reconcile

reconcile <- coverage %>%
  filter(signature_name %in% c(STATE_SIGS, ID_SIGS)) %>%
  select(lineage, domain, signature_name,
         py_n_defined = n_defined, py_n_present = n_present,
         py_n_detectable = n_detectably_expressed, py_n_used = n_used) %>%
  left_join(avail %>% select(lineage, signature_name,
                             r_n_defined = n_defined, r_n_present = n_present,
                             r_n_detectable = n_detectably_expressed),
            by = c("lineage", "signature_name")) %>%
  mutate(defined_agrees = py_n_defined == r_n_defined,
         present_agrees = py_n_present == r_n_present)

show_table(
  reconcile %>% select(lineage, domain, signature_name, py_n_defined, r_n_defined,
                       py_n_present, r_n_present, defined_agrees, present_agrees),
  "Reconciliation of the two availability routes"
)

cat(sprintf(
  "n_defined agrees for %d of %d signature x lineage cells; n_present agrees for %d.\n",
  sum(reconcile$defined_agrees), nrow(reconcile), sum(reconcile$present_agrees)
))
```

The two routes differ on **detectability** by construction, and the difference is informative
rather than a discrepancy: the catalogue's `fraction_nonzero` is computed over *all* rows of a
released lineage, whereas the scoring pipeline computes it inside a **scoring domain** (section 4),
so the Immune numbers differ once Lymphocyte and Monocyte rows are excluded from the Micro/PVM
denominator.

```{r}
#| label: gene-availability-domain

show_table(
  coverage %>%
    arrange(domain, signature_name) %>%
    select(domain, signature_name, n_defined, n_present, n_detectably_expressed, n_used,
           fraction_present, fraction_detectably_expressed, n_rows_scored),
  "Availability as the scoring pipeline saw it, per scoring domain"
)
```

Read directly: coverage of the released feature set is high for every scored signature, and the
losses are concentrated in **detectability**, not in **presence**. A symbol that is present but
non-zero in under 10% of a domain's rows is dropped from that domain's score, which is why the
smallest sets lose proportionally the most.

---

## 4. Normalization at the existing pseudobulk-row level

### 4.1 The exact recipe

The released `X` is **summed raw UMI counts**, with no `layers` and no normalized copy, so a
conventional bulk-RNA workflow applies and a single-cell recipe would be wrong. `edgeR` and `limma`
are **not installed** in this project's `r-quarto` environment (it carries tidyverse, knitr and
rmarkdown only), so TMM is implemented directly in
[`src/seaad/omics/state_scoring.py`](../../src/seaad/omics/state_scoring.py) against the published
definition, with edgeR's defaults, rather than approximated by something else.

```{r}
#| label: normalization-recipe

norm <- score_prov$normalization
show_table(
  tibble(
    step = c("input", "library size", "method", "log-ratio trim", "abundance trim",
             "reference row", "rescaling", "log-CPM", "prior count",
             "single-cell normalization", "rows collapsed before normalization"),
    definition = c(
      norm$input, norm$library_size, norm$method,
      as.character(norm$logratio_trim), as.character(norm$sum_trim),
      norm$reference_selection, norm$rescaling, norm$log_cpm,
      as.character(norm$prior_count),
      ifelse(isTRUE(norm$single_cell_normalization_used), "yes", "no"),
      ifelse(isTRUE(score_prov$rows_collapsed_before_scoring), "yes", "no")
    )
  ),
  "The exact normalization, as recorded by the pipeline that ran it"
)
```

In symbols, for row $i$ and gene $g$, with $L_i$ the library size and $f_i$ the TMM factor:

$$
p_i = \text{prior} \cdot \frac{L_i f_i}{\overline{L f}}, \qquad
\text{logCPM}_{gi} = \log_2\!\left( 10^6 \cdot \frac{y_{gi} + p_i}{L_i f_i + 2 p_i} \right).
$$

### 4.2 Scoring domains, and why the Immune lineage is split

TMM assumes most genes are not differentially expressed between the rows being compared. Pooling
microglia with lymphocytes violates that, so normalization factors, detectability and gene
standardization are all computed inside a **scoring domain** rather than inside a released lineage.

```{r}
#| label: domains

show_table(
  map_dfr(names(registry_j$provenance$domains), function(d) {
    spec <- registry_j$provenance$domains[[d]]
    tibble(domain = d, lineage = spec$lineage, supertype_rule = spec$supertype_rule,
           taxonomy_note = spec$taxonomy_note)
  }),
  "Scoring domains. The released Immune taxonomy is preserved, not collapsed."
)

show_table(
  normtab %>%
    group_by(domain) %>%
    summarise(n_rows = n(), n_donors = n_distinct(donor_id), n_regions = n_distinct(brain_region),
              n_supertypes = n_distinct(supertype), .groups = "drop") %>%
    arrange(domain),
  "Rows per scoring domain"
)
```

### 4.3 What normalization had to absorb

```{r}
#| label: normalization-integrity

integrity <- normtab %>%
  summarise(
    rows = n(),
    library_size_equals_released_total_umi = all(computed_library_size == total_umi),
    min_library_size = min(computed_library_size),
    median_library_size = median(computed_library_size),
    max_library_size = max(computed_library_size),
    library_size_ratio = max(computed_library_size) / min(computed_library_size)
  )

show_table(integrity, "Integrity check: the recomputed library size reproduces the released total UMI exactly")

show_table(
  normtab %>%
    group_by(domain) %>%
    summarise(n = n(),
              tmm_min = min(tmm_factor), tmm_q25 = quantile(tmm_factor, .25),
              tmm_median = median(tmm_factor), tmm_q75 = quantile(tmm_factor, .75),
              tmm_max = max(tmm_factor), .groups = "drop") %>%
    arrange(domain),
  "TMM factors by domain"
)
```

```{r}
#| label: fig-tmm
#| fig-width: 8
#| fig-height: 4.2
#| fig-cap: "TMM factor against nuclei support. The factor is near 1 for well-supported rows and becomes both extreme and erratic below roughly ten nuclei, which is the first sign that low-support rows are a measurement problem rather than a biological one."

normtab %>%
  filter(domain %in% c("Micro/PVM", "Astrocyte", "Oligodendrocyte", "OPC")) %>%
  ggplot(aes(n_nuclei, tmm_factor)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_hline(yintercept = 1, colour = "firebrick", linewidth = 0.4) +
  scale_x_log10() + scale_y_log10() +
  facet_wrap(~ domain, nrow = 1) +
  labs(x = "Number of nuclei (log)", y = "TMM factor (log)")
```

The library-size range across rows is enormous, because a row can be one nucleus or two thousand.
That is the reason normalization is done *before* any aggregation decision: a raw-count sum across
repeated rows would silently weight each repeat by its own library size, which is precisely the
behaviour section 9 tests rather than assumes.

---

## 5. Row-level module scores

### 5.1 The scoring rule

```{r}
#| label: scoring-rule

show_table(
  tibble(
    field = c("primary score", "secondary score", "gene inclusion",
              "detectability rule", "standardization scope",
              "parameters tuned against pathology / morphology / cognition"),
    definition = c(
      score_prov$scoring$primary,
      score_prov$scoring$secondary,
      "present in the released features AND detectably expressed in the domain AND non-constant",
      sprintf("non-zero in at least %g of the domain's rows", score_prov$detectability$fraction),
      "within scoring domain, across all rows of that domain",
      ifelse(isTRUE(score_prov$scoring$parameters_tuned_against_outcomes), "yes", "no")
    )
  ),
  "The scoring rule. One unweighted mean, no tuned parameter."
)
```

For domain $D$, signature $S$ and row $i \in D$:

$$
z_{gi} = \frac{\text{logCPM}_{gi} - \operatorname{mean}_{j \in D} \text{logCPM}_{gj}}
               {\operatorname{sd}_{j \in D} \text{logCPM}_{gj}}, \qquad
\text{state\_score}_i = \frac{1}{|S_D|}\sum_{g \in S_D} z_{gi}.
$$

The secondary score, `score_logcpm_mean`, is the same mean without the standardization, so any
conclusion that depends on standardizing is visible as such.

### 5.2 Scores exist at the released row grain, with their support carried

```{r}
#| label: score-grain

show_table(
  scores %>%
    filter(signature_name %in% STATE_SIGS) %>%
    group_by(domain, signature_name) %>%
    summarise(n_rows = n(), n_donors = n_distinct(donor_id),
              n_donor_region_supertype = n_distinct(paste(donor_id, brain_region, supertype)),
              n_genes_used = first(n_genes_used),
              mean = mean(score_z_mean), sd = sd(score_z_mean),
              min = min(score_z_mean), max = max(score_z_mean), .groups = "drop") %>%
    arrange(domain, signature_name),
  "Row-level state scores: one per donor x region x supertype x sample/library"
)

show_table(
  scores %>% slice_head(n = 3) %>%
    select(prepared_obs_index, donor_id, brain_region, supertype, sample_name, library_prep,
           assay_method, n_nuclei, total_umi, signature_name, score_z_mean),
  "Three rows, to show that every score keeps its identity and its support"
)
```

Every score row carries `donor_id`, `brain_region`, `supertype`, `sample_name`, `library_prep`,
`assay_method`, `Number of nuclei` and `total_umi`, exactly as required, so no downstream step has
to re-derive the grain.

### 5.3 Is a score dominated by one or two genes?

```{r}
#| label: gene-contributions

domination <- contrib %>%
  group_by(domain, signature_name) %>%
  summarise(
    n_genes = n(),
    max_gene_corr = max(corr_gene_z_with_score, na.rm = TRUE),
    top_gene = gene_symbol[which.max(corr_gene_z_with_score)],
    median_gene_corr = median(corr_gene_z_with_score, na.rm = TRUE),
    min_leave_one_out_corr = min(leave_one_out_corr, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(family = case_when(signature_name %in% NULL_SIGS ~ "null control",
                            signature_name %in% ID_SIGS ~ "identity control",
                            TRUE ~ "candidate state"))

show_table(
  domination %>% arrange(family, domain, desc(max_gene_corr)) %>%
    select(family, domain, signature_name, n_genes, top_gene, max_gene_corr,
           median_gene_corr, min_leave_one_out_corr),
  "Gene-level domination diagnostics. A leave-one-out correlation near 1 means no single gene carries the score."
)
```

`leave_one_out_corr` is the correlation between the score and the same score recomputed without
that gene. Values at or near 1 everywhere mean the mean is genuinely distributed over its genes; a
value that fell away from 1 would identify a signature that is really a single-gene readout. The
smallest sets are where to look, because they have the least room to average.

```{r}
#| label: fig-contributions
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Per-gene correlation with the signature score it belongs to. Candidate state signatures are compared against the size-matched null draws, which show what correlation structure arises from gene count alone."

contrib %>%
  mutate(family = case_when(signature_name %in% NULL_SIGS ~ "null control",
                            signature_name %in% ID_SIGS ~ "identity control",
                            TRUE ~ "candidate state")) %>%
  ggplot(aes(reorder(signature_name, corr_gene_z_with_score, median), corr_gene_z_with_score,
             colour = family)) +
  geom_boxplot(outlier.size = 0.5, linewidth = 0.35) +
  coord_flip() +
  labs(x = NULL, y = "correlation of gene z with the signature score", colour = NULL) +
  theme(legend.position = "top")
```

---

## 6. Repeated rows as measurement-reliability data

This is the section the rest of the notebook depends on. It answers:

> Can repeated library/sample measurements be combined after scoring, and does the combination
> require adjustment for assay method?

using only the internal structure of the release. No morphology, no cognition.

### 6.1 The replicate structure that exists

```{r}
#| label: replicate-structure

rep_structure <- scores %>%
  filter(signature_name %in% STATE_SIGS) %>%
  distinct(domain, donor_id, brain_region, supertype, prepared_obs_index, assay_method) %>%
  group_by(domain, donor_id, brain_region, supertype) %>%
  summarise(n_rows = n(), n_methods = n_distinct(assay_method), .groups = "drop") %>%
  mutate(kind = case_when(n_rows == 1 ~ "single measurement",
                          n_methods == 1 ~ "repeated, same assay method",
                          TRUE ~ "repeated, mixed assay method"))

show_table(
  rep_structure %>% count(domain, kind) %>%
    pivot_wider(names_from = kind, values_from = n, values_fill = 0) %>% arrange(domain),
  "Donor x region x supertype groups by replicate structure"
)

show_table(
  scores %>% distinct(prepared_obs_index, assay_method, domain) %>% count(domain, assay_method) %>%
    pivot_wider(names_from = assay_method, values_from = n, values_fill = 0) %>% arrange(domain),
  "Prepared rows by released assay method"
)
```

The mixed-method groups are the valuable ones: within a mixed-method group the donor, the region
and the supertype are held fixed by construction, so a difference between the two rows cannot be a
biological contrast between donors. It is measurement.

### 6.2 Agreement between repeated measurements

```{r}
#| label: pairs

pairs <- scores %>%
  filter(signature_name %in% c(STATE_SIGS, ID_SIGS, NULL_SIGS)) %>%
  replicate_pairs(value_col = "score_z_mean") %>%
  mutate(family = case_when(signature_name %in% NULL_SIGS ~ "null control",
                            signature_name %in% ID_SIGS ~ "identity control",
                            TRUE ~ "candidate state"))

cat(sprintf("%s within-group replicate pairs across %s signatures.\n",
            format(nrow(pairs), big.mark = ","), n_distinct(pairs$signature_name)))

show_table(
  pairs %>% agreement_summary(by = c("family", "domain", "signature_name", "assay_pair")) %>%
    arrange(family, domain, signature_name, assay_pair) %>%
    select(family, domain, signature_name, assay_pair, n_pairs, median_abs_difference,
           mean_difference, repeatability_sd, loa_lower, loa_upper),
  "Bland-Altman style agreement of repeated sample/library measurements, by assay-method match"
)
```

`repeatability_sd` is the within-group measurement standard deviation implied by the pairwise
differences, in score units. Because the primary score is standardized within domain, the whole
between-row spread of a signature is close to 1, so a repeatability SD can be read directly as a
fraction of the total spread.

### 6.3 Intraclass correlation

```{r}
#| label: icc

icc_tbl <- scores %>%
  filter(signature_name %in% c(STATE_SIGS, ID_SIGS, NULL_SIGS)) %>%
  group_by(domain, signature_name) %>%
  group_modify(~ icc_oneway(.x$score_z_mean, paste(.x$donor_id, .x$brain_region, .x$supertype))) %>%
  ungroup() %>%
  mutate(family = case_when(signature_name %in% NULL_SIGS ~ "null control",
                            signature_name %in% ID_SIGS ~ "identity control",
                            TRUE ~ "candidate state"))

show_table(
  icc_tbl %>% arrange(family, domain, desc(icc)) %>%
    select(family, domain, signature_name, n_groups, n_rows, icc, var_between, var_within),
  "One-way ICC(1,1) over donor x region x supertype groups. Replicates are repeated measurements, not donors."
)
```

```{r}
#| label: fig-icc
#| fig-width: 7.5
#| fig-height: 4
#| fig-cap: "ICC(1,1) per signature. The dashed line marks the size-matched null-control level for the same domain; a candidate state signature is only worth carrying if it clears it."

null_level <- icc_tbl %>% filter(family == "null control") %>% select(domain, null_icc = icc)

icc_tbl %>%
  filter(family != "null control") %>%
  left_join(null_level, by = "domain") %>%
  ggplot(aes(reorder(signature_name, icc), icc, fill = family)) +
  geom_col(width = 0.6) +
  geom_hline(aes(yintercept = null_icc), linetype = 2, colour = "firebrick", linewidth = 0.35) +
  coord_flip() +
  facet_wrap(~ domain, scales = "free_y", ncol = 2) +
  labs(x = NULL, y = "ICC(1,1)", fill = NULL) +
  theme(legend.position = "top")
```

The null controls are the important column here. A **size-matched random gene set reproduces at
least as well as every candidate state signature in its domain**, and in the astrocyte domain the
random 18-gene draw reproduces slightly better than the 18-gene activation axis. Two things follow,
and they are different:

- Row-level scores *are* reproducible across repeated library measurements. That is a real property
  of the pipeline, and it is what sections 9 and 10 need.
- That reproducibility is **not signature-specific**. Most of the between-group variance any of
  these scores carries is shared structure that a donor x region x supertype imposes on any
  sufficiently large gene set: overall recovery, complexity, composition inside the supertype. A
  high ICC is therefore evidence about the *measurement*, not evidence that the signature is
  tracking its intended biology.

The direct test of signature specificity is whether a candidate score is distinguishable from its
domain's null draw at the row level.

```{r}
#| label: signature-specificity

null_by_domain <- scores %>%
  filter(signature_name %in% NULL_SIGS) %>%
  select(prepared_obs_index, domain, null_score = score_z_mean)

specificity <- scores %>%
  filter(signature_name %in% c(STATE_SIGS, ID_SIGS)) %>%
  select(prepared_obs_index, domain, signature_name, score_z_mean) %>%
  inner_join(null_by_domain, by = c("prepared_obs_index", "domain")) %>%
  group_by(domain, signature_name) %>%
  summarise(n_rows = n(),
            pearson_with_null = cor(score_z_mean, null_score),
            .groups = "drop") %>%
  left_join(registry %>% select(signature_name, role, n_genes), by = "signature_name") %>%
  arrange(domain, desc(pearson_with_null))

show_table(
  specificity %>% select(domain, signature_name, role, n_genes, n_rows, pearson_with_null),
  "How much of each score is the generic axis the null control also measures"
)
```

A correlation near 1 would mean the signature is indistinguishable from a random gene set of the
same size, i.e. a pure recovery axis. Values below that leave room for signature-specific variation,
and the residual room is what a later disease-state analysis would actually be using. This table,
together with the ICC table, is the honest statement of what the current signatures can and cannot
support.

### 6.4 Do the differences depend on support or on assay method?

```{r}
#| label: fig-bland-altman
#| fig-width: 8.5
#| fig-height: 4.5
#| fig-cap: "Difference between paired repeated measurements against the weaker member's nuclei count. If replicate disagreement were biological, it would not funnel with support."

pairs %>%
  filter(family == "candidate state") %>%
  ggplot(aes(support_min, difference, colour = assay_pair)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_point(alpha = 0.25, size = 0.5) +
  scale_x_log10() +
  facet_wrap(~ domain, nrow = 1) +
  labs(x = "nuclei in the weaker member of the pair (log)",
       y = "difference in state score", colour = NULL) +
  theme(legend.position = "top")
```

```{r}
#| label: difference-drivers

drivers <- pairs %>%
  filter(family == "candidate state") %>%
  group_by(domain, signature_name) %>%
  summarise(
    n_pairs = n(),
    spearman_absdiff_vs_min_nuclei = cor(abs_difference, support_min, method = "spearman"),
    spearman_absdiff_vs_min_umi = cor(abs_difference, umi_min, method = "spearman"),
    median_absdiff_same_method = median(abs_difference[assay_pair == "same method"]),
    median_absdiff_mixed_method = median(abs_difference[assay_pair == "mixed method"]),
    .groups = "drop"
  ) %>%
  mutate(mixed_minus_same = median_absdiff_mixed_method - median_absdiff_same_method)

show_table(
  drivers %>% arrange(domain, signature_name),
  "What drives replicate disagreement: support, and assay-method match"
)
```

A negative rank correlation between `abs_difference` and support means disagreement shrinks as
support grows, which is the signature of measurement noise rather than biology.

### 6.5 Is there a systematic assay-method offset?

```{r}
#| label: assay-design

design <- pairs %>%
  filter(family == "candidate state", assay_pair == "mixed method") %>%
  distinct(domain, donor_id, brain_region, supertype, method_a, method_b) %>%
  mutate(pair = paste(pmin(method_a, method_b), pmax(method_a, method_b), sep = " vs ")) %>%
  count(domain, pair)

show_table(design %>% pivot_wider(names_from = pair, values_from = n, values_fill = 0),
           "Which method pairs are actually observed inside a donor x region x supertype")
```

The replicate design is a **chain**, not a star: some methods never share a
donor x region x supertype with the reference. Offsets are therefore estimated on the directly
observed contrasts and then propagated along that chain, with the number of steps recorded so a
chained estimate is never mistaken for a direct one.

```{r}
#| label: assay-offsets

contrasts <- pairs %>% filter(family == "candidate state") %>% method_pair_contrasts()

show_table(
  contrasts %>% arrange(domain, signature_name, first_method, second_method) %>%
    select(domain, signature_name, first_method, second_method, n_pairs,
           median_offset, mean_offset, ci_lower, ci_upper, wilcoxon_p),
  "Directly observed assay-method contrasts, estimated only within donor x region x supertype"
)

offsets <- method_offsets(pairs %>% filter(family == "candidate state"), reference = "10Xv3.1")

show_table(
  offsets %>% arrange(signature_name, assay_method) %>%
    select(signature_name, reference_method, assay_method, method_offset, n_steps_from_reference),
  "Additive offsets relative to 10Xv3.1, with the chain length behind each one"
)

show_table(
  offsets %>% group_by(assay_method) %>%
    summarise(n_signatures = n(), n_estimable = sum(!is.na(method_offset)),
              median_offset = median(method_offset, na.rm = TRUE),
              steps = paste(sort(unique(n_steps_from_reference)), collapse = ","), .groups = "drop"),
  "Offset coverage by method. A method with no path to the reference would appear as not estimable."
)
```

Every one of the 26 directly observed contrasts points the same way: **`10xMulti` reads lower than
both singleome methods**, by roughly 0.1 to 0.4 score units, across all four domains and all
thirteen signatures. That is a platform effect on a score, not a biological contrast, and because it
is one-directional it does not average out when repeated measurements are combined.

```{r}
#| label: fig-assay
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "Within-group assay-method contrasts, for every directly observed method pair. Donor, region and supertype are fixed inside every pair, so a non-zero centre is a method effect and not a cohort effect."

pairs %>%
  filter(family == "candidate state", assay_pair == "mixed method") %>%
  mutate(pair = paste(pmin(method_a, method_b), pmax(method_a, method_b), sep = " -> "),
         signed = if_else(method_a == pmax(method_a, method_b),
                          value_a - value_b, value_b - value_a)) %>%
  ggplot(aes(signed, signature_name, colour = pair)) +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  geom_boxplot(outlier.size = 0.4, linewidth = 0.35) +
  labs(x = "score(second method) - score(first method), within donor x region x supertype",
       y = NULL, colour = NULL) +
  theme(legend.position = "top")
```

---

## 7. Low-support observations

No minimum-nucleus cutoff is imposed. Instead the reliability curve is described, and only then is
a candidate rule proposed. Nothing is filtered in this notebook.

```{r}
#| label: support-curve

support <- pairs %>%
  filter(family == "candidate state") %>%
  support_reliability(support_col = "support_min")

show_table(
  support %>% filter(n_pairs >= 5) %>% arrange(domain, signature_name, support_bin),
  "Replicate disagreement by nuclei support of the weaker member (bins with at least 5 pairs)"
)
```

```{r}
#| label: fig-support
#| fig-width: 8.5
#| fig-height: 4.5
#| fig-cap: "Repeatability standard deviation against nuclei support. The score is standardized within domain, so a repeatability SD of 1 means the repeated measurement carries no information about the group at all."

support %>%
  filter(n_pairs >= 5) %>%
  ggplot(aes(support_bin, repeatability_sd, group = signature_name, colour = signature_name)) +
  geom_line(linewidth = 0.4) + geom_point(size = 0.8) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40", linewidth = 0.3) +
  facet_wrap(~ domain, nrow = 1, scales = "free_x") +
  labs(x = "nuclei in the weaker member of the pair", y = "repeatability SD (score units)",
       colour = NULL) +
  theme(legend.position = "bottom", legend.text = element_text(size = 6),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

```{r}
#| label: support-umi

support_umi <- pairs %>%
  filter(family == "candidate state") %>%
  mutate(umi_bin = cut(umi_min, breaks = c(0, 1e3, 1e4, 3e4, 1e5, 3e5, 1e6, Inf),
                       include.lowest = TRUE, dig.lab = 8)) %>%
  group_by(domain, umi_bin) %>%
  summarise(n_pairs = n(), median_abs_difference = median(abs_difference),
            repeatability_sd = sd(difference) / sqrt(2), .groups = "drop")

show_table(support_umi %>% arrange(domain, umi_bin),
           "The same curve against total UMI of the weaker member, pooled over candidate signatures")
```

```{r}
#| label: support-threshold

# A descriptive, not-yet-applied candidate rule: the smallest pre-declared bin at
# which the repeatability SD of every candidate signature in that domain first
# falls below half the within-domain spread of the score.
# Two criteria, not one, because a single cut-off would hide how sensitive the
# answer is to where the line is drawn. The score is standardized within domain,
# so a repeatability SD of r corresponds to a reliability of about 1 - r^2/2.
candidate_rule <- support %>%
  filter(n_pairs >= 5) %>%
  group_by(domain, support_bin) %>%
  summarise(worst_repeatability_sd = max(repeatability_sd, na.rm = TRUE),
            n_signatures = n_distinct(signature_name), n_pairs = sum(n_pairs), .groups = "drop") %>%
  arrange(domain, support_bin) %>%
  mutate(meets_lenient = worst_repeatability_sd < 0.5,
         meets_strict = worst_repeatability_sd < 0.3)

show_table(candidate_rule, "Repeatability of the worst candidate signature in each domain, by support bin")

first_bin <- function(column) {
  candidate_rule %>% filter(.data[[column]]) %>% group_by(domain) %>% slice(1) %>% ungroup() %>%
    transmute(domain, criterion = column, first_bin = support_bin,
              worst_repeatability_sd, n_pairs)
}
first_pass <- bind_rows(first_bin("meets_lenient"), first_bin("meets_strict")) %>%
  arrange(domain, criterion)

show_table(first_pass,
           "Empirical candidate support thresholds under two criteria. NOT applied anywhere in this notebook.")
```

```{r}
#| label: support-consequence

consequence <- scores %>%
  filter(signature_name %in% STATE_SIGS) %>%
  distinct(domain, prepared_obs_index, donor_id, brain_region, supertype, n_nuclei) %>%
  mutate(across(n_nuclei, as.numeric)) %>%
  group_by(domain) %>%
  summarise(
    n_rows = n(),
    rows_under_5 = sum(n_nuclei < 5), rows_under_10 = sum(n_nuclei < 10),
    rows_under_20 = sum(n_nuclei < 20),
    pct_under_10 = 100 * mean(n_nuclei < 10),
    donor_region_lost_at_10 = {
      keep <- n_nuclei >= 10
      n_distinct(paste(donor_id, brain_region)[!keep]) -
        n_distinct(intersect(paste(donor_id, brain_region)[!keep], paste(donor_id, brain_region)[keep]))
    },
    .groups = "drop"
  )

show_table(consequence,
           "What a threshold would cost. Reported so the cost is visible before any rule is adopted.")
```

The point of this section is that **support is a measurement covariate, not a nuisance to be
thresholded away by convention**. The curve above is descriptive; any rule adopted later should be
justified by it and applied once, explicitly, with the cost table above recomputed.

---

## 8. Supertype dependence

Molecular state and supertype composition are different quantities and are kept apart here. A
supertype is a **stratum**; how much of a lineage sits in a given supertype is **composition**
(section 11), and what the cells in that supertype are expressing is **state**.

### 8.1 The released Immune taxonomy is preserved

```{r}
#| label: immune-taxonomy

show_table(
  taxonomy %>% filter(lineage == "Immune") %>%
    select(supertype, n_donors, n_regions, total_nuclei, taxonomy_interpretation, semantic_status),
  "Released Immune supertypes, exactly as the release names them"
)
```

`Lymphocyte` and `Monocyte` are **not** merged into the resident-myeloid axis: they are separate
scoring domains, and the primary resident-myeloid state is computed over `Micro-PVM_*` rows only.
Within that family the release does not separate microglia from perivascular macrophages, so the
axis is labelled with the conservative biological label **Micro/PVM** and is never described as
pure microglia.

```{r}
#| label: monocyte-caution

show_table(
  normtab %>% filter(domain %in% c("Micro/PVM", "Lymphocyte", "Monocyte")) %>%
    group_by(domain) %>%
    summarise(n_rows = n(), n_donors = n_distinct(donor_id),
              median_nuclei = median(n_nuclei), max_nuclei = max(n_nuclei),
              median_umi = median(total_umi), .groups = "drop"),
  "Why the Immune split matters: the non-myeloid domains are tiny and would otherwise dilute the axis"
)
```

### 8.2 State score by supertype

```{r}
#| label: fig-supertype
#| fig-width: 9
#| fig-height: 7
#| fig-cap: "Row-level state scores by released supertype. Spread across supertypes within a domain is a reminder that the supertype is a stratum: a donor-level score computed without conditioning on it would partly re-measure composition."

scores %>%
  filter(signature_name %in% STATE_SIGS) %>%
  ggplot(aes(score_z_mean, supertype)) +
  geom_boxplot(outlier.size = 0.3, linewidth = 0.3) +
  facet_wrap(~ signature_name, scales = "free", ncol = 3) +
  labs(x = "state score (within-domain z mean)", y = NULL) +
  theme(strip.text = element_text(size = 7), axis.text.y = element_text(size = 6))
```

```{r}
#| label: supertype-variance

variance_share <- scores %>%
  filter(signature_name %in% c(STATE_SIGS, NULL_SIGS)) %>%
  group_by(domain, signature_name) %>%
  group_modify(function(.x, .y) {
    fit_st <- summary(aov(score_z_mean ~ supertype, data = .x))[[1]]
    fit_rg <- summary(aov(score_z_mean ~ brain_region, data = .x))[[1]]
    tibble(
      supertype_variance_share = fit_st[["Sum Sq"]][1] / sum(fit_st[["Sum Sq"]]),
      region_variance_share    = fit_rg[["Sum Sq"]][1] / sum(fit_rg[["Sum Sq"]]),
      n_rows = nrow(.x)
    )
  }) %>%
  ungroup() %>%
  mutate(family = if_else(signature_name %in% NULL_SIGS, "null control", "candidate state"))

show_table(
  variance_share %>% arrange(family, domain, desc(supertype_variance_share)),
  "Share of row-level score variance explained by supertype and, separately, by region"
)
```

These are **marginal** shares from two separate one-way decompositions, not a partition of a joint
model; they are reported to size the effects, not to attribute them. Two readings matter. First,
supertype dominates region by roughly an order of magnitude, so the supertype has to stay a stratum
rather than be averaged over on the way to a donor x region value. Second, the **null controls carry
almost the same supertype share as the candidates**, which says again that most of this structure is
generic: supertypes differ in recovery and complexity, and any gene set picks that up. Region shares
are also attenuated here because the denominator is row-level variance, which includes the
measurement error quantified in section 6; a donor x region-level decomposition would put region
higher.

### 8.3 State score by region

```{r}
#| label: fig-region
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Row-level state scores by brain region, for the candidate state signatures."

scores %>%
  filter(signature_name %in% STATE_SIGS) %>%
  ggplot(aes(score_z_mean, brain_region)) +
  geom_boxplot(outlier.size = 0.3, linewidth = 0.3) +
  facet_wrap(~ signature_name, scales = "free_x", ncol = 3) +
  labs(x = "state score (within-domain z mean)", y = NULL) +
  theme(strip.text = element_text(size = 7))
```

```{r}
#| label: state-vs-identity

identity_vs_state <- scores %>%
  filter(signature_name %in% c(STATE_SIGS, ID_SIGS)) %>%
  select(prepared_obs_index, domain, signature_name, score_z_mean) %>%
  pivot_wider(names_from = signature_name, values_from = score_z_mean)

id_corr <- map_dfr(intersect(STATE_SIGS, names(identity_vs_state)), function(s) {
  dom <- registry$domain[match(s, registry$signature_name)]
  control <- registry %>% filter(role == "identity_control", domain == dom) %>% pull(signature_name)
  if (!length(control) || !control %in% names(identity_vs_state)) return(NULL)
  d <- identity_vs_state %>% filter(domain == dom)
  tibble(domain = dom, signature_name = s, identity_control = control,
         pearson_with_identity = cor(d[[s]], d[[control]], use = "complete.obs"))
})

show_table(id_corr, "Is a candidate state score really a lineage-identity readout?")
```

A state score that correlated near 1 with its lineage-identity control would be measuring how much
of the lineage programme the library recovered, not a disease-relevant state. This table is the
check; it is reported for every candidate, not only where it looks good, and the result is
unflattering: **the three oligodendrocyte myelination axes sit at 0.91 to 0.96 against the
oligodendrocyte identity control.** Myelin genes are oligodendrocyte identity genes, so that is
biologically unsurprising, but it means those axes cannot be read as a myelination *state*
independent of how much oligodendrocyte programme the library recovered. The astrocyte activation
axis is lower at 0.85, and the small regulation set lower again at 0.63.

---

## 9. Aggregation policies, compared after scoring

Five policies are compared. None is chosen by its relationship to any downstream biological
outcome; the comparison is agreement and robustness only.

```{r}
#| label: policies

policy_defs <- tibble::tribble(
  ~policy, ~definition, ~what_it_assumes,
  "equal_weight_mean",
    "unweighted mean of the normalized state scores of the repeated rows",
    "each repeated measurement is an equally good estimate of the same quantity",
  "nucleus_weighted_mean",
    "mean weighted by Number of nuclei",
    "a measurement's precision scales with the nuclei behind it",
  "raw-count pooled score",
    "sum raw counts across repeated rows, then normalize and score once",
    "the repeats are one experiment; implicitly weights by library size (SENSITIVITY ONLY)",
  "assay_adjusted_mean",
    "subtract a per-method additive offset, then take the unweighted mean",
    "assay method shifts the score additively by a fixed amount",
  "assay_adjusted_nucleus_weighted_mean",
    "subtract the offset, then weight by Number of nuclei",
    "both of the above at once: additive method shift, precision rising with nuclei"
)

show_table(policy_defs, "The five aggregation policies")
```

```{r}
#| label: aggregate

meth_off <- offsets %>% select(signature_name, assay_method, method_offset)

# A method with no estimable offset is adjusted by zero, which is an assumption,
# so its size is reported rather than left implicit.
unadjusted <- scores %>%
  filter(signature_name %in% STATE_SIGS) %>%
  left_join(meth_off, by = c("signature_name", "assay_method")) %>%
  summarise(rows = n(), rows_without_an_estimable_offset = sum(is.na(method_offset)))

show_table(unadjusted, "How many scored rows the assay adjustment cannot reach")

agg <- scores %>%
  filter(signature_name %in% STATE_SIGS) %>%
  aggregate_policies(offsets = meth_off, value_col = "score_z_mean")

pooled_wide <- pooled %>%
  filter(signature_name %in% STATE_SIGS) %>%
  select(donor_id, brain_region, supertype, signature_name, pooled_score = score_z_mean)

agg <- agg %>% left_join(pooled_wide, by = c("donor_id", "brain_region", "supertype", "signature_name"))

cat(sprintf("%s donor x region x supertype x signature cells; %s of them from more than one measurement.\n",
            format(nrow(agg), big.mark = ","),
            format(sum(agg$n_measurements > 1), big.mark = ",")))
```

The pooled arm is standardized against the *pooled* row set, which is a different reference
population from the row-level one. Its **rank** agreement with the other policies is therefore the
interpretable quantity; an absolute difference against it also carries that change of reference and
is reported only for completeness.

```{r}
#| label: policy-agreement

multi <- agg %>% filter(n_measurements > 1)

agreement <- bind_rows(
  policy_agreement(multi, "equal_weight_mean", "nucleus_weighted_mean"),
  policy_agreement(multi, "equal_weight_mean", "umi_weighted_mean"),
  policy_agreement(multi, "equal_weight_mean", "assay_adjusted_mean"),
  policy_agreement(multi, "equal_weight_mean", "assay_adjusted_nucleus_weighted_mean"),
  policy_agreement(multi, "equal_weight_mean", "pooled_score"),
  policy_agreement(multi, "nucleus_weighted_mean", "pooled_score")
)

show_table(
  agreement %>% arrange(policy_a, policy_b, signature_name),
  "Agreement between aggregation policies, over groups that actually have repeated measurements"
)
```

```{r}
#| label: fig-policies
#| fig-width: 8
#| fig-height: 4.2
#| fig-cap: "Equal-weight mean against the three alternatives, over groups with repeated measurements. Deviation from the diagonal is where the policy choice would change a donor x region x supertype value."

multi %>%
  select(signature_name, equal_weight_mean, nucleus_weighted_mean, assay_adjusted_mean,
         assay_adjusted_nucleus_weighted_mean, pooled_score) %>%
  pivot_longer(-c(signature_name, equal_weight_mean), names_to = "policy", values_to = "value") %>%
  filter(!is.na(value)) %>%
  ggplot(aes(equal_weight_mean, value)) +
  geom_abline(slope = 1, intercept = 0, colour = "firebrick", linewidth = 0.35) +
  geom_point(alpha = 0.15, size = 0.5) +
  facet_wrap(~ policy, nrow = 1) +
  labs(x = "equal-weight mean of normalized scores", y = "alternative policy")
```

```{r}
#| label: policy-robustness

robustness <- multi %>%
  group_by(signature_name) %>%
  summarise(
    n_groups = n(),
    sd_equal = sd(equal_weight_mean),
    sd_nucleus = sd(nucleus_weighted_mean),
    sd_assay_adjusted = sd(assay_adjusted_mean, na.rm = TRUE),
    sd_assay_adjusted_nucleus = sd(assay_adjusted_nucleus_weighted_mean, na.rm = TRUE),
    sd_pooled = sd(pooled_score, na.rm = TRUE),
    median_within_group_sd = median(within_group_sd, na.rm = TRUE),
    .groups = "drop"
  )

show_table(robustness,
           "Spread of each aggregated estimate, beside the median within-group measurement SD it has to average over")
```

```{r}
#| label: policy-influence

influence <- multi %>%
  mutate(shift_nucleus = abs(nucleus_weighted_mean - equal_weight_mean),
         shift_pooled = abs(pooled_score - equal_weight_mean),
         shift_assay = abs(assay_adjusted_mean - equal_weight_mean),
         nuclei_imbalance = n_nuclei_total / n_measurements) %>%
  group_by(signature_name) %>%
  summarise(
    spearman_shift_nucleus_vs_imbalance = cor(shift_nucleus, nuclei_imbalance, method = "spearman",
                                              use = "complete.obs"),
    median_shift_nucleus = median(shift_nucleus, na.rm = TRUE),
    median_shift_pooled = median(shift_pooled, na.rm = TRUE),
    median_shift_assay = median(shift_assay, na.rm = TRUE),
    .groups = "drop"
  )

show_table(influence, "How far each alternative moves the estimate away from the equal-weight mean")
```

### 9.1 What the reliability results actually support

Three facts from sections 6 and 7 constrain the choice, and none of them involves an outcome:

1. **The assay offset is systematic, one-directional and present in every signature.** `10xMulti`
   scores below `10Xv3.1` in all thirteen contrasts, at p effectively zero, by roughly 0.1 to 0.4
   score units. A systematic shift does not average out; it biases whichever
   donor x region x supertype happens to be measured on the other platform.
2. **Repeated measurements are not equally precise.** Repeatability SD falls by roughly an order of
   magnitude from the lowest to the highest support bin, so treating two rows as interchangeable is
   contradicted by the data whenever their support differs.
3. **The policies nevertheless agree closely** (Pearson 0.90 to 0.998 depending on the pair), so the
   choice changes few values — but it changes them exactly where support or platform is unbalanced,
   which is where the estimate is weakest to begin with.

The simplest rule consistent with all three is the **nucleus-weighted mean of assay-adjusted
row-level scores**. It is one weighted mean and one additive offset, with no tuned parameter. The
unweighted assay-adjusted mean is a defensible alternative *once* a support rule is applied, because
removing the least-supported rows is what makes equal weighting reasonable; the two are reported
side by side rather than one being hidden. The raw-count pooled score is **not** recommended: it
agrees least with every other policy, and it implicitly weights by library size, which is the one
weighting no reliability result here supports.

**Row-level scores are not overwritten.** `row_state_scores.csv` remains the authoritative
row-grain file; the aggregated table is a separate derivative, and the pooled arm lives in
`pooled_state_scores.csv`.

---

## 10. Provisional donor x region x supertype state table

```{r}
#| label: provisional-table

provisional <- agg %>%
  left_join(registry %>% select(signature_name, feature_origin), by = "signature_name") %>%
  mutate(
    source_release = prep_prov$source_release,
    aggregation_method = "nucleus_weighted_mean_of_assay_adjusted_row_level_normalized_scores",
    state_score = assay_adjusted_nucleus_weighted_mean,
    support_status = case_when(
      n_nuclei_total >= 20 ~ "well supported",
      n_nuclei_total >= 10 ~ "marginal support",
      TRUE ~ "low support"
    )
  ) %>%
  select(donor_id, brain_region, lineage, supertype, signature_name, state_score,
         n_measurements, n_nuclei_total, aggregation_method, support_status,
         source_release, feature_origin,
         domain, equal_weight_mean, nucleus_weighted_mean, assay_adjusted_mean, pooled_score,
         within_group_sd, n_assay_methods, assay_methods, total_umi_total)

show_table(
  provisional %>% slice_head(n = 8) %>%
    select(donor_id, brain_region, lineage, supertype, signature_name, state_score,
           n_measurements, n_nuclei_total, aggregation_method, support_status, feature_origin),
  "The provisional donor x region x supertype state table, first eight rows"
)

show_table(
  provisional %>%
    group_by(lineage, signature_name) %>%
    summarise(n_cells = n(), n_donors = n_distinct(donor_id),
              n_donor_region = n_distinct(paste(donor_id, brain_region)),
              well_supported = sum(support_status == "well supported"),
              marginal = sum(support_status == "marginal support"),
              low = sum(support_status == "low support"), .groups = "drop"),
  "Size of the provisional table, and how much of it is well supported"
)

show_table(
  provisional %>% filter(brain_region %in% ROI) %>%
    count(brain_region, support_status) %>%
    pivot_wider(names_from = support_status, values_from = n, values_fill = 0),
  "The same, restricted to the candidate ROI set (which is still not frozen)"
)
```

This table is an **exploratory derivative, not a final model matrix**. The support bands are
descriptive labels attached to every row; nothing has been dropped.

---

## 11. Composition, kept separate

```{r}
#| label: composition

comp <- composition_table(normtab)

show_table(
  comp %>% slice_head(n = 8) %>%
    select(donor_id, brain_region, lineage, supertype, n_nuclei, relative_abundance),
  "The compositional table, first eight rows: donor x region x lineage x supertype"
)

show_table(
  comp %>% group_by(lineage) %>%
    summarise(n_cells = n(), n_donors = n_distinct(donor_id),
              total_nuclei = sum(n_nuclei),
              median_relative_abundance = median(relative_abundance), .groups = "drop"),
  "Relative abundance summary, within the four downloaded glial lineages"
)

cat(unique(comp$abundance_denominator), "\n", unique(comp$abundance_caveat), "\n", sep = "")
```

Three restrictions are recorded on the table itself, not only in prose:

- the denominator is **the four downloaded glial lineages in that donor x region**, not all cells,
  because no neuronal object has been downloaded (section 13);
- these are **naive proportions** carrying dissociation yield and sampling as well as biology, and
  are not an inferential composition model;
- composition is **never** combined with an expression-state score. The two are separate files and
  separate tables throughout.

```{r}
#| label: composition-vs-state

comp_state <- provisional %>%
  inner_join(comp %>% select(donor_id, brain_region, supertype, relative_abundance),
             by = c("donor_id", "brain_region", "supertype")) %>%
  group_by(signature_name) %>%
  summarise(n = n(),
            spearman_state_vs_abundance = cor(state_score, relative_abundance,
                                              method = "spearman", use = "complete.obs"),
            .groups = "drop")

show_table(comp_state,
           "Diagnostic only: how much a state score already tracks the supertype's own abundance")
```

That diagnostic exists so a later modelling step cannot accidentally treat a composition effect as
a state effect. It is not an argument for combining them — and the numbers make the warning concrete:
every candidate state score tracks its own supertype's relative abundance at Spearman 0.6 to 0.89.
State and composition are **not** close to orthogonal at this grain, so a later model that used both
would need to say explicitly which one it is attributing an effect to.

---

## 12. Region and disease-stage behaviour, as QC only

These plots characterise the measurement. **No signature is selected by association strength here,
and no $P/H/T \rightarrow$ morphology model is fitted.**

```{r}
#| label: qc-join

cps <- read_local_cps(omics$cps_by_region)

qnp <- read_qnp_workbook(paths$qnp_2026)
qnp_long <- qnp_primary_long(qnp) %>%
  filter(variable %in% c("abeta_percent_positive_area", "ptau_percent_positive_area"),
         status == "observed") %>%
  select(donor_id, brain_region, variable, value) %>%
  distinct()

# "Unambiguously" is checked, not assumed: one value per donor x region x marker,
# or the join does not happen.
stopifnot(!any(duplicated(qnp_long[c("donor_id", "brain_region", "variable")])))
qnp_p <- pivot_wider(qnp_long, names_from = variable, values_from = value)

qc <- provisional %>%
  left_join(cps, by = c("donor_id", "brain_region")) %>%
  left_join(qnp_p, by = c("donor_id", "brain_region"))

show_table(
  qc %>%
    summarise(
      cells = n(),
      with_CPS_Local = sum(!is.na(CPS_Local)),
      with_local_abeta = sum(!is.na(abeta_percent_positive_area)),
      with_local_ptau = sum(!is.na(ptau_percent_positive_area))
    ),
  "How much of the provisional table joins unambiguously to the stage fields"
)

# The two stage sources are independent, so their coverage being identical is
# checked rather than assumed.
qnp_grid <- distinct(qnp_p, donor_id, brain_region)
cps_grid <- distinct(cps, donor_id, brain_region)
cat(sprintf("CPS donor x region cells not covered by the QNP grey-matter grid: %d\n",
            nrow(anti_join(cps_grid, qnp_grid, by = c("donor_id", "brain_region")))))

show_table(
  qc %>% group_by(brain_region) %>%
    summarise(cells = n(), with_CPS_Local = sum(!is.na(CPS_Local)),
              with_local_abeta = sum(!is.na(abeta_percent_positive_area)),
              with_local_ptau = sum(!is.na(ptau_percent_positive_area)), .groups = "drop"),
  "Join coverage by region. LEC has no published local CPS, so it is absent by construction."
)
```

### 12.1 Raw scores

```{r}
#| label: fig-qc-cps
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Provisional state score against the published local pseudo-progression score. Descriptive only."

qc %>%
  filter(!is.na(CPS_Local)) %>%
  ggplot(aes(CPS_Local, state_score)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.4, colour = "firebrick") +
  facet_wrap(~ signature_name, scales = "free_y", ncol = 3) +
  labs(x = "CPS_Local (published, neuropathology-derived)", y = "provisional state score") +
  theme(strip.text = element_text(size = 7))
```

```{r}
#| label: fig-qc-path
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Provisional state score against local Ab (6E10) and pTau (AT8) percent positive area, grey matter. Descriptive only."

qc %>%
  select(signature_name, state_score, abeta_percent_positive_area, ptau_percent_positive_area) %>%
  pivot_longer(c(abeta_percent_positive_area, ptau_percent_positive_area),
               names_to = "marker", values_to = "percent_positive_area") %>%
  filter(!is.na(percent_positive_area)) %>%
  ggplot(aes(percent_positive_area, state_score, colour = marker)) +
  geom_point(alpha = 0.12, size = 0.4) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.4) +
  facet_wrap(~ signature_name, scales = "free", ncol = 3) +
  labs(x = "percent positive area (grey matter)", y = "provisional state score", colour = NULL) +
  theme(legend.position = "top", strip.text = element_text(size = 7))
```

### 12.2 Region-standardized scores, kept separate

```{r}
#| label: fig-qc-standardized
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "The same relationship after standardizing the score within region. Held separate from the raw version above; neither replaces the other."

qc_z <- qc %>%
  filter(!is.na(CPS_Local)) %>%
  group_by(signature_name) %>%
  group_modify(~ add_region_standardized(.x, "state_score")) %>%
  ungroup()

qc_z %>%
  ggplot(aes(CPS_Local, state_score_z_within_region)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.4, colour = "firebrick") +
  facet_wrap(~ signature_name, scales = "free_y", ncol = 3) +
  labs(x = "CPS_Local", y = "state score, z within region") +
  theme(strip.text = element_text(size = 7))
```

```{r}
#| label: qc-summary

qc_corr <- qc_z %>%
  group_by(signature_name, lineage) %>%
  summarise(
    n = n(),
    spearman_raw_vs_CPS = cor(state_score, CPS_Local, method = "spearman", use = "complete.obs"),
    spearman_regionz_vs_CPS = cor(state_score_z_within_region, CPS_Local,
                                  method = "spearman", use = "complete.obs"),
    .groups = "drop"
  )

show_table(
  qc_corr %>% arrange(lineage, signature_name),
  "Descriptive rank correlations with CPS_Local. Reported for every candidate, used to select none."
)
```

The raw and region-standardized versions are reported side by side on purpose. A relationship that
survives region standardization is within-region; one that does not is carried by the between-region
gradient in both the score and the pathology, which is a different claim.

---

## 13. Neurons are not downloaded

```{r}
#| label: neurons

manifest <- read_seaad_s3_manifest(omics$s3_manifest)

neuronal <- manifest %>%
  filter(family == "pseudobulk (donor x region x supertype)",
         !str_detect(file, "Immune|Astrocyte|Oligodendrocyte|OPC|Ependymal|Endothelial|VLMC|Microglia")) %>%
  transmute(file, size_mb = round(size_mb, 1)) %>%
  arrange(desc(size_mb))

show_table(neuronal, "Candidate neuronal pseudobulk objects. Inventory only; nothing was fetched.")

cat(sprintf("Total if fetched: %.2f GB across %d objects. Not downloaded in this task.\n",
            sum(neuronal$size_mb) / 1000, nrow(neuronal)))
```

The neuronal molecular-state definition remains unresolved for the reasons fixed in notebook 01:
HIP carries no layer-defined excitatory subclasses, so an excitatory-based neuronal $T$ is undefined
there, and an all-neuron pseudobulk confounds state with subtype composition and selective loss.
The inventory above is retained; nothing beyond it was done.

---

## Stop condition: the nine questions

```{r}
#| label: answers-inputs

state_icc  <- icc_tbl %>% filter(family == "candidate state")
null_icc   <- icc_tbl %>% filter(family == "null control") %>% select(domain, null_icc = icc)
icc_vs_null <- state_icc %>% left_join(null_icc, by = "domain") %>%
  mutate(clears_null = icc > null_icc)

worst_offset <- contrasts %>% slice_max(abs(median_offset), n = 1)
n_offset_sig <- sum(contrasts$wilcoxon_p < 0.05, na.rm = TRUE)
n_offset_chained <- sum(offsets$n_steps_from_reference > 1, na.rm = TRUE)
n_offset_missing <- sum(is.na(offsets$method_offset))

pol <- agreement %>% filter(policy_a == "equal_weight_mean")
```

```{r}
#| label: answers

answers <- tibble::tribble(
  ~question, ~answer,

  "1. Which predefined glial state signatures are defensible candidates?",
  sprintf(paste0("%d resolved external candidates were scored (%d Micro/PVM, %d Astrocyte, ",
                 "%d Oligodendrocyte, %d OPC), all ontology- or Hallmark-defined and copied ",
                 "verbatim from MSigDB %s. %d further candidates are carried unresolved with no ",
                 "genes, including the only human AD-derived state definitions (Sun 2023 MG4/MG8) ",
                 "and the SEA-AD-published option. No SEA-AD-published named gene set exists."),
          length(STATE_SIGS),
          sum(registry$role == "candidate_state" & registry$resolution_status == "resolved" & registry$domain == "Micro/PVM"),
          sum(registry$role == "candidate_state" & registry$resolution_status == "resolved" & registry$domain == "Astrocyte"),
          sum(registry$role == "candidate_state" & registry$resolution_status == "resolved" & registry$domain == "Oligodendrocyte"),
          sum(registry$role == "candidate_state" & registry$resolution_status == "resolved" & registry$domain == "OPC"),
          registry_j$provenance$msigdb_release,
          sum(registry$resolution_status == "unresolved")),

  "2. Are their genes adequately represented in the released pseudobulk?",
  sprintf(paste0("Yes for presence: fraction_present ranges %.2f to %.2f across scored signatures ",
                 "(median %.2f). Detectability is the binding constraint: %d of %d scored ",
                 "signature x domain cells lose genes to the %g non-zero-fraction rule, and the ",
                 "smallest sets lose proportionally most."),
          min(as.numeric(coverage$fraction_present), na.rm = TRUE),
          max(as.numeric(coverage$fraction_present), na.rm = TRUE),
          median(as.numeric(coverage$fraction_present), na.rm = TRUE),
          sum(coverage$n_detectably_expressed < coverage$n_present),
          nrow(coverage), score_prov$detectability$fraction),

  "3. Are row-level scores reproducible across repeated sample/library measurements?",
  sprintf(paste0("Yes as measurements, but the reproducibility is not signature-specific. Over %s ",
                 "replicate pairs, candidate-state ICC(1,1) runs %.2f to %.2f (median %.2f) and the ",
                 "median absolute replicate difference is %.2f score units against a within-domain ",
                 "SD of about 1. However only %d of %d candidates exceed their domain's ",
                 "size-matched null control, whose ICC is %.2f to %.2f, and each candidate ",
                 "correlates with that null draw at Pearson %.2f to %.2f. Row-level scoring is ",
                 "therefore reproducible; what it reproduces is largely a generic recovery and ",
                 "composition axis shared by any gene set of the same size."),
          format(nrow(pairs), big.mark = ","),
          min(state_icc$icc, na.rm = TRUE), max(state_icc$icc, na.rm = TRUE),
          median(state_icc$icc, na.rm = TRUE),
          median(pairs$abs_difference[pairs$family == "candidate state"], na.rm = TRUE),
          sum(icc_vs_null$clears_null, na.rm = TRUE), nrow(icc_vs_null),
          min(null_icc$null_icc, na.rm = TRUE), max(null_icc$null_icc, na.rm = TRUE),
          min(specificity$pearson_with_null[specificity$role == "candidate_state"], na.rm = TRUE),
          max(specificity$pearson_with_null[specificity$role == "candidate_state"], na.rm = TRUE)),

  "4. Does assay method materially affect the scores?",
  sprintf(paste0("Yes, and systematically. Offsets are estimated within donor x region x ",
                 "supertype, so they cannot be cohort effects. %d of %d directly observed ",
                 "signature x method-pair contrasts reach p < 0.05 (Wilcoxon), and every one of ",
                 "the 26 points the same way: 10xMulti reads lower than both other methods. The ",
                 "largest median offset is %.2f score units (%s, %s minus %s). ",
                 "Mixed-method pairs disagree more than same-method pairs for %d of %d ",
                 "signatures. The design is a chain, so %d of %d method offsets are reached in ",
                 "more than one step and %d are not estimable at all."),
          n_offset_sig, nrow(contrasts),
          worst_offset$median_offset[1], worst_offset$signature_name[1],
          worst_offset$second_method[1], worst_offset$first_method[1],
          sum(drivers$mixed_minus_same > 0, na.rm = TRUE), nrow(drivers),
          n_offset_chained, nrow(offsets), n_offset_missing),

  "5. Is there evidence for a minimum support criterion?",
  sprintf(paste0("Yes, and it is empirical rather than conventional: replicate disagreement falls ",
                 "monotonically with nuclei support, and absolute difference correlates with ",
                 "support at Spearman %.2f to %.2f. A candidate rule is tabulated per domain. ",
                 "Nothing has been filtered: %.0f%% of scored rows sit below 10 nuclei."),
          min(drivers$spearman_absdiff_vs_min_nuclei, na.rm = TRUE),
          max(drivers$spearman_absdiff_vs_min_nuclei, na.rm = TRUE),
          100 * mean(as.numeric(normtab$n_nuclei) < 10)),

  "6. What aggregation rule should convert repeated measurements into donor x region x supertype estimates?",
  sprintf(paste0("The nucleus-weighted mean of assay-adjusted row-level scores. The assay offset ",
                 "is systematic and one-directional, so it must be removed rather than averaged ",
                 "over, and replicate precision rises steeply with nuclei, so the rows are not ",
                 "interchangeable. Against the unweighted mean it agrees at Pearson %.3f (median ",
                 "across signatures), the nucleus-weighted mean alone at %.3f and the ",
                 "assay-adjusted unweighted mean at %.3f, so the choice moves few values but moves ",
                 "them exactly where support or platform is unbalanced. The raw-count pooled score ",
                 "agrees least (%.3f) and is a sensitivity arm only."),
          median(pol$pearson[pol$policy_b == "assay_adjusted_nucleus_weighted_mean"], na.rm = TRUE),
          median(pol$pearson[pol$policy_b == "nucleus_weighted_mean"], na.rm = TRUE),
          median(pol$pearson[pol$policy_b == "assay_adjusted_mean"], na.rm = TRUE),
          median(pol$pearson[pol$policy_b == "pooled_score"], na.rm = TRUE)),

  "7. How strongly do state scores vary by supertype and region?",
  sprintf(paste0("Supertype explains %.0f%% to %.0f%% of row-level score variance (median %.0f%%); ",
                 "region explains %.0f%% to %.0f%% (median %.0f%%). Supertype dominates region, ",
                 "which is why it must stay a stratum rather than be averaged over."),
          100 * min(variance_share$supertype_variance_share[variance_share$family == "candidate state"]),
          100 * max(variance_share$supertype_variance_share[variance_share$family == "candidate state"]),
          100 * median(variance_share$supertype_variance_share[variance_share$family == "candidate state"]),
          100 * min(variance_share$region_variance_share[variance_share$family == "candidate state"]),
          100 * max(variance_share$region_variance_share[variance_share$family == "candidate state"]),
          100 * median(variance_share$region_variance_share[variance_share$family == "candidate state"])),

  "8. What provisional state variables are ready for later donor x ROI modelling?",
  sprintf(paste0("A long table of %s donor x region x supertype x signature cells over %d donors ",
                 "and %d regions, of which %s are well supported (>= 20 nuclei). Restricted to the ",
                 "candidate ROI set it holds %s cells. It is exploratory, not a model matrix."),
          format(nrow(provisional), big.mark = ","),
          n_distinct(provisional$donor_id), n_distinct(provisional$brain_region),
          format(sum(provisional$support_status == "well supported"), big.mark = ","),
          format(sum(provisional$brain_region %in% ROI), big.mark = ",")),

  "9. Which uncertainties remain before defining final T variables?",
  paste0("No scored signature is simultaneously human, AD-derived and state-level, and none is ",
         "distinguishable from a size-matched random gene set on reliability alone; ",
         "sample_name semantics remain undocumented, so the replicate structure is interpreted as ",
         "measurement rather than known biology; the supertype stratum has no agreed collapsing ",
         "rule; microglia are not separated from PVM inside Micro-PVM_*; the support rule is ",
         "proposed but unapplied; and the neuronal axis is undefined.")
)

show_table(answers, "The nine stop-condition questions, answered from the tables above")
```

---

## Report

### Files changed

```{r}
#| label: files-changed

show_table(
  tibble::tribble(
    ~path, ~status, ~what,
    "results/analysis/02_seaad_omics_state_exploration.qmd", "new", "this notebook",
    "R/seaad_state.R", "new",
      "reliability, aggregation and composition helpers; no data is written from R",
    "src/seaad/omics/signature_registry.py", "new",
      "signature curation; resolves genes from MSigDB or refuses and marks unresolved",
    "src/seaad/omics/state_scoring.py", "new",
      "TMM / logCPM at the released row grain, module scoring, pooled sensitivity arm",
    "src/seaad/cli/omics_build_signature_registry.py", "new", "CLI for the registry",
    "src/seaad/cli/omics_state_scores.py", "new", "CLI for normalization and scoring",
    "src/seaad/tests/test_state_scoring.py", "new",
      "TMM properties, taxonomy domains, unresolved-signature refusal, end-to-end row grain",
    "src/seaad/pyproject.toml", "modified", "two new console scripts",
    "data/derivatives/sea-ad/omics_state/", "new",
      "registry, gene lists, row normalization, row scores, pooled scores, coverage, contributions, provenance"
  ),
  "Files changed by this task. Acquisition and preparation were not touched."
)
```

### Summary

**Signatures examined and provenance.** `r length(STATE_SIGS)` resolved external candidates plus
`r length(ID_SIGS)` lineage-identity controls and `r length(NULL_SIGS)` size-matched null controls
were scored; `r sum(registry$resolution_status == "unresolved")` further candidates are carried
with no genes. Every resolved definition is copied verbatim from MSigDB
`r registry_j$provenance$msigdb_release`, with the source GMT URL, size and SHA-256 recorded.
The SEA-AD-published route resolves to a per-gene effect-size table rather than a named gene set,
and is recorded as a labelled secondary alternative that was deliberately not retrieved.

**Normalization and scoring.** Summed raw UMI counts, row-sum library sizes, TMM factors
(Robinson & Oshlack 2010; log-ratio trim `r score_prov$normalization$logratio_trim`, abundance trim
`r score_prov$normalization$sum_trim`, reference by upper quartile, geometric-mean rescaling),
then edgeR-style log2 CPM with prior `r score_prov$normalization$prior_count`. TMM is implemented
in this repository because edgeR is absent from the project's R environment. Scores are the
unweighted mean of within-domain gene z-scores; the unstandardized mean logCPM is kept beside it.
No single-cell normalization; no repeated row summed before scoring; no parameter tuned against any
outcome.

**Gene coverage.** Presence is high and detectability is the binding constraint; every absent
symbol is listed by name, and nothing is fuzzy-matched.

**Replicate reliability, assay effects, support.** Quantified in sections 6 and 7 from
`r format(nrow(pairs), big.mark = ",")` within-group replicate pairs. Reproducibility is good in
absolute terms and **not signature-specific**: size-matched random gene sets reproduce at least as
well as every candidate, so ICC here is evidence about the measurement rather than about the
biology. Assay-method offsets are estimated only inside donor x region x supertype, where donor,
region and supertype are fixed by construction; they are systematic and one-directional. Replicate
disagreement falls steeply and monotonically with nuclei support.

**Recommended aggregation rule.** The nucleus-weighted mean of **assay-adjusted** row-level scores.
The assay offset is systematic, one-directional and significant in all thirteen contrasts, so it is
removed rather than averaged over; replicate precision rises steeply with nuclei, so rows are not
treated as interchangeable. The unweighted assay-adjusted mean is retained beside it as the
defensible simpler alternative once a support rule is applied, and the raw-count pooled arm is a
sensitivity analysis only. Row-level scores are never overwritten.

**Provisional state variables.** The long table in section 10, with its support bands, plus the
separate compositional table in section 11.

### Unresolved

```{r}
#| label: unresolved-summary

show_table(
  tibble::tribble(
    ~issue, ~why_it_matters, ~what_would_resolve_it,
    "No candidate is distinguishable from a size-matched random gene set on reliability",
      "high ICC is a property of the measurement, not evidence that a signature tracks its biology",
      "a signature with genuine state structure, or a specificity criterion adopted before selection",
    "No human, AD-derived, state-level signature is scored",
      "every scored axis is an ontology or Hallmark biological axis, not a disease state",
      "recover Sun et al. 2023 Table S2 (MG4, MG8) and the Habib 2020 DAA list verbatim",
    "sample_name semantics are undocumented",
      "repeated rows are interpreted as measurement; if they were tissue blocks they would be biology",
      "a statement from SEA-AD, or a released specimen/tissue-block field",
    "Micro-PVM is not resolved into microglia and PVM",
      "the resident-myeloid axis is Micro/PVM, not microglia",
      "a finer released taxonomy, or a marker-based split that would itself need validating",
    "The support rule is proposed but not applied",
      "a large share of rows sit below ten nuclei",
      "one explicit decision, applied once, with the cost table recomputed",
    "Supertype stratification has no collapsing rule",
      "the eventual T is per donor x region, and supertype has to be handled on the way there",
      "a stated rule: condition, weight by abundance, or select one supertype per lineage",
    "The neuronal axis is undefined",
      "HIP has no layer-defined excitatory subclasses; all-neuron pooling confounds state with loss",
      "a defensible neuronal grouping present in every candidate ROI"
  ),
  "What remains open"
)
```

### The single next scientific decision

> **Resolve, or abandon, a human AD-derived glial state signature before anything else.**

The measurement pipeline is in place: normalization, scoring, replicate reliability, the assay
offset and the support curve are all quantified, and an aggregation rule follows from them without
reference to any outcome. What is *not* in place is a signature worth measuring. Every scored axis
is an ontology or Hallmark biological axis; none is a disease state in the sense the eventual $T$
block requires; and section 6.3 shows none of them is distinguishable from a size-matched random
gene set on reliability alone.

The choice is therefore between recovering a human, AD-cohort-derived state definition verbatim —
Sun et al. 2023 Table S2 is the concrete target, with Habib 2020 DAA behind it — and accepting that
$T$ will be built on generic biological axes with that limitation stated in the model. Everything
downstream (the support rule, the supertype collapsing rule, the ROI freeze) is cheaper to settle
after that choice than before it, because each of them would otherwise have to be redone against a
different signature set.
