---
title: "SEA-AD molecular state: choosing the T variables"
subtitle: "Which single Micro/PVM and Astrocyte measurements go forward, whether background correction changes them, and how much of them P and H already explain"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    df-print: default
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose and scope

`02_seaad_omics_state_exploration.qmd` built the measurement: TMM, logCPM, row-level module
scores, the assay-method offset, the replicate-reliability curve and an aggregation rule that
follows from them. It ended on one sentence — *resolve, or abandon, a human AD-derived glial
state signature before anything else* — because every axis it had scored was an ontology or
Hallmark biological axis, and none of them was distinguishable from a size-matched random gene
set on reliability alone.

This notebook is a **decision notebook, not another inventory**. It does five things and stops:

1. resolves that outstanding signature question and **reduces** the candidate set to one primary
   and one secondary axis per lineage;
2. adds exactly one scoring sensitivity — a **matched-background (competitive) score** — and asks
   whether it materially changes the measurement;
3. applies the already-supported assay-adjusted, nucleus-weighted aggregation and collapses the
   supertype stratum under one stated rule;
4. writes **one** small analysis-ready donor x region molecular-state table;
5. joins it to the existing $P/H$ block and asks the only question that matters before morphology:
   **does each score retain appreciable donor x region variation after global stage and the
   existing $P/H$ measurements are accounted for?**

Deliberately **not** done: any morphology, any cognition, any neuronal download, any further
normalization or scoring framework, any hypothesis test, any ROI freeze. No signature is selected
by its association with CPS, $P$, $H$, morphology or cognition — the selection in section 2 is
declared and justified *before* any association appears in this notebook, and nothing later
revises it.

```{r}
#| label: setup

my_packages <- c("tidyverse", "knitr", "jsonlite", "readxl", "splines")
invisible(lapply(my_packages, library, character.only = TRUE))

find_root <- function(p = getwd()) {
  while (!dir.exists(file.path(p, ".git"))) {
    if (identical(dirname(p), p)) stop("project root not found")
    p <- dirname(p)
  }
  p
}
PROJECT_ROOT <- find_root()
source(file.path(PROJECT_ROOT, "R", "seaad_explore.R"))
source(file.path(PROJECT_ROOT, "R", "seaad_state.R"))
source(file.path(PROJECT_ROOT, "R", "seaad_analysis_ready.R"))

paths    <- seaad_paths(PROJECT_ROOT)
omics    <- seaad_omics_paths(PROJECT_ROOT)
state    <- seaad_state_paths(PROJECT_ROOT)
analysis <- seaad_analysis_paths(PROJECT_ROOT)

show_table <- function(x, caption = NULL, digits = 3) {
  kable(x, caption = caption, digits = digits, format.args = list(big.mark = ""))
}
theme_set(theme_minimal(base_size = 10))

# The candidate multimodal ROI set. Still a candidate: section 8 reports every
# released region, and nothing here freezes the set.
ROI <- c("MTG", "DFC", "MEC", "STG", "V1C", "HIP")
```

```{r}
#| label: selection

# ---------------------------------------------------------------------------
# THE SELECTION. Declared here, at the top of the notebook, before any number
# in it has been computed, so that no table below can be read as having chosen
# it. The justification is in section 2; the criteria are biological construct,
# human/AD relevance, published provenance, gene availability and
# interpretability, and nothing else.
# ---------------------------------------------------------------------------
SELECTED <- tibble::tribble(
  ~axis,        ~role,        ~signature_name,
  "Micro/PVM",  "primary",    "sun_2023_MG4_lipid_processing",
  "Micro/PVM",  "secondary",  "sun_2023_MG8_inflammatory_II",
  "Astrocyte",  "primary",    "astrocyte_activation_GOBP",
  "Oligo/OPC",  "secondary",  "myelin_maintenance_GOBP",
  "Oligo/OPC",  "secondary",  "oligodendrocyte_differentiation_GOBP"
)

PRIMARY_MICRO <- "sun_2023_MG4_lipid_processing"
PRIMARY_ASTRO <- "astrocyte_activation_GOBP"
# Kept out of the carried-forward set but scored throughout as the
# dataset-independent comparator for the Micro/PVM axis.
COMPARATOR_MICRO <- "microglial_glial_activation_GOBP"
```

---

## 0. What this notebook reads, and what it writes

```{r}
#| label: inputs

inputs <- tibble::tribble(
  ~role, ~path,
  "signature registry (rebuilt)",        state$registry_csv,
  "signature genes",                     state$signature_genes,
  "Sun 2023 Table S1 supplement",        file.path(analysis$supplements_dir,
                                                   "sun_2023_cell_table_s1.xlsx"),
  "row-level state scores",              state$row_scores,
  "row-level competitive scores",        analysis$competitive_scores,
  "matched-background summary",          analysis$background_summary,
  "signature gene coverage",             state$gene_coverage,
  "pooled sensitivity scores",           state$pooled_scores,
  "published donor x region CPS",        omics$cps_by_region,
  "QNP workbook (P and H blocks)",       paths$qnp_2026
) %>%
  mutate(exists = file.exists(path),
         size_mb = round(ifelse(file.exists(path), file.size(path) / 1e6, NA_real_), 2),
         file = basename(path)) %>%
  select(role, file, exists, size_mb)

show_table(inputs, "Everything this notebook reads. Nothing new was downloaded from SEA-AD.")
```

It writes exactly one derivative, `molecular_state_analysis_table.csv`, in section 6. Composition
stays where notebook 02 put it: a separate table, never merged with a state score.

```{r}
#| label: load

registry   <- read_csv(state$registry_csv, show_col_types = FALSE)
registry_j <- fromJSON(state$registry_json, simplifyVector = FALSE)
coverage   <- read_csv(state$gene_coverage, show_col_types = FALSE)
scores     <- read_state_scores(state$row_scores)
comp       <- read_competitive_scores(analysis$competitive_scores)
bg_summary <- read_csv(analysis$background_summary, show_col_types = FALSE)
comp_prov  <- fromJSON(analysis$competitive_prov, simplifyVector = FALSE)
prep_prov  <- fromJSON(state$preparation_prov, simplifyVector = FALSE)
score_prov <- fromJSON(state$scoring_prov, simplifyVector = FALSE)

role_of <- registry %>% select(signature_name, role, feature_origin, resolution_status)
scores  <- scores %>% left_join(role_of, by = "signature_name")
# The competitive arm corrects the size-matched null controls as well as the
# candidates, so that section 3 can compare like with like.
comp    <- comp %>% left_join(role_of %>% select(signature_name, role), by = "signature_name")
COMP_SIGS <- comp %>% filter(role == "candidate_state") %>% pull(signature_name) %>% unique()

STATE_SIGS <- registry %>%
  filter(role == "candidate_state", resolution_status == "resolved") %>% pull(signature_name)
ID_SIGS <- registry %>% filter(role == "identity_control") %>% pull(signature_name)
NULL_SIGS <- registry %>% filter(role == "null_control") %>% pull(signature_name)
```

---

## 1. The measurement pipeline is frozen

Nothing in this notebook re-opens normalization, scoring or the reliability analysis. The chain
below is taken as settled by notebook 02 and reused unchanged.

```{r}
#| label: frozen

show_table(
  tibble::tribble(
    ~step, ~rule, ~settled_in,
    "input",        "summed raw UMI counts on the released row grain",        "notebook 02, s.4",
    "library size", "row sum of raw counts",                                  "notebook 02, s.4",
    "scaling",      sprintf("TMM, log-ratio trim %g, abundance trim %g, upper-quartile reference",
                            score_prov$normalization$logratio_trim,
                            score_prov$normalization$sum_trim),              "notebook 02, s.4",
    "expression",   sprintf("edgeR-style log2 CPM, prior count %g",
                            score_prov$normalization$prior_count),           "notebook 02, s.4",
    "score",        "unweighted mean of within-domain gene z-scores",         "notebook 02, s.5",
    "assay",        "additive per-method offset from mixed-method replicate pairs only",
                                                                             "notebook 02, s.6.5",
    "aggregation",  "nucleus-weighted mean of assay-adjusted row scores",     "notebook 02, s.9",
    "sensitivity",  "raw-count pooled arm, kept beside and never substituted", "notebook 02, s.9"
  ),
  "The frozen chain. One thing is added to it in section 3, and nothing is replaced."
)
```

One implementation note found while re-running the pipeline, recorded because it is a property of
the frozen code rather than of this task:

```{r}
#| label: tie-note

show_table(
  tibble::tribble(
    ~observation, ~scope, ~consequence,
    paste("The TMM reference row is chosen as the row whose upper quartile is closest to the",
          "domain mean upper quartile. In the Lymphocyte and Monocyte domains almost every row",
          "is so sparse that its upper quartile falls back to 1/n_genes, so 688 of 786",
          "Lymphocyte rows tie to ten significant figures and the choice is settled by",
          "floating-point noise, which differs between numpy builds."),
    "Lymphocyte and Monocyte only",
    paste("None. Neither domain carries a scored signature. The Micro/PVM, Astrocyte,",
          "Oligodendrocyte and OPC reference rows are identical across numpy 2.2 and 2.4, and",
          "re-running the scorer on the previous registry reproduced row_state_scores.csv",
          "byte for byte.")
  ),
  "A latent tie in the frozen normalization, and why it does not touch any state score"
)
```

---

## 2. Reducing the signature set

### 2.1 The criteria, and the criteria that were refused

```{r}
#| label: criteria

show_table(
  tibble::tribble(
    ~criterion, ~used, ~why,
    "biological construct", "yes",
      "the axis has to name something the eventual model can be about",
    "human and AD relevance", "yes",
      "a mouse or non-brain definition measures an analogy, not the thing",
    "published provenance", "yes",
      "membership must be copyable verbatim from a versioned source",
    "gene availability", "yes",
      "a set whose genes are largely absent or undetectable is not measurable here",
    "interpretability", "yes",
      "a direction of effect has to be statable in advance",
    "association with SEA-AD CPS", "NO", "would select the axis by the answer",
    "association with QNP P or H", "NO", "would select the axis by the answer",
    "association with morphology or cognition", "NO", "out of scope and circular",
    "which signature gives the strongest result", "NO", "not a criterion at all"
  ),
  "What the selection in the setup chunk was, and was not, allowed to use"
)
```

### 2.2 Micro/PVM: the human AD-derived state definition is now resolved

Notebook 02's single blocking gap was that no scored Micro/PVM axis was simultaneously human,
AD-derived and state-level. That gap is closed. Sun et al. 2023 (Cell 186:4386-4403) release their
microglial state markers as a **machine-readable supplementary workbook**, not as a figure image:
Table S1, page 2, one row per (state, gene) with the authors' own `FindAllMarkers` output.

```{r}
#| label: sun-provenance

sun <- registry_j$provenance$supplements[[1]]
show_table(
  tibble::tibble(
    field = c("citation", "sheet", "state column", "gene column", "groups released",
              "MG4 genes", "MG8 genes", "SHA-256", "filter applied here"),
    value = c(sun$citation, sun$sheet, sun$group_column, sun$gene_column,
              as.character(sun$n_groups),
              as.character(sun$group_sizes$MG4), as.character(sun$group_sizes$MG8),
              substr(sun$sha256, 1, 24), sun$released_filter)
  ),
  "Provenance of the two newly resolved Micro/PVM signatures"
)
```

The file is pinned by SHA-256 and re-verified on every registry build, and **no threshold of this
project's own is applied**: every released row of the named state is taken, in published order.
Choosing a p-value or fold-change cut-off would have made the membership our choice rather than
the publication's.

Why **MG4** is primary and **MG8** secondary, on the criteria above and nothing else:

- *construct.* MG4 is the human lipid-processing state. The authors' own GO enrichment of exactly
  this marker set (Table S1 page 3) is led by regulation of cholesterol storage, cholesterol
  efflux, lipid storage and lipoprotein biosynthesis, over `ABCA1`, `APOE`, `TREM2`, `PPARG`,
  `MSR1` and `GPNMB`. That is the human analogue of the disease-associated microglial state, which
  is the resident-myeloid construct this project's $T$ block is about.
- *human and AD relevance.* Derived from ROSMAP post-mortem human cortex across the AD spectrum,
  by unsupervised clustering of microglial nuclei. MG8 shares this exactly.
- *distinctness.* MG8 is led instead by cytokine-mediated signalling and response to
  lipopolysaccharide (`CD86`, `CIITA`, `IL15`, `IL10RA`, `TNFRSF1B`, `IRAK3`, with `LRRK2`,
  `SPON1`, `FOXP1` among its top markers). It shares MG4's provenance but not its biology, which
  is what makes it a useful secondary rather than a near-duplicate.

`microglial_glial_activation_GOBP` (GO:0061900) is **not** carried forward but is kept scored
throughout as the dataset-independent comparator: it is the fallback if the Sun axes turn out to
be unmeasurable here, and it is the reference against which "does an AD-derived definition behave
differently from a generic activation axis" can be asked later.

### 2.3 Astrocyte: a primary axis, and no defensible secondary

`astrocyte_activation_GOBP` (GO:0048143, reactive astrocyte state) is the primary. It is ontology
defined, human-applicable, dataset-independent, and consistent with Escartin et al. 2021, who
recommend describing astrocyte reactivity by measured markers rather than by an A1/A2 polarity.

**No secondary astrocyte axis is retained**, because none is clearly justified:

```{r}
#| label: astro-alternatives

show_table(
  tibble::tribble(
    ~alternative, ~status, ~why_not_carried,
    "astrocyte_activation_regulation_GOBP (GO:0061889)", "scored, not selected",
      "eight genes. A regulatory set that small is not a state axis and is expected to be unstable; kept only as a printed comparator.",
    "Liddelow 2017 PAN / A1 / A2", "unresolved",
      "cassette membership exists only inside a figure image. Image extraction was explicitly out of scope, and Escartin 2021 advises against the polarity anyway.",
    "Habib 2020 disease-associated astrocyte", "unresolved",
      "mouse-derived. A mouse definition fails the human-relevance criterion that MG4 passes.",
    "a Sun-equivalent human AD astrocyte state", "not available",
      "no equivalent machine-readable human astrocyte state cassette was identified from the provenance already collected."
  ),
  "Why the astrocyte axis carries one signature and not two"
)
```

This asymmetry is a real limitation and is carried into the decision table rather than papered
over: the Micro/PVM axis has a human AD-derived definition and the astrocyte axis does not.

### 2.4 Are the selected genes actually there?

```{r}
#| label: availability

sel_cov <- coverage %>%
  inner_join(SELECTED %>% select(signature_name, axis, role), by = "signature_name") %>%
  bind_rows(coverage %>% filter(signature_name == COMPARATOR_MICRO) %>%
              mutate(axis = "Micro/PVM", role = "comparator")) %>%
  filter(domain %in% c("Micro/PVM", "Astrocyte", "Oligodendrocyte", "OPC")) %>%
  transmute(axis, role, signature_name, domain,
            n_defined, n_present, n_detectably_expressed, n_used,
            fraction_present = as.numeric(fraction_present),
            fraction_usable = n_used / n_defined) %>%
  arrange(axis, role)

show_table(sel_cov,
  "Gene availability of the selected axes inside their own scoring domain")
```

```{r}
#| label: availability-note

cat(sprintf(
  "MG4 keeps %d of %d defined symbols as usable genes in Micro/PVM (%.0f%%); MG8 keeps %d of %d (%.0f%%).\n",
  sel_cov$n_used[sel_cov$signature_name == "sun_2023_MG4_lipid_processing"],
  sel_cov$n_defined[sel_cov$signature_name == "sun_2023_MG4_lipid_processing"],
  100 * sel_cov$fraction_usable[sel_cov$signature_name == "sun_2023_MG4_lipid_processing"],
  sel_cov$n_used[sel_cov$signature_name == "sun_2023_MG8_inflammatory_II"],
  sel_cov$n_defined[sel_cov$signature_name == "sun_2023_MG8_inflammatory_II"],
  100 * sel_cov$fraction_usable[sel_cov$signature_name == "sun_2023_MG8_inflammatory_II"]))
```

One reading caveat travels with both Sun axes and is recorded on the registry entry itself: scored
on **Micro/PVM pseudobulk**, an MG4 marker score measures a *shift of the whole resident-myeloid
pool towards that state*, not the abundance of MG4 cells. At this grain the two are not separable,
and the released taxonomy does not separate microglia from perivascular macrophages either.

---

## 3. One scoring sensitivity: a matched-background score

### 3.1 What the correction is, and what it is matched on

Notebook 02's sharpest negative result was that size-matched *random* gene sets reproduce across
replicate libraries as well as any candidate signature, and that every candidate correlates
strongly with its domain's null draw. That is the signature of a shared axis — recovery depth and
overall expression level — riding on every gene at once. The competitive score subtracts it:

$$\text{competitive score} \;=\; \underbrace{\overline{z}_{\text{signature}}}_{\text{as before}} \;-\; \underbrace{\overline{z}_{\text{matched background}}}_{\text{new}}$$

```{r}
#| label: background-rule

bg <- comp_prov$background
show_table(
  tibble::tibble(
    field = c("matched on", "binning", "eligible pool", "draws", "seed", "draw rule",
              "what it is not"),
    value = c(paste(unlist(bg$matched_on), collapse = "; "), bg$binning, bg$pool,
              as.character(bg$n_draws), as.character(bg$seed), bg$draw_rule,
              bg$not_an_empirical_null)
  ),
  "How the matched background is constructed. The seed is recorded in the provenance file."
)
```

The background is **not** a random gene set: each background gene replaces one signature gene from
the same joint stratum of mean abundance and detection frequency, so the two sets carry the same
technical profile and differ only in gene identity. A few dozen draws are enough to pin down the
mean that is subtracted, and no more than that is wanted — this is a sensitivity, not an
empirical-null framework.

```{r}
#| label: background-quality

show_table(
  bg_summary %>%
    transmute(domain, signature_name, n_draws, n_signature_genes_matched,
              n_eligible_background_genes, n_strata_used,
              sig_log2cpm = signature_mean_log2_cpm,
              bg_log2cpm = background_mean_log2_cpm,
              sig_detect = signature_mean_detection_fraction,
              bg_detect = background_mean_detection_fraction,
              median_background_sd_z, se_of_background_mean = standard_error_of_background_mean,
              widened_draws) %>%
    arrange(domain, signature_name),
  "Did the matching work, and is the subtracted mean stable? Signature and background abundance and detection should agree closely, and the standard error of the background mean should be small against a within-domain score SD of about 1."
)
```

### 3.2 Does the correction materially change the measurement?

```{r}
#| label: raw-vs-comp-row

raw_vs_comp_row <- comp %>%
  group_by(domain, signature_name, role) %>%
  summarise(n_rows = n(),
            pearson = cor(score_z_mean, competitive_score),
            spearman = cor(score_z_mean, competitive_score, method = "spearman"),
            sd_raw = sd(score_z_mean),
            sd_competitive = sd(competitive_score),
            median_background = median(background_mean_z),
            .groups = "drop") %>%
  arrange(domain, signature_name)

show_table(raw_vs_comp_row,
  "Row level: raw score against competitive score, per signature")
```

```{r}
#| label: fig-raw-vs-comp
#| fig-width: 9
#| fig-height: 3.6
#| fig-cap: "Row-level raw score against matched-background-corrected score, for the selected axes and the Micro/PVM comparator. The dashed line is the identity."

comp %>%
  filter(signature_name %in% c(SELECTED$signature_name, COMPARATOR_MICRO)) %>%
  ggplot(aes(score_z_mean, competitive_score)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, linewidth = 0.3, colour = "grey50") +
  geom_point(alpha = 0.10, size = 0.4) +
  facet_wrap(~ signature_name, nrow = 1, scales = "free") +
  labs(x = "raw mean z", y = "competitive score") +
  theme_minimal(base_size = 20)
```

---

## 4. Repeated-measure aggregation, applied

The rule is the one notebook 02 supported and is applied here without re-deriving it: subtract the
signature-specific additive assay offset estimated **only** from mixed-method replicate pairs
inside a donor x region x supertype, then take the nucleus-weighted mean. The competitive score is
a different variable, so its offsets are re-estimated on it rather than borrowed from the raw
score.

```{r}
#| label: aggregate

pairs_raw  <- replicate_pairs(scores, value_col = "score_z_mean")
offsets_raw <- method_offsets(pairs_raw)

pairs_comp  <- replicate_pairs(comp, value_col = "competitive_score")
offsets_comp <- method_offsets(pairs_comp)

agg_raw <- aggregate_policies(scores, offsets_raw, value_col = "score_z_mean") %>%
  select(donor_id, brain_region, supertype, signature_name, lineage, domain,
         n_measurements, n_nuclei_total, n_assay_methods,
         state_score = assay_adjusted_nucleus_weighted_mean,
         equal_weight_mean, nucleus_weighted_mean, within_group_sd)

agg_comp <- aggregate_policies(comp, offsets_comp, value_col = "competitive_score") %>%
  select(donor_id, brain_region, supertype, signature_name,
         competitive_score = assay_adjusted_nucleus_weighted_mean)

supertype_level <- agg_raw %>%
  left_join(agg_comp, by = c("donor_id", "brain_region", "supertype", "signature_name"))

show_table(
  supertype_level %>%
    filter(signature_name %in% c(SELECTED$signature_name, COMPARATOR_MICRO)) %>%
    group_by(domain, signature_name) %>%
    summarise(cells = n(), donors = n_distinct(donor_id),
              donor_regions = n_distinct(paste(donor_id, brain_region)),
              with_competitive = sum(!is.na(competitive_score)), .groups = "drop"),
  "Donor x region x supertype estimates. Row-level scores are preserved untouched in the derivative files; nothing below overwrites them."
)
```

```{r}
#| label: icc

# Technical reliability of both scorings, reusing notebook 02's estimator: a
# one-way ICC over repeated sample/library measurements inside a fixed
# donor x region x supertype. This is a property of the measurement, not
# evidence that a signature tracks its biology - notebook 02 established that a
# size-matched random set reaches the same ICC as any candidate.
# `icc_oneway()` returns a one-row tibble, so the scalar is pulled out here.
icc_of <- function(rows, value_col) {
  rows %>%
    group_by(signature_name, domain) %>%
    summarise(icc = icc_oneway(.data[[value_col]],
                               paste(donor_id, brain_region, supertype))$icc,
              .groups = "drop")
}

icc_raw <- icc_of(scores %>% filter(signature_name %in% comp$signature_name), "score_z_mean") %>%
  rename(icc_raw = icc)
icc_comp <- icc_of(comp, "competitive_score") %>% rename(icc_competitive = icc)

null_floor <- icc_raw %>%
  inner_join(icc_comp, by = c("signature_name", "domain")) %>%
  inner_join(role_of %>% select(signature_name, role), by = "signature_name") %>%
  filter(role == "null_control") %>%
  select(domain, null_icc_raw = icc_raw, null_icc_competitive = icc_competitive)

reliability <- icc_raw %>%
  inner_join(icc_comp, by = c("signature_name", "domain")) %>%
  inner_join(role_of %>% select(signature_name, role), by = "signature_name") %>%
  filter(role == "candidate_state") %>%
  left_join(null_floor, by = "domain") %>%
  mutate(clears_null_raw = icc_raw > null_icc_raw,
         clears_null_competitive = icc_competitive > null_icc_competitive,
         # How far above the floor, not just whether. A bare ">" would put an
         # axis that barely clears in the same box as one that clears fivefold.
         margin_over_null = icc_competitive - null_icc_competitive) %>%
  select(signature_name, domain, icc_raw, null_icc_raw, clears_null_raw,
         icc_competitive, null_icc_competitive, clears_null_competitive,
         margin_over_null) %>%
  arrange(domain, desc(icc_competitive))

show_table(reliability,
  "ICC(1,1) across repeated sample/library measurements, for both scorings, each against its own domain's size-matched null control put through the same correction")
```

Notebook 02's negative result is reproduced exactly in the `icc_raw` column: no candidate clears
its domain's null control on the **raw** score, so raw reproducibility is a property of the
measurement rather than evidence about the signature. The `icc_competitive` column is the new
information, and it separates the candidates sharply - which is the point of the correction. It is
a **technical-reliability** criterion computed from repeated measurements of the same
donor x region x supertype; no outcome, and no association with CPS, $P$, $H$ or morphology, enters
it.

Read the two null columns together. The correction does what it is supposed to do to a random gene
set: the size-matched null falls from an ICC of 0.68-0.77 raw to a small residual once its own
matched background is subtracted. Against that floor the candidates separate into two groups, and
the separation is not subtle - the Sun axes sit several times above it while the ontology
activation axes sit just above it.

```{r}
#| label: carry-rule

# The rule, stated before it is applied. An axis is only carried forward if its
# background-corrected score is more reproducible across repeated libraries than
# a size-matched random gene set that has been through the same correction. An
# axis that fails it is `unresolved`: the definition is chosen and defensible,
# but no reliable measurement of it exists in this release.
carry_status <- function(sig) {
  row <- reliability %>% filter(signature_name == sig)
  if (!nrow(row) || is.na(row$clears_null_competitive[1])) return("unresolved")
  if (row$clears_null_competitive[1]) "carry_forward" else "unresolved"
}

show_table(
  SELECTED %>%
    filter(axis != "Oligo/OPC") %>%
    rowwise() %>%
    mutate(status_under_the_rule = carry_status(signature_name)) %>%
    ungroup(),
  "The rule applied to the selected axes. The signatures were fixed before any of these numbers existed; what the rule decides is whether each one is *measurable* here, not whether it is the right construct."
)
```

```{r}
#| label: raw-vs-comp-aggregated

raw_vs_comp_agg <- supertype_level %>%
  filter(!is.na(competitive_score)) %>%
  group_by(domain, signature_name) %>%
  summarise(n = n(),
            pearson = cor(state_score, competitive_score),
            spearman = cor(state_score, competitive_score, method = "spearman"),
            .groups = "drop")

show_table(raw_vs_comp_agg,
  "The same agreement question after aggregation, on the assay-adjusted nucleus-weighted values")
```

```{r}
#| label: correction-verdict

verdict <- raw_vs_comp_row %>%
  filter(role == "candidate_state") %>%
  select(signature_name, row_pearson = pearson, sd_raw, sd_competitive) %>%
  left_join(raw_vs_comp_agg %>% select(signature_name, aggregated_pearson = pearson),
            by = "signature_name") %>%
  mutate(variance_removed = 1 - (sd_competitive / sd_raw)^2) %>%
  arrange(desc(variance_removed))

show_table(verdict,
  "How much the correction changes each score. `variance_removed` is the share of the raw score's variance that the matched background accounted for.")

cat(sprintf(paste0(
  "The correction is material, not cosmetic. Across the %d corrected candidate signatures the matched ",
  "background accounts for %.0f%% to %.0f%% of the raw score's variance, and raw and competitive ",
  "scores agree at Pearson %.2f to %.2f at row level (%.2f to %.2f after aggregation).
",
  "For the two carried-forward axes: %s, raw-vs-competitive Pearson %.2f at row level and %.2f ",
  "aggregated, %.0f%% of variance removed; %s, %.2f and %.2f, %.0f%% removed.
"),
  nrow(verdict),
  100 * min(verdict$variance_removed), 100 * max(verdict$variance_removed),
  min(verdict$row_pearson), max(verdict$row_pearson),
  min(verdict$aggregated_pearson, na.rm = TRUE), max(verdict$aggregated_pearson, na.rm = TRUE),
  PRIMARY_MICRO,
  verdict$row_pearson[verdict$signature_name == PRIMARY_MICRO],
  verdict$aggregated_pearson[verdict$signature_name == PRIMARY_MICRO],
  100 * verdict$variance_removed[verdict$signature_name == PRIMARY_MICRO],
  PRIMARY_ASTRO,
  verdict$row_pearson[verdict$signature_name == PRIMARY_ASTRO],
  verdict$aggregated_pearson[verdict$signature_name == PRIMARY_ASTRO],
  100 * verdict$variance_removed[verdict$signature_name == PRIMARY_ASTRO]))
```

This is the direct answer to notebook 02's negative result. There, size-matched *random* gene sets
reproduced across replicate libraries as well as any candidate signature, which meant the raw score
was largely measuring something shared by every gene set of that size. The numbers above say how
much: the matched background carries most of the raw score's variance, so the raw score and the
competitive score are **not** interchangeable and the choice between them is a real decision rather
than a robustness check.

Two consequences are recorded rather than acted on unilaterally:

- both scores are carried in the analysis table, `score` and `competitive_score`, so the choice
  stays open and visible to the next step;
- the Sun axes sit at the very top of the abundance range (mean log2 CPM near 8, detected in more
  than 90% of rows), which is why they are the hardest to background-match and why the residual
  mismatch in the table above is largest for them. A marker set of highly expressed genes is
  exactly the case where an uncorrected mean-z is most exposed to recovery depth.

---

## 5. Collapsing the supertype stratum

The eventual $T$ block is per donor x ROI, so the supertype stratum has to go somewhere. Notebook
02 left the rule open; it is settled here, once, and stated on the table:

> **Rule.** A lineage-level score is the nucleus-weighted mean of its supertypes' scores.

That is the same weighting already justified for repeated measurements, for the same reason: a
supertype seen in four nuclei is not evidence on a par with one seen in four hundred. The
consequence is stated rather than hidden — a nucleus-weighted lineage score is an
*abundance-weighted* average of within-supertype states, so it is not independent of composition.
`n_supertypes` and `dominant_supertype_share` travel with every row so a later model can see how
concentrated each value is.

```{r}
#| label: collapse

lineage_level <- collapse_supertypes_to_lineage(supertype_level)

show_table(
  lineage_level %>%
    filter(signature_name %in% c(SELECTED$signature_name, COMPARATOR_MICRO)) %>%
    group_by(domain, signature_name) %>%
    summarise(donor_regions = n(), donors = n_distinct(donor_id),
              regions = n_distinct(brain_region),
              median_supertypes = median(n_supertypes),
              median_dominant_share = median(dominant_supertype_share),
              well_supported = sum(support_status == "well supported"), .groups = "drop"),
  "Donor x region x lineage estimates after collapsing supertype"
)
```

---

## 6. The analysis-ready molecular-state table

```{r}
#| label: analysis-table

molecular_state <- lineage_level %>%
  inner_join(bind_rows(SELECTED,
                       tibble(axis = "Micro/PVM", role = "comparator",
                              signature_name = COMPARATOR_MICRO)),
             by = "signature_name") %>%
  mutate(
    signature = signature_name,
    score = state_score,
    feature_origin = registry$feature_origin[match(signature_name, registry$signature_name)],
    source_release = prep_prov$source_release,
    aggregation = "nucleus-weighted mean over supertypes of the nucleus-weighted mean of assay-adjusted row scores"
  ) %>%
  select(donor_id, brain_region, lineage, axis, selection_role = role, signature,
         score, competitive_score, n_nuclei, n_measurements, n_supertypes,
         dominant_supertype_share, support_status, feature_origin, source_release,
         aggregation) %>%
  arrange(axis, selection_role, signature, donor_id, brain_region)

write_csv(molecular_state, analysis$analysis_table)

show_table(molecular_state %>% slice_head(n = 8),
           "The analysis-ready molecular-state table, first eight rows")

show_table(
  molecular_state %>% count(axis, selection_role, signature, name = "donor_region_cells"),
  sprintf("Size of the table: %d rows written to %s",
          nrow(molecular_state), basename(analysis$analysis_table))
)
```

It is deliberately narrow. A metadata field is here only if it changes interpretation or analysis:
support, how many supertypes and measurements stand behind the value, where the feature definition
came from, and which release the counts came from. Everything else stays in the upstream
derivatives. **Composition is not in this table** and is not merged into it anywhere.

```{r}
#| label: write-provenance

write_json(
  list(
    created_utc = format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z"),
    written_by = "results/analysis/03_seaad_omics_analysis_ready.qmd",
    grain = "donor x brain region x lineage x signature",
    score = "assay-adjusted, nucleus-weighted row scores, nucleus-weighted over supertypes",
    competitive_score = comp_prov$definition,
    competitive_domains = comp_prov$domains_corrected,
    selection = as.list(SELECTED),
    selection_basis = paste("biological construct, human/AD relevance, published provenance,",
                            "gene availability, interpretability; never association with CPS,",
                            "QNP, morphology or cognition"),
    source_release = prep_prov$source_release,
    composition_merged = FALSE,
    imputation = "none"
  ),
  analysis$analysis_prov, auto_unbox = TRUE, pretty = TRUE
)
cat("wrote", basename(analysis$analysis_prov), "\n")
```

---

## 7. Oligodendrocyte and OPC stay secondary

The instruction for this axis is narrow: keep it, check it against the lineage-identity control,
and if a myelination score is extremely correlated with identity, label it and stop. No score is
invented to reduce that correlation, and the axis is not deleted.

```{r}
#| label: oligo-identity

oligo_wide <- supertype_level %>%
  filter(domain %in% c("Oligodendrocyte", "OPC")) %>%
  select(donor_id, brain_region, supertype, domain, signature_name, state_score) %>%
  pivot_wider(names_from = signature_name, values_from = state_score)

identity_corr <- map_dfr(
  registry %>% filter(domain == "Oligodendrocyte", role == "candidate_state") %>%
    pull(signature_name),
  function(sig) {
    d <- oligo_wide %>% filter(domain == "Oligodendrocyte")
    ok <- !is.na(d[[sig]]) & !is.na(d[["oligodendrocyte_identity_LEIN"]])
    tibble(signature_name = sig, n = sum(ok),
           pearson_with_identity = cor(d[[sig]][ok], d[["oligodendrocyte_identity_LEIN"]][ok]),
           spearman_with_identity = cor(d[[sig]][ok], d[["oligodendrocyte_identity_LEIN"]][ok],
                                        method = "spearman"))
  }) %>%
  mutate(classification = if_else(abs(pearson_with_identity) >= 0.9,
                                  "identity_associated", "retained as secondary")) %>%
  arrange(desc(abs(pearson_with_identity)))

show_table(identity_corr,
  "Oligodendrocyte state axes against the lineage-identity control (LEIN oligodendrocyte markers), at donor x region x supertype")
```

```{r}
#| label: opc-note

cat(paste(
  "The OPC domain has no scored lineage-identity control: the LEIN oligodendrocyte marker set is",
  "registered to the Oligodendrocyte lineage and is scored only there. The OPC differentiation",
  "axis is therefore retained as a secondary biological candidate without an identity check, and",
  "that gap is recorded rather than filled by inventing a control.\n"))
```

---

## 8. Joining the selected $T$ to the existing QNP block

```{r}
#| label: ph-join

cps <- read_local_cps(omics$cps_by_region)

qnp_wide <- read_qnp_workbook(paths$qnp_2026) %>%
  qnp_primary_long() %>%
  filter(status == "observed") %>%
  select(donor_id, brain_region, variable, value) %>%
  distinct()
stopifnot(!any(duplicated(qnp_wide[c("donor_id", "brain_region", "variable")])))
qnp_wide <- pivot_wider(qnp_wide, names_from = variable, values_from = value)

t_wide <- molecular_state %>%
  filter(signature %in% c(PRIMARY_MICRO, PRIMARY_ASTRO)) %>%
  mutate(block = if_else(signature == PRIMARY_MICRO, "MicroPVM", "Astro")) %>%
  select(donor_id, brain_region, block, score, competitive_score, n_nuclei) %>%
  pivot_wider(names_from = block,
              values_from = c(score, competitive_score, n_nuclei),
              names_glue = "{block}_{.value}")

# Exact identifiers only: no region is renamed, mapped or approximated on either
# side of this join, and nothing is imputed.
donor_roi <- qnp_wide %>%
  full_join(cps %>% select(donor_id, brain_region, CPS_Local), by = c("donor_id", "brain_region")) %>%
  full_join(t_wide, by = c("donor_id", "brain_region")) %>%
  select(donor_id, brain_region, CPS_Local,
         abeta_percent_positive_area, ptau_percent_positive_area,
         neun_cells_per_area, gfap_percent_positive_area,
         MicroPVM_score, MicroPVM_competitive_score, MicroPVM_n_nuclei,
         Astro_score, Astro_competitive_score, Astro_n_nuclei)

show_table(head(donor_roi, 6),
  "The exploratory donor x ROI table. No morphology, no cognition, no imputation.")
```

### 8.1 Coverage by ROI

```{r}
#| label: coverage-by-roi

coverage_roi <- donor_roi %>%
  group_by(brain_region) %>%
  summarise(
    n_rows = n(),
    n_CPS = sum(!is.na(CPS_Local)),
    n_QNP_PH = sum(!is.na(abeta_percent_positive_area) & !is.na(ptau_percent_positive_area) &
                   !is.na(neun_cells_per_area) & !is.na(gfap_percent_positive_area)),
    n_MicroPVM = sum(!is.na(MicroPVM_score)),
    n_Astro = sum(!is.na(Astro_score)),
    n_complete_PH_T = sum(!is.na(abeta_percent_positive_area) & !is.na(ptau_percent_positive_area) &
                          !is.na(neun_cells_per_area) & !is.na(gfap_percent_positive_area) &
                          !is.na(MicroPVM_score) & !is.na(Astro_score)),
    n_complete_with_CPS = sum(!is.na(CPS_Local) &
                          !is.na(abeta_percent_positive_area) & !is.na(ptau_percent_positive_area) &
                          !is.na(neun_cells_per_area) & !is.na(gfap_percent_positive_area) &
                          !is.na(MicroPVM_score) & !is.na(Astro_score)),
    .groups = "drop") %>%
  mutate(candidate_ROI = brain_region %in% ROI) %>%
  arrange(desc(candidate_ROI), desc(n_complete_PH_T))

show_table(coverage_roi,
  "Coverage by region. The six candidate ROIs first; every released region is shown, because the ROI set is not frozen here.")
```

```{r}
#| label: coverage-totals

cat(sprintf(
  "Across the six candidate ROIs: %d donor x ROI cells have complete P and H, %d have the Micro/PVM state, %d the astrocyte state, and %d have complete P + H + both T variables (%d of those also have a local CPS).\n",
  sum(coverage_roi$n_QNP_PH[coverage_roi$candidate_ROI]),
  sum(coverage_roi$n_MicroPVM[coverage_roi$candidate_ROI]),
  sum(coverage_roi$n_Astro[coverage_roi$candidate_ROI]),
  sum(coverage_roi$n_complete_PH_T[coverage_roi$candidate_ROI]),
  sum(coverage_roi$n_complete_with_CPS[coverage_roi$candidate_ROI])))
```

---

## 9. The redundancy question

> Does each molecular-state score retain appreciable donor x region variation after global stage
> and the existing $P/H$ measurements are accounted for?

This is a **characterisation**, not a test and not a selection step. No p-value is used, nothing is
chosen by the answer, and the residual is never promoted to the canonical molecular variable.

```{r}
#| label: correlations

covariates <- c("CPS_Local", "abeta_percent_positive_area", "ptau_percent_positive_area",
                "neun_cells_per_area", "gfap_percent_positive_area")

corr_tbl <- bind_rows(
  correlate_with(donor_roi, "MicroPVM_score", covariates),
  correlate_with(donor_roi, "MicroPVM_competitive_score", covariates),
  correlate_with(donor_roi, "Astro_score", covariates),
  correlate_with(donor_roi, "Astro_competitive_score", covariates)
)

show_table(
  corr_tbl %>%
    select(response, covariate, spearman) %>%
    pivot_wider(names_from = covariate, values_from = spearman) %>%
    left_join(corr_tbl %>% group_by(response) %>%
                summarise(n_min = min(n), n_max = max(n), .groups = "drop"),
              by = "response") %>%
    relocate(n_min, n_max, .after = response),
  "Spearman correlation of each selected state score with CPS and the existing P/H variables, at donor x ROI. `n` varies by covariate because NeuN and GFAP are not released for every region, so the range is given rather than one number."
)
```


```{r}
#| label: fig-cps
#| fig-width: 9
#| fig-height: 3.4
#| fig-cap: "Selected state scores against the published local pseudo-progression score, by region. Descriptive."

donor_roi %>%
  filter(!is.na(CPS_Local)) %>%
  select(brain_region, CPS_Local, MicroPVM_score, Astro_score) %>%
  pivot_longer(c(MicroPVM_score, Astro_score), names_to = "axis", values_to = "score") %>%
  filter(!is.na(score)) %>%
  ggplot(aes(CPS_Local, score, colour = brain_region)) +
  geom_point(alpha = 0.6, size = 0.9) +
  geom_smooth(aes(group = 1), method = "loess", se = FALSE, linewidth = 0.4, colour = "black") +
  facet_wrap(~ axis, nrow = 1, scales = "free_y") +
  labs(x = "CPS_Local", y = "state score", colour = NULL)+
  theme_minimal(base_size = 20)
```

```{r}
#| label: redundancy

redundancy <- bind_rows(
  redundancy_fit(donor_roi, "MicroPVM_score"),
  redundancy_fit(donor_roi, "MicroPVM_competitive_score"),
  redundancy_fit(donor_roi, "Astro_score"),
  redundancy_fit(donor_roi, "Astro_competitive_score")
)

show_table(
  redundancy %>%
    select(any_of(c("response", "n", "n_donors", "n_regions", "sd_raw", "sd_residual",
                    "r2_full", "fraction_variance_remaining")), starts_with("unique_r2_")),
  "T ~ ROI + ns(CPS, 3) + Abeta + pTau + NeuN + GFAP, fitted only to characterise. `unique_r2_*` is the drop in R-squared when that block is removed from the full model, so the four do not sum to the full R-squared."
)
```

```{r}
#| label: redundancy-read

cat(sprintf(
  "Micro/PVM: SD %.3f raw, %.3f residual, so %.0f%% of the score's spread survives ROI, stage and P/H.\nAstrocyte: SD %.3f raw, %.3f residual, so %.0f%% survives.\n",
  redundancy$sd_raw[redundancy$response == "MicroPVM_score"],
  redundancy$sd_residual[redundancy$response == "MicroPVM_score"],
  100 * redundancy$sd_residual[redundancy$response == "MicroPVM_score"] /
        redundancy$sd_raw[redundancy$response == "MicroPVM_score"],
  redundancy$sd_raw[redundancy$response == "Astro_score"],
  redundancy$sd_residual[redundancy$response == "Astro_score"],
  100 * redundancy$sd_residual[redundancy$response == "Astro_score"] /
        redundancy$sd_raw[redundancy$response == "Astro_score"]))
```

```{r}
#| label: redundancy-contrast

# Descriptive, and *not* a selection step: the signatures were fixed in the
# setup chunk before any of these numbers existed. What this compares is two
# scorings of the same fixed signature.
contrast <- redundancy %>%
  mutate(axis = if_else(str_detect(response, "^MicroPVM"), "Micro/PVM", "Astrocyte"),
         scoring = if_else(str_detect(response, "competitive"), "competitive", "raw")) %>%
  select(axis, scoring, r2_full, unique_r2_ROI, unique_r2_CPS, unique_r2_P, unique_r2_H)

show_table(contrast,
  "The same model fitted to the raw and the competitive version of each axis. Read the columns, not the totals: where the explained variance sits matters more here than how much of it there is.")
```

---

## 10. Decision table

```{r}
#| label: decision

pull1 <- function(tbl, sig, col) tbl[[col]][match(sig, tbl$signature_name)]
agree_micro <- raw_vs_comp_agg$pearson[raw_vs_comp_agg$signature_name == PRIMARY_MICRO]
agree_astro <- raw_vs_comp_agg$pearson[raw_vs_comp_agg$signature_name == PRIMARY_ASTRO]
reliability_of <- function(sig) {
  row <- reliability %>% filter(signature_name == sig)
  removed <- verdict$variance_removed[verdict$signature_name == sig]
  sprintf(paste("ICC(1,1) raw %.2f vs raw null %.2f (does not clear it); competitive %.2f vs",
                "corrected null %.2f, %s by %.2f. The matched background accounts for %.0f%% of",
                "the raw score's variance."),
          row$icc_raw[1], row$null_icc_raw[1],
          row$icc_competitive[1], row$null_icc_competitive[1],
          if (isTRUE(row$clears_null_competitive[1])) "clearing it" else "failing it",
          abs(row$margin_over_null[1]), 100 * removed)
}
r_micro <- redundancy %>% filter(response == "MicroPVM_score")
r_astro <- redundancy %>% filter(response == "Astro_score")
oligo_worst <- identity_corr %>% slice(1)

decision <- tibble::tribble(
  ~axis, ~primary_definition, ~score, ~technical_reliability, ~PH_redundancy, ~ROI_coverage, ~status,

  "Micro/PVM",
  sprintf("Sun et al. 2023 MG4 (lipid processing), %d of %d genes usable",
          pull1(sel_cov, PRIMARY_MICRO, "n_used"), pull1(sel_cov, PRIMARY_MICRO, "n_defined")),
  "assay-adjusted, nucleus-weighted mean z; matched-background arm beside it",
  reliability_of(PRIMARY_MICRO),
  sprintf("R2 %.2f to ROI + CPS + P + H; %.0f%% of SD survives",
          r_micro$r2_full, 100 * r_micro$sd_residual / r_micro$sd_raw),
  sprintf("%d donor x ROI cells with complete P/H + both T",
          sum(coverage_roi$n_complete_PH_T[coverage_roi$candidate_ROI])),
  carry_status(PRIMARY_MICRO),

  "Astrocyte",
  sprintf("GO:0048143 astrocyte activation, %d of %d genes usable",
          pull1(sel_cov, PRIMARY_ASTRO, "n_used"), pull1(sel_cov, PRIMARY_ASTRO, "n_defined")),
  "assay-adjusted, nucleus-weighted mean z; matched-background arm beside it",
  reliability_of(PRIMARY_ASTRO),
  sprintf("R2 %.2f to ROI + CPS + P + H; %.0f%% of SD survives",
          r_astro$r2_full, 100 * r_astro$sd_residual / r_astro$sd_raw),
  sprintf("%d donor x ROI cells with complete P/H + both T",
          sum(coverage_roi$n_complete_PH_T[coverage_roi$candidate_ROI])),
  carry_status(PRIMARY_ASTRO),

  "Oligo/OPC",
  "GO:0043217 myelin maintenance (Oligo) and GO:0048709 oligodendrocyte differentiation (OPC)",
  "assay-adjusted, nucleus-weighted mean z; no matched-background arm",
  sprintf("highest correlation with the LEIN identity control is %.2f (%s)",
          oligo_worst$pearson_with_identity, oligo_worst$signature_name),
  "not characterised against P/H in this notebook",
  "retained in the analysis table, not joined to P/H here",
  if (any(identity_corr$classification == "identity_associated")) "identity_associated" else "secondary"
)

show_table(decision, "The decision. Statuses are limited to carry_forward, secondary, identity_associated, unresolved.")
```

---

## Report

### Files changed

```{r}
#| label: files-changed

show_table(
  tibble::tribble(
    ~path, ~status, ~what,
    "results/analysis/03_seaad_omics_analysis_ready.qmd", "new", "this notebook",
    "R/seaad_analysis_ready.R", "new",
      "supertype collapsing, competitive-score reader, redundancy characterisation",
    "src/seaad/omics/xlsx.py", "new",
      "minimal stdlib XLSX reader; openpyxl is absent from this project's environments",
    "src/seaad/omics/signature_registry.py", "modified",
      "second resolution route: a SHA-256-pinned published supplement; Sun MG4 and MG8 resolved",
    "src/seaad/omics/state_scoring.py", "modified",
      "additive only: per-domain abundance accumulator and matched-background draws, both off by default",
    "src/seaad/omics/competitive_scoring.py", "new", "the competitive score",
    "src/seaad/cli/omics_competitive_scores.py", "new", "its CLI",
    "src/seaad/pyproject.toml", "modified", "one new console script",
    "data/derivatives/sea-ad/omics_state/", "regenerated + new",
      "registry with two more resolved signatures; row_competitive_scores.csv, background_matching_summary.csv, molecular_state_analysis_table.csv and their provenance"
  ),
  "Files changed. Acquisition and preparation were not touched."
)
```

### Summary

```{r}
#| label: stop-conditions

answers <- tibble::tribble(
  ~question, ~answer,

  "1. What single Micro/PVM molecular-state variable goes forward?",
  sprintf(paste("Sun et al. 2023 MG4, the human lipid-processing microglial state, resolved",
                "verbatim from Table S1 page 2 of the published supplement (SHA-256 pinned,",
                "no threshold applied here). %d of %d symbols are usable in the Micro/PVM",
                "domain, and its background-corrected score reaches ICC %.2f against a",
                "corrected-null floor of %.2f, so it is %s. MG8 (inflammatory) is the secondary",
                "alternative at ICC %.2f; GO:0061900 glial cell activation is kept scored as the",
                "dataset-independent comparator and reaches only %.2f corrected."),
          pull1(sel_cov, PRIMARY_MICRO, "n_used"), pull1(sel_cov, PRIMARY_MICRO, "n_defined"),
          reliability$icc_competitive[reliability$signature_name == PRIMARY_MICRO],
          reliability$null_icc_competitive[reliability$signature_name == PRIMARY_MICRO],
          carry_status(PRIMARY_MICRO),
          reliability$icc_competitive[reliability$signature_name == "sun_2023_MG8_inflammatory_II"],
          reliability$icc_competitive[reliability$signature_name == COMPARATOR_MICRO]),

  "2. What single Astrocyte molecular-state variable goes forward?",
  sprintf(paste("GO:0048143 astrocyte activation, and it is the weaker of the two. Only %d of %d",
                "symbols are usable; its corrected score reaches ICC %.2f against a",
                "corrected-null floor of %.2f, so it is **%s** under the stated rule but clears",
                "that floor far less convincingly than MG4 does, and there is no secondary",
                "astrocyte axis to fall back on - the regulation set keeps %d usable genes,",
                "Liddelow A1/A2 is image-only, and Habib DAA is mouse-derived."),
          pull1(sel_cov, PRIMARY_ASTRO, "n_used"), pull1(sel_cov, PRIMARY_ASTRO, "n_defined"),
          reliability$icc_competitive[reliability$signature_name == PRIMARY_ASTRO],
          reliability$null_icc_competitive[reliability$signature_name == PRIMARY_ASTRO],
          carry_status(PRIMARY_ASTRO),
          coverage$n_used[coverage$signature_name == "astrocyte_activation_regulation_GOBP" &
                          coverage$domain == "Astrocyte"]),

  "3. Does matched-background correction materially change either?",
  sprintf(paste("Yes, materially, for both. The matched background accounts for %.0f%% of MG4's",
                "raw variance and %.0f%% of the astrocyte axis's, and raw and competitive scores",
                "agree at only Pearson %.2f and %.2f after aggregation. The background",
                "reproduces each signature's abundance and detection profile closely and the",
                "subtracted mean is stable over %d draws, so this is a property of the scores",
                "rather than of the correction. The two scorings are not interchangeable; both",
                "are carried in the analysis table and neither is deleted."),
          100 * verdict$variance_removed[verdict$signature_name == PRIMARY_MICRO],
          100 * verdict$variance_removed[verdict$signature_name == PRIMARY_ASTRO],
          agree_micro, agree_astro, comp_prov$background$n_draws),

  "4. How much of each score is already explained by CPS and P/H?",
  sprintf(paste("Micro/PVM R2 %.2f, astrocyte R2 %.2f against ROI + ns(CPS,3) + Abeta + pTau +",
                "NeuN + GFAP; %.0f%% and %.0f%% of each score's SD survives as residual."),
          r_micro$r2_full, r_astro$r2_full,
          100 * r_micro$sd_residual / r_micro$sd_raw,
          100 * r_astro$sd_residual / r_astro$sd_raw),

  "5. How many donor x ROI observations are complete?",
  sprintf(paste("%d across the six candidate ROIs with complete P, H and both selected T",
                "variables; %d of those also carry a published local CPS. Per-ROI counts are in",
                "section 8.1 and nothing is imputed."),
          sum(coverage_roi$n_complete_PH_T[coverage_roi$candidate_ROI]),
          sum(coverage_roi$n_complete_with_CPS[coverage_roi$candidate_ROI])),

  "6. Is there enough independent molecular variation to proceed to morphology?",
  sprintf(paste("Yes on the numbers, with one caveat. Micro/PVM keeps %.0f%% of its SD and the",
                "astrocyte axis %.0f%% after ROI, stage and P/H, over %d complete donor x ROI",
                "cells - both axes carry information that P and H do not. Note where the",
                "explained variance sits: for the raw Micro/PVM score it is almost all ROI",
                "(unique R2 %.2f) with CPS and P adding %.2f and %.2f, while for the competitive",
                "score ROI drops to %.2f and CPS and P rise to %.2f and %.2f. The caveat is that",
                "this makes the raw-versus-competitive choice a real modelling decision rather",
                "than a detail, and it is not settled here."),
          100 * r_micro$sd_residual / r_micro$sd_raw,
          100 * r_astro$sd_residual / r_astro$sd_raw,
          sum(coverage_roi$n_complete_PH_T[coverage_roi$candidate_ROI]),
          r_micro$unique_r2_ROI, r_micro$unique_r2_CPS, r_micro$unique_r2_P,
          contrast$unique_r2_ROI[contrast$axis == "Micro/PVM" & contrast$scoring == "competitive"],
          contrast$unique_r2_CPS[contrast$axis == "Micro/PVM" & contrast$scoring == "competitive"],
          contrast$unique_r2_P[contrast$axis == "Micro/PVM" & contrast$scoring == "competitive"])
)

# A zero-length argument makes sprintf() return character(0), which lands in the
# table as an empty cell rather than an error. Check for it rather than trusting
# the render.
stopifnot(all(nzchar(answers$answer)), !any(is.na(answers$answer)))

show_table(answers, "The six stop-condition questions")
```

### What is still open

```{r}
#| label: unresolved

show_table(
  tibble::tribble(
    ~issue, ~blocks_morphology, ~what_would_resolve_it,
    "Raw and competitive scorings of the same signature disagree, and the P/H model puts their explained variance in different places",
      "YES - this is the one open issue that does",
      paste("one decision, made once and stated: carry the raw mean-z, the",
            "matched-background score, or both as separate covariates. Everything needed to",
            "make it is in sections 3 and 9; nothing further has to be computed."),
    "The astrocyte axis has no human AD-derived definition, and its corrected score does not clear the corrected null",
      "only if an astrocyte T is required",
      "a machine-readable human astrocyte state cassette, if one exists; the Sun route worked for microglia and is the obvious template",
    "Micro-PVM is not resolved into microglia and perivascular macrophages",
      "no", "a finer released taxonomy",
    "The lineage score is abundance-weighted across supertypes, so it is not independent of composition",
      "no", "stated on the table; a later model must say which it attributes an effect to",
    "sample_name semantics remain undocumented",
      "no", "a statement from SEA-AD; no contradiction has arisen, so it stays closed",
    "The ROI set is still not frozen",
      "not yet", "the coverage table in section 8.1 plus the morphology ROI list",
    "The neuronal axis is undefined and nothing neuronal was downloaded",
      "no", "a defensible neuronal grouping present in every candidate ROI"
  ),
  "Open issues, and whether each one actually blocks morphology integration"
)
```
