---
title: "SEA-AD multiregional progression, tissue response and longitudinal cognitive phenotypes"
subtitle: "Released Global/Local CPS against conventional staging, MG4 and NeuN in regional tissue, and four-domain cognitive level and decline in ACT"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
    df-print: default
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

## 1. Scientific questions and cohort

This notebook links global AD progression, regional pathological involvement and cellular tissue
response to longitudinal cognitive phenotypes. Morphology is still being processed.

Its first pass established four results from direct multiregional QNP:

- MG4 is mainly donor-level and follows donor-wide tau.
- Regional NeuN is weakly lower where regional tau is higher.
- Executive decline behaves differently from memory decline.
- Multiregional tau accounts for part of the cognitive "discordance" left by the MTG-fitted CPS.

Those results are kept (Sections 3 and 6). This amended version moves the progression
representation onto the **released multiregion progression scores**, and it characterizes **all
four cognitive domains**.

1. **Disease heterogeneity.** How heterogeneous is quantitative progression within Braak and ADNC
   categories? How does local involvement vary among donors at comparable global progression?
   Which anatomical patterns appear across the available regions?
2. **Longitudinal cognitive phenotypes.** How do terminal CASI, the global slope, and the level
   and rate of memory, executive, language and visuospatial function relate to global versus
   anatomically matched local pTau progression?
3. **Tissue response.** Does MG4 follow global or region-specific pTau progression? Does regional
   NeuN vary with local pTau progression? Does MG4 carry tissue-integrity information at
   comparable local pathology?

Everything is cross-sectional postmortem tissue plus antemortem cognition. No association below
is causal, establishes temporal order or mediation, or measures circuit function.

**Progression source and definitions.** `Global_and_Local_CPS.20260105.csv` from the multiregion
release is used as released. Its columns are `CPS_Global`, `CPS_Global_pTau`, `CPS_Global_ABeta`,
`CPS_Local`, `CPS_Local_pTau` and `CPS_Local_ABeta`.

- **Not re-derived.** Nothing is re-estimated, averaged across regions, or combined into a new
  score.
- **Global and local are not independent exposures.** They are never entered as if they were, and
  local is not treated as a residual from global.
- **Distinct from notebook 04's CPS.** `CPS_Global` is not `SpecimenMetadata.csv`'s
  `Continuous Pseudo-progression Score` (the Gabitto MTG CPS; notebook 01). The Gabitto-CPS
  analyses from this notebook's first pass are kept for reference in Section 6 and are not
  repeated.

**Inputs, all reused unchanged:**

| Block | Source | Defined in |
|---|---|---|
| Released Global/Local CPS (all; Aβ; pTau) | `Global_and_Local_CPS.20260105.csv` | release; grid checked in notebook 01 |
| Braak, ADNC, co-pathology, `Severely Affected Donor` | donor-metadata workbook | notebooks 00, 04 |
| Regional QNP (AT8, 6E10, NeuN, GFAP; Grey matter) | `qnp_primary_long()`, June 2026 workbook | notebook 00 |
| MG4 and astrocyte activation, competitive scores | `molecular_state_analysis_table.csv` | notebook 03 |
| Terminal CASI | `Last CASI Score`, `Interval from last CASI in months` | notebook 04 |
| Harmonized longitudinal cognition and released slopes | `sea-ad_cohort_harmonized_cognitive_scores_20241213.xlsx` | notebook 00 |
| snRNA-seq nuclei per library (severely-affected check only) | `supertype_abundances_by_library.csv`, `row_state_scores.csv` | notebooks 01–02 |

Nothing was downloaded, rescored, re-aggregated or re-normalized.

**Not available locally:**

- The Kavli research strategy, the ACT DUA specifications, the reuse literature census and any
  resilience survey.
- `CASI_bifactor_scores`.
- A codebook for the released slopes, or their standard errors.

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(knitr)
  library(cowplot)
  library(scales)
})

find_root <- function(p = getwd()) {
  while (!dir.exists(file.path(p, ".git"))) {
    if (identical(dirname(p), p)) stop("project root not found")
    p <- dirname(p)
  }
  p
}
PROJECT_ROOT <- find_root()
source(file.path(PROJECT_ROOT, "R", "seaad_explore.R"))
source(file.path(PROJECT_ROOT, "R", "seaad_analysis_ready.R"))
source(file.path(PROJECT_ROOT, "R", "seaad_state.R"))
source(file.path(PROJECT_ROOT, "R", "seaad_multiscale.R"))

paths    <- seaad_paths(PROJECT_ROOT)
analysis <- seaad_analysis_paths(PROJECT_ROOT)
omics    <- seaad_omics_paths(PROJECT_ROOT)
state_p  <- seaad_state_paths(PROJECT_ROOT)

FIG_DIR <- file.path(PROJECT_ROOT, "results", "analysis", "figures", "05_multiregional")
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)

show_table <- function(x, caption = NULL, digits = 2, align = NULL) {
  kable(x, caption = caption, digits = digits, align = align,
        format.args = list(big.mark = ""))
}

ROI3 <- c("MTG", "DFC", "MEC")
SIG_MG4   <- "sun_2023_MG4_lipid_processing"
SIG_ASTRO <- "astrocyte_activation_GOBP"
NEUN_FLOOR_DONORS <- c("H21.33.020", "H20.33.028")   # near-zero NeuN immunoreactivity in every core ROI
B_BOOT <- 2000
BOOT_SEED <- 20260925
set.seed(BOOT_SEED)
```

```{r}
#| label: design-system

# --- figure design system ---------------------------------------------------
# Shared with notebook 04: font, inks, theme, point glyph, ROI palette,
# pseudo-log axis and export path. Palettes added here, checked with the
# dataviz validator (scripts/validate_palette.js):
#   variance components  aqua / blue / orange, reference slots 1-3, all-pairs pass;
#                        every segment also carries its share as text.
#   ADNC                 notebook 04's 4-step single-hue blue ramp (--ordinal pass).
#   Global pTau tertile  3-step single-hue blue ramp (--ordinal pass).
#   Local pTau heatmap   single-hue orange sequential ramp, so it cannot be read
#                        as the ADNC ramp.
# ROI is always encoded by colour *and* point shape; model variants by glyph.
FIG_FONT <- "Arial"
FIG_WIDTH_MM <- 180
INK   <- "#0b0b0b"
INK2  <- "#52514e"
MUTED <- "#8a8984"
GRID  <- "#ebeae6"
OUTLINE <- "#2b2b29"
MISSING <- "#bdbcb6"

PAL_ROI   <- c(MTG = "#008300", DFC = "#4a3aa7", MEC = "#e87ba4")
SHAPE_ROI <- c(MTG = 21, DFC = 22, MEC = 24)
VC_LEVELS <- c("ROI", "donor", "donor x ROI")
VC_LABELS <- c(ROI = "Region", donor = "Donor-wide", `donor x ROI` = "Within-donor + error")
PAL_VC  <- c(ROI = "#1baf7a", donor = "#2a78d6", `donor x ROI` = "#eb6834")
ADNC_LEVELS <- c("Not AD", "Low", "Intermediate", "High")
PAL_ADNC <- setNames(c("#86b6ef", "#3987e5", "#1c5cab", "#0d366b"), ADNC_LEVELS)
GT_LEVELS <- c("low Global pTau", "mid Global pTau", "high Global pTau")
PAL_GT <- setNames(c("#86b6ef", "#3987e5", "#0d366b"), GT_LEVELS)
HEAT_LOW <- "#fbefe6"; HEAT_HIGH <- "#7a2a0a"

theme_fig <- function(base_size = 8) {
  theme_classic(base_size = base_size, base_family = FIG_FONT) +
    theme(
      text = element_text(colour = INK),
      axis.text = element_text(colour = INK2, size = rel(0.9)),
      axis.title = element_text(colour = INK, size = rel(0.95)),
      axis.line = element_line(colour = MUTED, linewidth = 0.3),
      axis.ticks = element_line(colour = MUTED, linewidth = 0.3),
      panel.grid.major.y = element_line(colour = GRID, linewidth = 0.3),
      plot.title = element_text(face = "bold", size = rel(1.05), margin = margin(b = 2)),
      plot.subtitle = element_text(colour = INK2, size = rel(0.85), margin = margin(b = 4),
                                   lineheight = 1.1),
      plot.caption = element_text(colour = MUTED, size = rel(0.8), hjust = 0),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", hjust = 0, size = rel(0.95)),
      legend.title = element_text(size = rel(0.9), colour = INK, margin = margin(r = 12)),
      legend.text = element_text(size = rel(0.85), colour = INK2),
      legend.key.size = unit(3.2, "mm"),
      legend.key.spacing.x = unit(2.5, "mm"),
      legend.title.position = "left",
      legend.background = element_blank(),
      plot.margin = margin(8, 8, 6, 8),
      plot.background = element_rect(fill = "white", colour = NA)
    )
}
theme_set(theme_fig())

point_args <- list(shape = 21, size = 1.9, stroke = 0.2, colour = OUTLINE, alpha = 0.9)
roi_fill   <- function(...) scale_fill_manual(values = PAL_ROI, name = "Region", ...)
roi_shape  <- function(...) scale_shape_manual(values = SHAPE_ROI, name = "Region", ...)
adnc_fill  <- function(...) scale_fill_manual(values = PAL_ADNC, name = "ADNC", drop = FALSE, ...)

shared_legend <- function(p, position = "top") {
  cowplot::get_legend(p + theme(legend.position = position,
                                legend.justification = "center",
                                legend.box.margin = margin(0, 0, 0, 0)))
}
no_legend <- theme(legend.position = "none")

compose <- function(..., labels = "AUTO", label_size = 10, align = "hv", axis = "tblr") {
  plot_grid(..., labels = labels, label_size = label_size, label_fontfamily = FIG_FONT,
            label_fontface = "bold", align = align, axis = axis)
}

save_figure <- function(plot, name, height_mm, width_mm = FIG_WIDTH_MM) {
  plot <- ggdraw(plot) + theme(plot.background = element_rect(fill = "white", colour = NA))
  base <- file.path(FIG_DIR, name)
  ggsave(paste0(base, ".pdf"), plot, width = width_mm, height = height_mm, units = "mm",
         device = cairo_pdf)
  ggsave(paste0(base, ".png"), plot, width = width_mm, height = height_mm, units = "mm",
         dpi = 400, device = ragg::agg_png, bg = "white")
  paste0(base, ".png")
}

fmt_ci <- function(est, lo, hi, digits = 2) {
  sprintf(paste0("%.", digits, "f [%.", digits, "f, %.", digits, "f]"), est, lo, hi)
}
fmt_sci <- function(est, lo, hi, digits = 2) {
  sprintf(paste0("%+.", digits, "f [%+.", digits, "f, %+.", digits, "f]"), est, lo, hi)
}
zstd <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
scor <- function(a, b) {
  ok <- !is.na(a) & !is.na(b)
  if (sum(ok) < 5) NA_real_ else cor(a[ok], b[ok], method = "spearman")
}
qr <- function(x, d = 0) sprintf(paste0("%.", d, "f [%.", d, "f–%.", d, "f]"),
                                 median(x, na.rm = TRUE), quantile(x, .25, na.rm = TRUE),
                                 quantile(x, .75, na.rm = TRUE))

# Coefficient/interval panel shared by all figures; `variant` is encoded by
# glyph (filled = primary, open = sensitivity), never by colour.
VARIANT_SHAPES <- c(21, 1, 2, 0, 5)
forest_panel <- function(d, x_lab, title, variants) {
  d <- d %>% mutate(variant = factor(variant, levels = variants))
  dodge <- position_dodge(width = 0.6)
  ggplot(d, aes(estimate, term, shape = variant)) +
    geom_vline(xintercept = 0, colour = MUTED, linewidth = 0.3) +
    geom_linerange(aes(xmin = lo, xmax = hi, group = variant), position = dodge,
                   colour = INK, linewidth = 0.45) +
    geom_point(aes(group = variant), position = dodge, size = 1.8, stroke = 0.45,
               colour = INK, fill = INK) +
    scale_shape_manual(values = setNames(VARIANT_SHAPES[seq_along(variants)], variants),
                       name = NULL, drop = FALSE) +
    scale_y_discrete(limits = rev) +
    labs(title = title, x = x_lab, y = NULL) +
    theme(panel.grid.major.y = element_blank(),
          panel.grid.major.x = element_line(colour = GRID, linewidth = 0.3),
          legend.position = "bottom", legend.direction = "vertical",
          legend.title.position = "top")
}
```

```{r}
#| label: donor-table

donor_meta <- read_excel(paths$donor_metadata, guess_max = 1e5)
specimen   <- read.csv(paths$specimen_index, check.names = FALSE)
stopifnot(
  "one row per donor in each source" =
    !anyDuplicated(donor_meta$`Donor ID`) && !anyDuplicated(specimen$`Donor ID`),
  "the two sources carry the same 84 donors" =
    setequal(donor_meta$`Donor ID`, specimen$`Donor ID`) && nrow(donor_meta) == 84L
)

donor <- donor_meta %>%
  transmute(
    donor_id = `Donor ID`, cohort = `Primary Study Name`,
    age = `Age at Death`, sex = Sex, education = `Years of education`,
    braak = Braak, adnc = `Overall AD neuropathological Change`,
    severe = coalesce(`Severely Affected Donor`, "N") == "Y",
    casi = `Last CASI Score`, casi_interval = `Interval from last CASI in months`,
    late = LATE, arteriolo = Arteriolosclerosis
  ) %>%
  left_join(specimen %>% transmute(donor_id = `Donor ID`,
                                   cps = `Continuous Pseudo-progression Score`),
            by = "donor_id") %>%
  mutate(has_casi = !is.na(casi),
         is_act = cohort == "ACT",
         braak_stage = factor(str_remove(braak, "Braak "),
                              levels = c("0", "I", "II", "III", "IV", "V", "VI")),
         braak_group = factor(if_else(braak_stage %in% c("0", "I", "II", "III"), "0–III",
                                      as.character(braak_stage)),
                              levels = c("0–III", "IV", "V", "VI")),
         adnc = factor(adnc, levels = ADNC_LEVELS),
         late_2_3 = late %in% c("LATE Stage 2", "LATE Stage 3"),
         arteriolo_severe = arteriolo == "Severe")
stopifnot(
  "CASI present exactly for the ACT donors" = identical(donor$has_casi, donor$is_act),
  "no missing CPS, age, sex, education, Braak or ADNC" =
    !anyNA(donor[c("cps", "age", "sex", "education", "braak_stage", "adnc")]),
  "the severely-affected field is only Y or blank" =
    all(donor_meta$`Severely Affected Donor` %in% c("Y", NA))
)

# Notebook 04's baseline terminal model (Gabitto CPS). Retained only for the
# first-pass results in Section 6; nothing new is conditioned on it.
CASI_FORMULA <- casi ~ cps + age + sex + education + casi_interval
casi_set <- donor %>% filter(has_casi)
casi_fit <- lm(CASI_FORMULA, data = casi_set)
stopifnot("the CASI model uses all 69 ACT donors" = nobs(casi_fit) == 69L)
casi_set <- casi_set %>% mutate(adj_cognition = residuals(casi_fit))
donor <- donor %>% left_join(casi_set %>% select(donor_id, adj_cognition), by = "donor_id")
```

```{r}
#| label: progression-table

# The released multiregion progression scores, used as released.
cps_rel <- read_csv(omics$cps_by_region, show_col_types = FALSE) %>%
  select(donor_id = `Donor ID`, brain_region = `Brain Region`,
         CPS_Global, CPS_Global_ABeta, CPS_Global_pTau, CPS_Local, CPS_Local_ABeta, CPS_Local_pTau)
stopifnot(
  "one released CPS row per donor x region" = !anyDuplicated(cps_rel[c("donor_id", "brain_region")]),
  "every CPS donor is a known donor" = all(cps_rel$donor_id %in% donor$donor_id),
  "the Global scores are constant within donor" =
    cps_rel %>% group_by(donor_id) %>%
      summarise(k = n_distinct(CPS_Global) + n_distinct(CPS_Global_pTau) + n_distinct(CPS_Global_ABeta)) %>%
      pull(k) %>% `==`(3) %>% all(),
  "no missing released score" = !anyNA(cps_rel)
)
glob <- cps_rel %>% distinct(donor_id, CPS_Global, CPS_Global_ABeta, CPS_Global_pTau)
stopifnot("Global CPS for all 84 donors" = nrow(glob) == 84L)
donor <- donor %>% left_join(glob, by = "donor_id") %>%
  mutate(gp_tertile = factor(GT_LEVELS[findInterval(CPS_Global_pTau,
                                                    quantile(CPS_Global_pTau, c(1/3, 2/3))) + 1],
                             levels = GT_LEVELS))
REGIONS_CPS <- cps_rel %>% group_by(brain_region) %>%
  summarise(m = median(CPS_Local_pTau), .groups = "drop") %>% arrange(desc(m)) %>% pull(brain_region)
local_wide <- cps_rel %>% select(donor_id, brain_region, CPS_Local_pTau) %>%
  pivot_wider(names_from = brain_region, values_from = CPS_Local_pTau, names_prefix = "lp_")
```

```{r}
#| label: region-table

qnp_obs <- read_qnp_workbook(paths$qnp_2026) %>%
  qnp_primary_long() %>%
  filter(status == "observed") %>%
  select(donor_id, brain_region, variable, value)
stopifnot("one QNP value per donor x region x variable" =
            !anyDuplicated(qnp_obs[c("donor_id", "brain_region", "variable")]))
qnp_wide <- pivot_wider(qnp_obs, names_from = variable, values_from = value)

state_tbl <- read_csv(analysis$analysis_table, show_col_types = FALSE)
stopifnot("one molecular value per donor x region x signature" =
            !anyDuplicated(state_tbl[c("donor_id", "brain_region", "signature")]))
state_wide <- state_tbl %>%
  filter(signature %in% c(SIG_MG4, SIG_ASTRO)) %>%
  mutate(key = recode(signature, !!SIG_MG4 := "mg4", !!SIG_ASTRO := "astro")) %>%
  select(donor_id, brain_region, key, competitive_score, support_status) %>%
  pivot_wider(names_from = key, values_from = c(competitive_score, support_status))

donor_roi <- qnp_wide %>%
  full_join(state_wide, by = c("donor_id", "brain_region")) %>%
  full_join(cps_rel %>% select(donor_id, brain_region, CPS_Local_pTau, CPS_Local_ABeta),
            by = c("donor_id", "brain_region")) %>%
  left_join(glob %>% select(donor_id, CPS_Global_pTau), by = "donor_id") %>%
  mutate(neun_per_mm2 = neun_cells_per_area * 1e6)
stopifnot(
  "donor x ROI keys unique after joining QNP, molecular state and released CPS" =
    !anyDuplicated(donor_roi[c("donor_id", "brain_region")]),
  "every donor x ROI row maps to a known donor" = all(donor_roi$donor_id %in% donor$donor_id)
)

# --- core-region analysis scale (unchanged from the first pass) ------------------
sd_within_roi <- function(y, g) {
  sd(y - ave(y, g, FUN = function(v) mean(v, na.rm = TRUE)), na.rm = TRUE)
}
core <- donor_roi %>%
  filter(brain_region %in% ROI3) %>%
  group_by(brain_region) %>%
  mutate(
    at8  = normal_score(ptau_percent_positive_area),
    ab   = normal_score(abeta_percent_positive_area),
    neun = normal_score(neun_per_mm2),
    gfap = normal_score(gfap_percent_positive_area),
    at8_log = zstd(log10(ptau_percent_positive_area + 0.001)),
    ab_log  = zstd(log10(abeta_percent_positive_area + 0.001)),
    neun_flag = neun_per_mm2 < 0.1 * median(neun_per_mm2, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(mg4   = competitive_score_mg4 / sd_within_roi(competitive_score_mg4, brain_region),
         astro = competitive_score_astro / sd_within_roi(competitive_score_astro, brain_region),
         brain_region = factor(brain_region, levels = ROI3))

qnp_complete_ids <- core %>%
  group_by(donor_id) %>%
  summarise(ok = n() == 3 && !anyNA(c(at8, ab, neun, gfap)), .groups = "drop") %>%
  filter(ok) %>% pull(donor_id)
tissue <- core %>%
  filter(donor_id %in% qnp_complete_ids) %>%
  within_between(c("at8", "ab", "neun", "gfap", "at8_log", "ab_log"))

balanced_ids <- tissue %>%
  group_by(donor_id) %>%
  summarise(ok = !anyNA(c(mg4, astro)), .groups = "drop") %>%
  filter(ok) %>% pull(donor_id)
bal <- tissue %>% filter(donor_id %in% balanced_ids) %>% within_between(c("mg4", "astro"))

stopifnot(
  "the tissue table has exactly the three core ROIs per donor" = all(count(tissue, donor_id)$n == 3),
  "the balanced table is balanced" = all(count(bal, donor_id)$n == 3),
  "every molecular value used is well supported" =
    all(donor_roi$support_status_mg4[!is.na(donor_roi$competitive_score_mg4)] == "well supported"),
  "released local CPS present on every core tissue row" = !anyNA(tissue$CPS_Local_pTau)
)
dropped_qnp <- setdiff(unique(core$donor_id), qnp_complete_ids)
```

```{r}
#| label: cognition-table

cognition <- suppressMessages(read_excel(paths$cognition, guess_max = 1e5,
                                         .name_repair = "unique_quiet")) %>%
  rename(donor_id = `Donor ID`) %>%
  left_join(donor %>% select(donor_id, cohort, age), by = "donor_id")
stopifnot("harmonized study label agrees with donor cohort" =
            all((cognition$study == "ACT") == (cognition$cohort == "ACT")))

act_visits <- cognition %>% filter(cohort == "ACT")
DOMAINS <- tibble(
  domain = c("Memory", "Executive", "Language", "Visuospatial"),
  key    = c("mem", "exf", "lan", "vsp"),
  visit  = c("MEM_E", "EXF_E", "LAN_E", "VSP_E"),
  slope  = c("slope_zmem0", "slope_zexf0", "slope_zlan0", "slope_zvsp0")
)
SLOPE_GLOBAL <- "slope_zcasi_irt0"
act_slopes <- act_visits %>%
  group_by(donor_id) %>%
  summarise(across(all_of(c(SLOPE_GLOBAL, DOMAINS$slope)), ~ n_distinct(.x, na.rm = TRUE), .names = "nd_{.col}"),
            across(all_of(c(SLOPE_GLOBAL, DOMAINS$slope)), ~ first(.x)), .groups = "drop")
stopifnot("every released slope is constant within donor" =
            all(select(act_slopes, starts_with("nd_")) <= 1))
act_slopes <- select(act_slopes, -starts_with("nd_"))

dom_sum <- set_names(map(DOMAINS$visit, ~ donor_domain_summary(act_visits, .x)), DOMAINS$key)

act <- casi_set %>%
  left_join(act_slopes, by = "donor_id") %>%
  left_join(dom_sum$mem %>% select(donor_id, n_visits, first_ybd, last_ybd, span_years),
            by = "donor_id")
for (k in DOMAINS$key) {
  act <- act %>% left_join(dom_sum[[k]] %>%
                             select(donor_id, !!paste0(k, "_last") := last_score,
                                    !!paste0(k, "_level") := window_mean,
                                    !!paste0(k, "_ols") := ols_rate), by = "donor_id")
}
act <- act %>% left_join(glob, by = "donor_id") %>% left_join(local_wide, by = "donor_id") %>%
  mutate(gp10 = CPS_Global_pTau * 10)
stopifnot(
  "the cognition cohort is the 69 ACT donors" = nrow(act) == 69L && all(act$cohort == "ACT"),
  "every ACT donor has all five released slopes" = !anyNA(act[c(SLOPE_GLOBAL, DOMAINS$slope)]),
  "every ACT donor has >= 3 memory visits" = all(act$n_visits >= 3),
  "every ACT donor has a released Global pTau CPS" = !anyNA(act$CPS_Global_pTau)
)

adrc_long <- cognition %>% filter(cohort != "ACT") %>%
  group_by(donor_id) %>%
  summarise(n_mem = sum(!is.na(MEM_E)), has_slope = any(!is.na(slope_zmem0)), .groups = "drop")
```

### Cohort and coverage

```{r}
#| label: cohort

roi_n <- function(d, cond, levels = ROI3) {
  d %>% filter({{ cond }}) %>% count(brain_region) %>%
    mutate(brain_region = factor(brain_region, levels = levels)) %>% arrange(brain_region) %>%
    mutate(s = paste(brain_region, n)) %>% pull(s) %>% paste(collapse = " · ")
}
all3 <- function(d, cond) {
  d %>% group_by(donor_id) %>% summarise(k = sum({{ cond }}), .groups = "drop") %>%
    filter(k == 3) %>% nrow()
}
act_ids <- donor$donor_id[donor$is_act]
partB <- tissue %>% filter(!is.na(mg4))

cohort_tbl <- tribble(
  ~`Analysis set`, ~Donors, ~`Donor × ROI rows`, ~Unit, ~`Used in`,
  "All SEA-AD donors", sprintf("84 (ACT 69, ADRC 15; severely affected %d)", sum(donor$severe)),
    "–", "donor", "§2",
  "Released Global CPS (all, Aβ, pTau)", "84", "–", "donor", "§2–5",
  "Released Local CPS (all, Aβ, pTau)", sprintf("%d", n_distinct(cps_rel$donor_id)),
    roi_n(cps_rel, TRUE, REGIONS_CPS), "donor × ROI", "§2 (all regions); §3–5",
  "Regional QNP (AT8, 6E10, NeuN, GFAP), core",
    sprintf("%d in all three ROIs", length(qnp_complete_ids)),
    roi_n(core, !is.na(at8) & !is.na(ab) & !is.na(neun) & !is.na(gfap)), "donor × ROI", "§3",
  "MG4 competitive score, all regions", sprintf("%d in all three core ROIs", all3(core, !is.na(mg4))),
    roi_n(donor_roi %>% filter(!is.na(CPS_Local_pTau)), !is.na(competitive_score_mg4), REGIONS_CPS),
    "donor × ROI", "§3",
  "Tissue set, primary (core QNP complete; MG4 where present)",
    sprintf("%d", n_distinct(partB$donor_id)), sprintf("%d rows", nrow(partB)), "donor × ROI", "§3",
  "Tissue set, balanced (QNP + MG4 + astrocyte in all three)",
    sprintf("%d (ACT %d)", length(balanced_ids), sum(balanced_ids %in% act_ids)),
    sprintf("%d rows", nrow(bal)), "donor × ROI", "§3, §5",
  "Terminal CASI", "69 (all ACT)", "–", "donor", "§4–5",
  "Harmonized cognition, ACT",
    sprintf("69; visits %s; all four domains and all five released slopes", qr(act$n_visits)),
    "–", "donor × visit", "§4–5",
  "Harmonized cognition, ADRC",
    sprintf("%d with visits; 0 with released slopes", nrow(adrc_long)), "–", "donor × visit", "not modelled"
)
show_table(cohort_tbl, caption = "Cohorts. Rows in the tissue sets are donor × ROI and are never counted as donors. Cognition is ACT-only throughout; ADRC donors enter only disease and tissue analyses. No analysis requires every region.")

timing <- list(span = qr(act$span_years), first = qr(act$first_ybd), last = qr(act$last_ybd),
               casi_int = qr(act$casi_interval / 12, 1), last_max = max(act$last_ybd))
```

**Timing of the ACT cognitive data** (whole years; median [IQR]):

- follow-up spans `r timing$span` years;
- the first visit falls `r timing$first` years before death;
- the last harmonized visit falls `r timing$last` years before death (at most `r timing$last_max`);
- the last CASI falls `r timing$casi_int` years before death.

**Coverage decisions.**

- **Regions.** Local CPS covers nine regions. AnG, FI and ITG were profiled only in the 43
  "AD-spectrum" donors, so they enter region-wise and are never forced into a complete panel.
- **Donor-wide direct QNP.** Always the MTG/DFC/MEC mean, from donors with QNP in all three.
  `r paste(dropped_qnp, collapse = ", ")` lacks MEC QNP and is left out of that decomposition only.

---

## 2. Disease heterogeneity: conventional staging and released progression

**Staging as the reference.** Braak stage and ADNC serve as the conventional frame; neither is
read as disease time. Every quantity in this section is descriptive:

- the spread of `CPS_Global_pTau` inside each stage;
- the spread of `CPS_Local_pTau` in each region among donors in the same `CPS_Global_pTau`
  tertile;
- each region's median local pTau as global progression rises, which gives its anatomical
  ordering.

```{r}
#| label: staging-heterogeneity

stage_summary <- function(g) {
  donor %>% group_by(level = {{ g }}) %>%
    summarise(Donors = n(),
              `Global pTau CPS, median [IQR]` = qr(CPS_Global_pTau, 2),
              Range = sprintf("%.2f–%.2f", min(CPS_Global_pTau), max(CPS_Global_pTau)),
              `Global CPS, median` = sprintf("%.2f", median(CPS_Global)),
              .groups = "drop")
}
stage_tbl <- bind_rows(
  stage_summary(braak_stage) %>% filter(Donors > 0) %>% mutate(Staging = "Braak", level = as.character(level)),
  stage_summary(adnc) %>% mutate(Staging = "ADNC", level = as.character(level))
) %>% relocate(Staging) %>% rename(Category = level)
show_table(stage_tbl, caption = "Released Global pTau CPS within conventional categories, all 84 donors. Descriptive only.")

gp_by <- function(v, lvl) donor$CPS_Global_pTau[donor[[v]] == lvl]
stg <- list(
  rho_braak = cor(as.integer(donor$braak_stage), donor$CPS_Global_pTau, method = "spearman"),
  rho_adnc = cor(as.integer(donor$adnc), donor$CPS_Global_pTau, method = "spearman"),
  b5_n = length(gp_by("braak_stage", "V")),
  b5_below_b4max = sum(gp_by("braak_stage", "V") < max(gp_by("braak_stage", "IV"))),
  b5_above_b6min = sum(gp_by("braak_stage", "V") > min(gp_by("braak_stage", "VI"))),
  b5_overlap = sum(gp_by("braak_stage", "V") < max(gp_by("braak_stage", "IV")) |
                   gp_by("braak_stage", "V") > min(gp_by("braak_stage", "VI"))),
  high_n = length(gp_by("adnc", "High")),
  high_below = sum(gp_by("adnc", "High") < max(gp_by("adnc", "Intermediate"))),
  rho_specimen = cor(donor$CPS_Global, donor$cps, method = "spearman"),
  rho_global_ptau_ab = cor(donor$CPS_Global_pTau, donor$CPS_Global_ABeta, method = "spearman")
)
stopifnot("Braak V spans part of the Braak IV and VI ranges" =
            stg$b5_below_b4max > 0 && stg$b5_above_b6min > 0)
```

```{r}
#| label: local-heterogeneity

loc <- cps_rel %>%
  left_join(donor %>% select(donor_id, gp_tertile), by = "donor_id") %>%
  left_join(qnp_wide %>% select(donor_id, brain_region, ptau_percent_positive_area,
                                abeta_percent_positive_area), by = c("donor_id", "brain_region")) %>%
  mutate(brain_region = factor(brain_region, levels = REGIONS_CPS))
local_tbl <- loc %>% group_by(brain_region) %>%
  summarise(
    Donors = n(),
    `ρ local vs Global pTau` = scor(CPS_Local_pTau, CPS_Global_pTau),
    `Median local pTau by Global pTau tertile (low / mid / high)` = paste(
      sprintf("%.2f", tapply(CPS_Local_pTau, gp_tertile, median)), collapse = " / "),
    `IQR within tertile (low / mid / high)` = paste(
      sprintf("%.2f", tapply(CPS_Local_pTau, gp_tertile, IQR)), collapse = " / "),
    `ρ local pTau vs direct AT8` = scor(CPS_Local_pTau, ptau_percent_positive_area),
    `ρ local Aβ vs direct 6E10` = scor(CPS_Local_ABeta, abeta_percent_positive_area),
    .groups = "drop") %>%
  rename(Region = brain_region)
show_table(local_tbl, caption = "Released local pTau CPS by region, ordered by overall median. Global tertiles are over all 84 donors. Tertiles pool different donors per region (AnG, FI, ITG: the 43 AD-spectrum donors only). Descriptive.")

lt <- function(r, col) local_tbl[[col]][local_tbl$Region == r]
low_t <- loc %>% filter(gp_tertile == GT_LEVELS[1]) %>% group_by(brain_region) %>%
  summarise(m = median(CPS_Local_pTau), .groups = "drop") %>% arrange(desc(m))
mid_iqr <- loc %>% filter(gp_tertile == GT_LEVELS[2]) %>% group_by(brain_region) %>%
  summarise(iqr = IQR(CPS_Local_pTau), lo = quantile(CPS_Local_pTau, .1),
            hi = quantile(CPS_Local_pTau, .9), .groups = "drop")
stopifnot(
  "local pTau CPS is a near-monotone rescaling of direct AT8 in every region" =
    all(local_tbl$`ρ local pTau vs direct AT8` > 0.95),
  "MEC and HIP lead at low global progression" = all(low_t$brain_region[1:2] %in% c("MEC", "HIP")),
  "V1C is the last-involved region" = tail(REGIONS_CPS, 1) == "V1C"
)
```

**Reading.**

- **Within staging, progression is heterogeneous.** Global pTau CPS rises with Braak stage
  (ρ `r sprintf("%.2f", stg$rho_braak)`) and ADNC (ρ `r sprintf("%.2f", stg$rho_adnc)`), but the
  categories overlap widely. Of the `r stg$b5_n` Braak V donors, `r stg$b5_below_b4max` sit below
  the highest Braak IV value and `r stg$b5_above_b6min` above the lowest Braak VI value. Among the
  `r stg$high_n` High-ADNC donors, `r stg$high_below` sit below the highest Intermediate value.
  Braak V and High ADNC each span most of the global pTau range.
- **At comparable global progression, local involvement still varies.** Local and global pTau
  correlate at ρ `r sprintf("%.2f", min(local_tbl[[3]]))`–`r sprintf("%.2f", max(local_tbl[[3]]))`
  across regions. In the middle global tertile, the local pTau IQR is
  `r sprintf("%.2f", min(mid_iqr$iqr))`–`r sprintf("%.2f", max(mid_iqr$iqr))` on a 0–1 scale.
  It is widest in the neocortical association regions reached at mid-stage
  (`r paste(as.character(mid_iqr$brain_region[order(-mid_iqr$iqr)][1:3]), collapse = ", ")`),
  where donors of the same global stage differ most in how far tau has spread.
- **Anatomical pattern.** At low global progression, local pTau is already highest in
  `r paste(as.character(low_t$brain_region[1:2]), collapse = " and ")` (medians
  `r paste(sprintf("%.2f", low_t$m[1:2]), collapse = " and ")`), while V1C is essentially
  uninvolved until the high tertile. That is the expected medial-temporal → association-cortex →
  primary-cortex sequence. The two neocortical core regions differ in shape: DFC already carries
  low-level pTau at low global stage (median `r sprintf("%.2f", low_t$m[low_t$brain_region == "DFC"])`
  vs MTG `r sprintf("%.2f", low_t$m[low_t$brain_region == "MTG"])`) but rises mainly in the high
  tertile, whereas MTG rises steadily through mid-stage.
- **What the released local scores are.** Local pTau CPS correlates with direct AT8 at ρ ≥
  `r sprintf("%.2f", min(local_tbl[["ρ local pTau vs direct AT8"]]))` in every region, and local Aβ
  CPS with direct 6E10 at ρ ≥ `r sprintf("%.2f", min(local_tbl[["ρ local Aβ vs direct 6E10"]]))`.
  In rank terms they are the direct stain measurements placed on a common progression scale,
  not independent observations. For the same reason, direct QNP and CPS are never entered as
  separate exposures of one pathology in the same model.
- **Global pTau vs global CPS vs Gabitto CPS.** `CPS_Global` correlates with the Gabitto MTG CPS
  at ρ `r sprintf("%.2f", stg$rho_specimen)`, and global pTau with global Aβ at ρ
  `r sprintf("%.2f", stg$rho_global_ptau_ab)`. The three are related but distinct released
  quantities.

---

## 3. Tissue response: MG4, NeuN and regional pathology

### 3.1 Direct QNP decomposition (first pass, retained)

Every regional measure is split into a **donor-wide** mean over MTG, DFC and MEC and a **regional
deviation** from it. The fit is Mundlak within/between; the within-donor coefficient equals the
donor fixed-effects estimate.

- **QNP measures:** within-ROI normal scores.
- **MG4 and astrocyte activation:** in pooled within-ROI SD units.
- **Intervals:** `r B_BOOT`-resample donor-cluster bootstrap. (`nlme` and `lme4` are not
  installed.)

```{r}
#| label: variance-components

vc_scale <- bal %>%
  mutate(`AT8 (pTau)` = log10(ptau_percent_positive_area + 0.001),
         `6E10 (Aβ)` = log10(abeta_percent_positive_area + 0.001),
         `NeuN density` = neun_per_mm2,
         GFAP = log10(gfap_percent_positive_area),
         `MG4 state` = competitive_score_mg4,
         `Astrocyte activation` = competitive_score_astro)
VC_VARS <- c("AT8 (pTau)", "6E10 (Aβ)", "NeuN density", "GFAP", "MG4 state", "Astrocyte activation")
vc <- map_dfr(VC_VARS, ~ variance_components(vc_scale, .x)) %>%
  mutate(component = factor(component, levels = VC_LEVELS),
         variable = factor(variable, levels = rev(VC_VARS)))
vc_share <- function(v, comp = NULL) {
  x <- vc %>% filter(variable == v)
  if (is.null(comp)) x$donor_share_after_roi[1] else x$share[x$component == comp]
}
```

```{r}
#| label: partB-model

MG4_TERMS <- c(at8_donor = "AT8, donor-wide", at8_dev = "AT8, regional deviation",
               ab_donor = "6E10, donor-wide", ab_dev = "6E10, regional deviation",
               brain_regionDFC = "DFC vs MTG", brain_regionMEC = "MEC vs MTG")
fit_state <- function(outcome, log_scale = FALSE) {
  terms <- if (log_scale) c("at8_log_donor", "at8_log_dev", "ab_log_donor", "ab_log_dev")
           else c("at8_donor", "at8_dev", "ab_donor", "ab_dev")
  f <- reformulate(c("brain_region", terms), response = outcome)
  function(d) {
    cf <- coef(lm(f, data = d))
    out <- cf[c(terms, "brain_regionDFC", "brain_regionMEC")]
    names(out) <- names(MG4_TERMS)
    out
  }
}
partB_data <- function(outcome) tissue %>% filter(!is.na(.data[[outcome]]))

state_fits <- bind_rows(
  cluster_bootstrap(partB_data("mg4"), fit_state("mg4"), B_BOOT, seed = BOOT_SEED) %>%
    mutate(outcome = "MG4", variant = "Primary (all rows)"),
  cluster_bootstrap(bal, fit_state("mg4"), B_BOOT, seed = BOOT_SEED) %>%
    mutate(outcome = "MG4", variant = "Balanced donors"),
  cluster_bootstrap(partB_data("mg4"), fit_state("mg4", log_scale = TRUE), B_BOOT, seed = BOOT_SEED) %>%
    mutate(outcome = "MG4", variant = "Log-scale pathology"),
  cluster_bootstrap(partB_data("astro"), fit_state("astro"), B_BOOT, seed = BOOT_SEED) %>%
    mutate(outcome = "Astrocyte activation", variant = "Primary (all rows)")
) %>% mutate(label = MG4_TERMS[term])
stopifnot("every bootstrap resample fitted" = all(state_fits$n_boot_ok == B_BOOT))

sf <- function(o, t, v = "Primary (all rows)") state_fits %>% filter(outcome == o, term == t, variant == v)
sf_ci <- function(o, t, v = "Primary (all rows)") { x <- sf(o, t, v); fmt_sci(x$estimate, x$lo, x$hi) }
stopifnot(
  "MG4 rises with donor-wide AT8 (interval excludes zero)" = sf("MG4", "at8_donor")$lo > 0,
  "within-donor AT8 deviation is smaller than the between-donor term for MG4" =
    sf("MG4", "at8_dev")$estimate < sf("MG4", "at8_donor")$estimate
)
```

<details><summary>Direct-QNP state model table</summary>

```{r}
state_fits %>%
  mutate(cell = fmt_sci(estimate, lo, hi)) %>%
  select(outcome, variant, label, cell) %>%
  pivot_wider(names_from = label, values_from = cell) %>%
  rename(State = outcome, Model = variant) %>%
  show_table(caption = sprintf("State ~ ROI + donor-wide and regional-deviation AT8 and 6E10 (SD of state per SD of pathology; donor-cluster bootstrap 95%% CI). Primary: %d donors, %d rows; balanced: %d donors, %d rows.", n_distinct(partB$donor_id), nrow(partB), length(balanced_ids), nrow(bal)))
```

</details>

**First-pass reading, unchanged.**

- **Donor-level states.** After region differences are removed, a donor's shared level carries
  `r percent(vc_share("MG4 state"), 1)` of MG4 variance and
  `r percent(vc_share("Astrocyte activation"), 1)` of astrocyte-activation variance.
- **MG4 and tau.** MG4 follows donor-wide AT8 (`r sf_ci("MG4", "at8_donor")` SD per SD) more than
  regional AT8 deviation (`r sf_ci("MG4", "at8_dev")`).
- **Amyloid** adds nothing (`r sf_ci("MG4", "ab_donor")`).
- **GFAP** is the measure dominated by within-donor regional variation.

### 3.2 MG4 against released global and local pTau progression

Three quantities on the same donor × ROI rows, each a separate model rather than one joint model
of global and local:

- **(a) Global:** `MG4 ~ ROI + CPS_Global_pTau`.
- **(b) Local, pooled:** `MG4 ~ ROI + CPS_Local_pTau`.
- **(c) Local, within donor:** `MG4 ~ ROI + CPS_Local_pTau + donor`, with donor fixed effects.
  Global progression is constant within a donor, so (c) asks whether a donor's more-involved
  regions carry more MG4, with no averaging of CPS across regions and no residualization.

(a) and (b) are also compared by R², on identical rows. Exposures are per SD over the model rows.

- **Primary:** the three core ROIs.
- **Sensitivity:** every region with MG4 and local CPS.

```{r}
#| label: mg4-cps

mg4_rows <- function(regions) {
  donor_roi %>%
    filter(brain_region %in% regions, !is.na(competitive_score_mg4), !is.na(CPS_Local_pTau)) %>%
    mutate(mg4 = competitive_score_mg4 / sd_within_roi(competitive_score_mg4, brain_region),
           gz = zstd(CPS_Global_pTau), lz = zstd(CPS_Local_pTau),
           brain_region = factor(brain_region))
}
fit_mg4_cps <- function(d) {
  mg <- lm(mg4 ~ brain_region + gz, d)
  ml <- lm(mg4 ~ brain_region + lz, d)
  mw <- lm(mg4 ~ brain_region + lz + factor(donor_id), d)
  c(global = unname(coef(mg)["gz"]), local = unname(coef(ml)["lz"]),
    within = unname(coef(mw)["lz"]),
    r2_global = summary(mg)$r.squared, r2_local = summary(ml)$r.squared,
    dr2 = summary(ml)$r.squared - summary(mg)$r.squared)
}
mg4_core <- mg4_rows(ROI3)
mg4_all  <- mg4_rows(REGIONS_CPS)
mg4_cps <- bind_rows(
  cluster_bootstrap(mg4_core, fit_mg4_cps, B_BOOT, seed = BOOT_SEED) %>% mutate(set = "Core ROIs"),
  cluster_bootstrap(mg4_all, fit_mg4_cps, B_BOOT, seed = BOOT_SEED) %>% mutate(set = "All regions")
)
stopifnot("every bootstrap resample fitted" = all(mg4_cps$n_boot_ok == B_BOOT))
mc <- function(t, s = "Core ROIs") mg4_cps %>% filter(term == t, set == s)
mc_ci <- function(t, s = "Core ROIs", d = 2) { x <- mc(t, s); fmt_sci(x$estimate, x$lo, x$hi, d) }
stopifnot(
  "MG4 rises with Global pTau CPS" = mc("global")$lo > 0,
  "local pTau does not fit MG4 better than global in either set" =
    mc("dr2")$lo < 0.02 && mc("dr2", "All regions")$lo < 0.02,
  "the within-donor local term includes zero in both sets" =
    mc("within")$lo < 0 && mc("within", "All regions")$lo < 0
)

mg4_cps %>%
  filter(term %in% c("global", "local", "within", "dr2")) %>%
  mutate(term = recode(term, global = "(a) Global pTau, SD per SD", local = "(b) Local pTau pooled, SD per SD",
                       within = "(c) Local pTau within donor, SD per SD", dr2 = "R²(b) − R²(a)"),
         cell = fmt_sci(estimate, lo, hi)) %>%
  select(term, set, cell) %>% pivot_wider(names_from = set, values_from = cell) %>%
  rename(Quantity = term) %>%
  show_table(caption = sprintf("MG4 against released pTau progression; donor-cluster bootstrap 95%% CI. Core: %d rows, %d donors; all regions: %d rows (%s), %d donors.", nrow(mg4_core), n_distinct(mg4_core$donor_id), nrow(mg4_all), roi_n(mg4_all, TRUE, REGIONS_CPS), n_distinct(mg4_all$donor_id)))
```

**Reading.** MG4 follows **global** pTau progression:

- **Global:** `r mc_ci("global")` SD per SD in the core ROIs; `r mc_ci("global", "All regions")`
  across all regions.
- **Local, pooled:** fits MG4 no better (R² difference `r mc_ci("dr2")`; all regions
  `r mc_ci("dr2", "All regions")`).
- **Local, within donor:** a donor's more-involved regions do not carry clearly more MG4
  (`r mc_ci("within")`; all regions `r mc_ci("within", "All regions")`).

This reproduces the direct-QNP result of 3.1 on the released progression scale. MG4 is
predominantly a donor-level response that scales with global pTau stage, and only weakly, if at
all, a regional one.

### 3.3 Regional NeuN, local pTau progression and MG4

NeuN is a direct QNP measurement (within-ROI normal score of NeuN-positive cells per area). It is
related to local pTau through the **pTau-specific** released score and through direct AT8, in
separate models, never jointly.

`CPS_Local_pTau` is released as pTau-specific. Its construction is not documented locally, but it
tracks direct AT8 at ρ ≥ 0.97 in every region (Section 2). Its correlation with NeuN matches that
of direct AT8, which gives no sign that NeuN is among its inputs. The combined `CPS_Local` and
the Gabitto CPS are not used with NeuN, since they may contain it.

**Models, donor fixed effects throughout:**

- **Within-donor coupling:** `NeuN ~ ROI + local pTau + donor`.
- **MG4 at comparable local pathology:** `NeuN ~ ROI + local pTau + MG4 + donor`.
- **First-pass direct-AT8 model**, for comparison: the Mundlak model with donor-wide and regional
  AT8 and MG4, on balanced donors.

```{r}
#| label: neun-flags

flagged <- core %>% filter(neun_flag) %>%
  transmute(donor_id, ROI = as.character(brain_region),
            `NeuN per mm²` = round(neun_per_mm2, 1), `AT8 % area` = signif(ptau_percent_positive_area, 2))
flag_donors <- unique(flagged$donor_id)
severe_ids <- donor$donor_id[donor$severe]
stopifnot("the two near-zero NeuN donors are the ones flagged in all three ROIs" =
            setequal(flagged %>% count(donor_id) %>% filter(n == 3) %>% pull(donor_id), NEUN_FLOOR_DONORS))
```

```{r}
#| label: neun-models

NEUN_DATA <- tissue %>% filter(!is.na(mg4)) %>% mutate(lz = zstd(CPS_Local_pTau))
fit_neun_fe <- function(d) {
  m1 <- lm(neun ~ brain_region + lz + factor(donor_id), d)
  m2 <- lm(neun ~ brain_region + lz + mg4 + factor(donor_id), d)
  m3 <- lm(neun ~ brain_region + at8 + mg4 + factor(donor_id), d)
  c(local_ptau = unname(coef(m1)["lz"]),
    mg4_given_local = unname(coef(m2)["mg4"]),
    mg4_given_at8 = unname(coef(m3)["mg4"]))
}
fit_neun_mundlak <- function(d) {
  f <- lm(neun ~ brain_region + at8_donor + at8_dev + mg4_donor + mg4_dev, d)
  c(at8_dev = unname(coef(f)["at8_dev"]), mg4_dev = unname(coef(f)["mg4_dev"]))
}
NEUN_VARIANTS <- list(`All donors` = character(0),
                      `Without the 2 NeuN-floor donors` = NEUN_FLOOR_DONORS,
                      `Without all severely affected donors` = severe_ids)
neun_fits <- imap_dfr(NEUN_VARIANTS, function(drop, nm) {
  bind_rows(
    cluster_bootstrap(NEUN_DATA %>% filter(!donor_id %in% drop), fit_neun_fe, B_BOOT, seed = BOOT_SEED),
    cluster_bootstrap(bal %>% filter(!donor_id %in% drop), fit_neun_mundlak, B_BOOT, seed = BOOT_SEED)
  ) %>% mutate(variant = nm,
               n_donors = n_distinct(NEUN_DATA$donor_id[!NEUN_DATA$donor_id %in% drop]))
})
stopifnot("every bootstrap resample fitted" = all(neun_fits$n_boot_ok == B_BOOT))
NEUN_LABELS <- c(local_ptau = "Local pTau CPS → NeuN (within donor)",
                 at8_dev = "Direct AT8 deviation → NeuN (first pass)",
                 mg4_given_local = "MG4 → NeuN at comparable local pTau CPS",
                 mg4_given_at8 = "MG4 → NeuN at comparable direct AT8",
                 mg4_dev = "MG4 deviation → NeuN (first pass)")
nf <- function(t, v = "All donors") neun_fits %>% filter(term == t, variant == v)
nf_ci <- function(t, v = "All donors") { x <- nf(t, v); fmt_sci(x$estimate, x$lo, x$hi) }
stopifnot(
  "local pTau and NeuN: negative, interval reaching zero" =
    nf("local_ptau")$estimate < 0 && nf("local_ptau")$hi > -0.02,
  "MG4 at comparable local pTau: positive point estimate in every variant" =
    all(neun_fits$estimate[neun_fits$term == "mg4_given_local"] > 0),
  "dropping the two NeuN-floor donors moves the MG4 terms by < 0.05 SD" =
    abs(nf("mg4_given_local", "Without the 2 NeuN-floor donors")$estimate - nf("mg4_given_local")$estimate) < 0.05 &&
    abs(nf("mg4_dev", "Without the 2 NeuN-floor donors")$estimate - nf("mg4_dev")$estimate) < 0.05
)

neun_fits %>%
  mutate(Term = NEUN_LABELS[term], cell = fmt_sci(estimate, lo, hi)) %>%
  select(Term, variant, cell) %>% pivot_wider(names_from = variant, values_from = cell) %>%
  show_table(caption = sprintf("NeuN (within-ROI normal score) models, SD of NeuN per SD of predictor; donor-cluster bootstrap 95%% CI. Fixed-effects rows: %d donors, %d rows (core ROIs with MG4); first-pass Mundlak rows: balanced donors. Severely affected donors: %d.", n_distinct(NEUN_DATA$donor_id), nrow(NEUN_DATA), length(severe_ids)))
```

**Reading.**

- **NeuN and local pTau progression.** Within donors, NeuN is lower where local pTau progression
  is higher (`r nf_ci("local_ptau")` SD per SD), and the interval reaches zero. Direct AT8 gives
  the same result (`r nf_ci("at8_dev")`). It is a modest, imprecise coupling, the same on both
  pathology representations.
- **MG4 at comparable local pathology.** It adds at most weak information:
  - at comparable local pTau CPS: `r nf_ci("mg4_given_local")`;
  - at comparable direct AT8: `r nf_ci("mg4_given_at8")`;
  - the first-pass regional-deviation estimate: `r nf_ci("mg4_dev")`.

  The direction is consistent (relatively MG4-high regions have slightly higher NeuN), but only
  the first-pass Mundlak interval clears zero. It does so barely, and not once the two NeuN-floor
  donors are removed (`r nf_ci("mg4_dev", "Without the 2 NeuN-floor donors")`). It is not evidence
  that MG4 preserves neurons.

### 3.4 Supporting check: the two donors with near-zero NeuN immunoreactivity

```{r}
#| label: severe-check

# snRNA-seq neuronal share of nuclei per donor x region, from existing derivatives:
# library-level supertype abundances, mapped to donors through the library IDs in
# row_state_scores.csv. A different tissue block from the QNP section.
lib_map <- read_csv(state_p$row_scores, show_col_types = FALSE, progress = FALSE,
                    col_select = c(donor_id, brain_region, library_prep)) %>% distinct()
stopifnot("each library maps to one donor x region" =
            !anyDuplicated(lib_map[c("library_prep", "brain_region")]))
neuronal_share <- read_csv(omics$abundances, show_col_types = FALSE) %>%
  group_by(library_prep, brain_region) %>%
  summarise(neu = sum(n_nuclei[source_object == "neuronal"]), tot = sum(n_nuclei), .groups = "drop") %>%
  inner_join(lib_map, by = c("library_prep", "brain_region")) %>%
  group_by(donor_id, brain_region) %>%
  summarise(neuronal_share = sum(neu) / sum(tot), .groups = "drop") %>%
  filter(brain_region %in% ROI3) %>%
  group_by(brain_region) %>% mutate(share_pct = percent_rank(neuronal_share)) %>% ungroup()

roi_med <- qnp_wide %>% filter(brain_region %in% ROI3) %>% group_by(brain_region) %>%
  summarise(neun_med = median(neun_cells_per_area * 1e6, na.rm = TRUE), .groups = "drop")
qnp_raw <- read_qnp_workbook(paths$qnp_2026)
cell_area <- map_dfr(qnp_raw[c("MTG", "DFC", "Hip-MEC")], ~ .x %>%
    select(`Donor ID`, `brain region`, `analysis region`, `average NeuN positive cell area`)) %>%
  filter(`analysis region` == "Grey matter", `brain region` %in% ROI3) %>%
  select(donor_id = `Donor ID`, brain_region = `brain region`,
         cell_area = `average NeuN positive cell area`) %>%
  group_by(brain_region) %>% mutate(area_ratio = cell_area / median(cell_area, na.rm = TRUE)) %>% ungroup()

severe_tbl <- core %>% filter(donor_id %in% NEUN_FLOOR_DONORS) %>%
  mutate(brain_region = as.character(brain_region)) %>%
  left_join(roi_med, by = "brain_region") %>%
  left_join(cell_area, by = c("donor_id", "brain_region")) %>%
  left_join(neuronal_share, by = c("donor_id", "brain_region")) %>%
  left_join(donor %>% select(donor_id, severe, braak, adnc, casi), by = "donor_id") %>%
  transmute(Donor = donor_id, `Severely affected` = if_else(severe, "Y", "–"),
            Stage = paste(braak, "·", adnc, "ADNC"), `Last CASI` = casi, ROI = brain_region,
            `NeuN / mm² (ROI median)` = sprintf("%.1f (%.0f)", neun_per_mm2, neun_med),
            `NeuN cell area vs median` = sprintf("%.2f×", area_ratio),
            `snRNA neuronal share (ROI percentile)` = sprintf("%.2f (%.0f%%)", neuronal_share, 100 * share_pct),
            `Local pTau CPS` = sprintf("%.2f", CPS_Local_pTau))
show_table(severe_tbl, caption = "The two donors with near-zero NeuN immunoreactivity in all core ROIs. snRNA-seq share = neuronal nuclei / all nuclei recovered and retained in the release across the donor's libraries in that region, from a separate tissue block. It reflects nuclei recovery, QC retention and SEA-AD's nuclei enrichment design as well as tissue composition, and is not a neuron count.")

sev <- list(
  both_severe = all(donor$severe[donor$donor_id %in% NEUN_FLOOR_DONORS]),
  n_severe = sum(donor$severe),
  flagged_rows_severe = sum(flagged$donor_id %in% severe_ids),
  share_pct_max = max(neuronal_share$share_pct[neuronal_share$donor_id %in% NEUN_FLOOR_DONORS]),
  severe_gp = median(donor$CPS_Global_pTau[donor$severe]),
  other_gp = median(donor$CPS_Global_pTau[!donor$severe])
)
stopifnot("both near-zero NeuN donors carry the published severely-affected flag" = sev$both_severe,
          "both are in the lowest decile of snRNA neuronal share in every core ROI" = sev$share_pct_max < 0.1)
```

**Reading.** Three things are kept apart here:

- **neuronal loss**, meaning fewer neurons;
- **reduced NeuN immunoreactivity or an altered neuronal phenotype**, meaning neurons present but
  poorly labelled;
- **differential recovery or representation of neuronal nuclei** in snRNA-seq.

**What the release documents.** Both donors carry the published `Severely Affected Donor` flag
(`r sev$n_severe` donors in all). `r sev$flagged_rows_severe` of the `r nrow(flagged)`
sub-10%-of-median NeuN rows belong to severely affected donors. The dataset description (Gabitto
et al., Research Square preprint rs-2921860) says these donors were identified by extreme
snRNA-seq/snATAC-seq QC:

- fewer UMIs and genes per nucleus;
- a lower fraction of nuclei passing QC;
- reduced nuclear-localized RNA.

The authors read this as transcriptional shutdown. They also report that these donors "had a
pronounced reduction in neuron labeling (not due entirely to cell loss) for NeuN" and steeper
late-life memory decline.

**What this notebook observes.**

- **NeuN immunoreactivity** is near zero in every core ROI (down to 0.3 cells/mm²), and the detected
  NeuN-positive profiles are about half to two-thirds of the usual size in MTG and DFC.
- **snRNA-seq neuronal share** is the lowest in every core ROI (≤ the
  `r round(100 * sev$share_pct_max)`th percentile).

**What that supports.** A severe neuronal phenotype with markedly reduced NeuN immunoreactivity.
Absolute neuronal loss is **neither established nor excluded**:

- reduced NeuN antigenicity can mimic loss in IHC;
- because these donors were defined partly by poor nuclei QC, their low snRNA-seq neuronal share
  may reflect differential recovery or QC retention of neuronal nuclei rather than neuron number.

The two observations are consistent with each other, but they are not independent confirmation of
depletion.

- **The MG4–NeuN reading does not depend on them.** Removing these two donors changes the MG4
  terms by less than 0.05 SD. Removing all severely affected donors moves them within their
  intervals (table in 3.3). They stay in every primary analysis, and the severely-affected flag is
  not used as an explanatory variable anywhere.

```{r}
#| label: fig1-build

# --- A. staging vs released global pTau progression ---------------------------------
f1a <- ggplot(donor, aes(braak_stage, CPS_Global_pTau)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, colour = INK2, fill = NA, linewidth = 0.3) +
  geom_point(aes(fill = adnc), shape = 21, size = 1.7, stroke = 0.2, colour = OUTLINE,
             position = position_jitter(width = 0.15, height = 0, seed = 11)) +
  geom_text(data = donor %>% count(braak_stage), aes(braak_stage, 1.07, label = n),
            size = 2.3, colour = INK2, family = FIG_FONT) +
  adnc_fill() +
  scale_x_discrete(drop = TRUE) +
  scale_y_continuous(limits = c(0, 1.1), breaks = seq(0, 1, 0.25)) +
  labs(title = "Global pTau progression within staging",
       x = "Braak stage (donor counts above)", y = "CPS_Global_pTau") +
  theme(legend.position = "bottom")

# --- B. donor x region local pTau --------------------------------------------------------
heat <- expand_grid(donor_id = donor$donor_id, brain_region = REGIONS_CPS) %>%
  left_join(cps_rel %>% select(donor_id, brain_region, CPS_Local_pTau), by = c("donor_id", "brain_region")) %>%
  mutate(brain_region = factor(brain_region, levels = rev(REGIONS_CPS)))
heat$x <- match(heat$donor_id, donor$donor_id[order(donor$CPS_Global_pTau)])
f1b <- ggplot(heat, aes(x, brain_region, fill = CPS_Local_pTau)) +
  geom_tile(colour = "white", linewidth = 0.15) +
  scale_fill_gradient(low = HEAT_LOW, high = HEAT_HIGH, na.value = MISSING, limits = c(0, 1),
                      name = "CPS_Local_pTau", breaks = c(0, 0.5, 1)) +
  scale_x_continuous(expand = expansion(0), breaks = c(1, 84), labels = c("lowest", "highest")) +
  labs(title = "Local pTau across regions",
       subtitle = "84 donors ordered by Global pTau; grey = region not profiled",
       x = "Donors, by CPS_Global_pTau", y = NULL) +
  theme(axis.line = element_blank(), axis.ticks = element_blank(), panel.grid.major.y = element_blank(),
        axis.text.x = element_text(hjust = c(0, 1)), plot.margin = margin(8, 14, 6, 8),
        legend.position = "bottom", legend.key.width = unit(8, "mm"), legend.key.height = unit(2.5, "mm"))

# --- C. local vs global in the core ROIs ----------------------------------------------------
f1c_data <- donor_roi %>% filter(brain_region %in% ROI3, !is.na(CPS_Local_pTau)) %>%
  mutate(brain_region = factor(brain_region, levels = ROI3))
f1c <- ggplot(f1c_data, aes(CPS_Global_pTau, CPS_Local_pTau)) +
  geom_point(aes(fill = brain_region, shape = brain_region), size = 1.4, stroke = 0.2,
             colour = OUTLINE, alpha = 0.85) +
  roi_fill() + roi_shape() +
  scale_x_continuous(limits = c(0, 1)) + scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Local at comparable global",
       x = "CPS_Global_pTau (donor)", y = "CPS_Local_pTau (region)") +
  theme(legend.position = "bottom")

# --- D. variance components (first pass) ------------------------------------------------------
f1d <- ggplot(vc, aes(share, variable, fill = component)) +
  geom_col(width = 0.7, colour = "white", linewidth = 0.5, position = position_stack(reverse = TRUE)) +
  geom_text(aes(label = if_else(share >= 0.07, percent(share, 1), "")),
            position = position_stack(vjust = 0.5, reverse = TRUE),
            size = 2.2, family = FIG_FONT, colour = "white", fontface = "bold") +
  scale_fill_manual(values = PAL_VC, labels = VC_LABELS, name = NULL) +
  scale_x_continuous(labels = percent, expand = expansion(0), breaks = seq(0, 1, 0.25)) +
  labs(title = "Where tissue variance sits",
       subtitle = sprintf("%d donors × MTG, DFC, MEC", length(balanced_ids)),
       x = "Share of variance", y = NULL) +
  guides(fill = guide_legend(nrow = 1)) +
  theme(panel.grid.major.y = element_blank(), axis.line.y = element_blank(),
        axis.ticks.y = element_blank(), legend.position = "bottom")

# --- E. MG4 against progression ----------------------------------------------------------------
f1e_data <- bind_rows(
  mg4_cps %>% filter(term %in% c("global", "local", "within")) %>%
    mutate(term = recode(term, global = "Global pTau CPS", local = "Local pTau CPS, pooled",
                         within = "Local pTau CPS, within donor"),
           variant = set),
  state_fits %>% filter(outcome == "MG4", variant == "Primary (all rows)", term %in% c("at8_donor", "at8_dev")) %>%
    mutate(term = recode(term, at8_donor = "Direct AT8, donor-wide", at8_dev = "Direct AT8, regional deviation"),
           variant = "Core ROIs")
) %>% mutate(term = factor(term, levels = c("Global pTau CPS", "Local pTau CPS, pooled",
                                            "Local pTau CPS, within donor", "Direct AT8, donor-wide",
                                            "Direct AT8, regional deviation")))
f1e <- forest_panel(f1e_data, "SD of MG4 per SD of pathology", "MG4 follows global progression",
                    variants = c("Core ROIs", "All regions")) +
  guides(shape = guide_legend(nrow = 1)) + theme(legend.direction = "horizontal")

# --- F. NeuN models -----------------------------------------------------------------------------
f1f_data <- neun_fits %>%
  mutate(term = factor(NEUN_LABELS[term], levels = NEUN_LABELS[c("local_ptau", "at8_dev", "mg4_given_local",
                                                                 "mg4_given_at8", "mg4_dev")]))
f1f <- forest_panel(f1f_data, "SD of NeuN per SD of predictor", "NeuN, local pathology and MG4",
                    variants = names(NEUN_VARIANTS)) +
  scale_y_discrete(limits = rev, labels = label_wrap(26))

fig1 <- plot_grid(
  compose(f1a, f1b, labels = c("A", "B"), align = "h", axis = "b", rel_widths = c(0.9, 1.1)),
  compose(f1c, f1d, labels = c("C", "D"), align = "h", axis = "b", rel_widths = c(0.75, 1.25)),
  compose(f1e, f1f, labels = c("E", "F"), align = "h", axis = "b"),
  ncol = 1, rel_heights = c(1, 0.95, 1.15)
)
fig1_png <- save_figure(fig1, "fig1_progression_tissue_response", height_mm = 230)
fig1_cap <- sprintf(paste0(
  "**Figure 1.** Disease progression and regional tissue response, all SEA-AD donors (ACT and ADRC). ",
  "**A** Released `CPS_Global_pTau` by Braak stage (no donor at stage I), fill = ADNC; boxes are quartiles. ",
  "**B** Released `CPS_Local_pTau` for every donor × region, donors ordered by Global pTau and regions by median local pTau; AnG, FI and ITG were profiled only in the 43 AD-spectrum donors. ",
  "**C** Local against global pTau progression in MTG, DFC and MEC (donor × ROI points). Local is not a residual of global, and no reference line is implied. ",
  "**D** Variance components of direct QNP and molecular measures over %d donors with all three core ROIs (first pass): region, donor-wide level, and within-donor regional variation plus measurement error. ",
  "**E** MG4 (SD units) against pathology in separate models: global pTau CPS; local pTau CPS pooled and within donor (donor fixed effects); first-pass direct-AT8 donor-wide and regional-deviation terms. Core ROIs: %d rows, %d donors; all regions: %d rows. ",
  "**F** NeuN (within-ROI normal score) within donors against local pTau CPS or direct AT8, and MG4 at comparable local pathology, with the two donors with near-zero NeuN immunoreactivity or all %d severely affected donors removed. NeuN density measures NeuN-immunoreactive profiles, not absolute neuron number. ",
  "Donor-cluster bootstrap 95%% CI (%d resamples); donors are the resampling unit. Associations are cross-sectional and not causal; QNP and snRNA-seq come from the same ROI, not the same microscopic field."),
  length(balanced_ids), nrow(mg4_core), n_distinct(mg4_core$donor_id), nrow(mg4_all),
  length(severe_ids), B_BOOT)
```

```{r}
#| label: fig1
#| fig-cap: !expr fig1_cap
#| out-width: "100%"

include_graphics(fig1_png)
```

---

## 4. Longitudinal cognitive phenotypes (ACT, 69 donors)

### 4.1 Released variables: documentation, observations and open questions

The harmonized workbook is a single sheet with no notes, codebook or methods text. The SEA-AD data
dictionary covers only `Last CASI Score` and its interval.

One focused search of the accessible methods literature found two usable sources:

- **The harmonization pipeline.** It is described in Mukherjee et al. (2023, *Neuropsychology* 37:
  409, doi 10.1037/neu0000835; full text not accessible here) and, in methods text, in the
  cohort-pooled genome-wide study of these scores (*Mol Neurodegener* 2023, PMC10286470).
- **The dataset paper.** The Gabitto et al. preprint cites Crane et al. (2022, *Alzheimer's &
  Dementia* 18:e064534) for the domain scores. Crane and Mukherjee are listed data contributors of
  the SEA-AD metadata release. The preprint's full Methods were not accessible.

```{r}
#| label: cognitive-provenance

prov <- tribble(
  ~Item, ~`Documented (source)`, ~`Observed here`, ~`Unresolved`,
  "`MEM_E`, `EXF_E`, `LAN_E`, `VSP_E`",
    "Harmonized, co-calibrated domain scores from confirmatory factor analysis (Mplus); anchor items shared across cohorts put studies on one metric; single-factor or bifactor model chosen per study; item parameters calibrated at the last visit and fixed to score every visit (PMC10286470; Mukherjee 2023)",
    "ACT visit-level SD 0.43–0.57 and positive means, so scores are not standardized to this cohort",
    "Reference sample and target mean/SD of the metric; which calibration release produced this file",
  "`*_E_SE`",
    "Co-calibrated scores carry a per-score standard error; the pooled genetic study excluded scores with SE > 0.6 (PMC10286470)",
    "Present at visit level; see the SE table below",
    "Whether SEA-AD applied any SE-based exclusion",
  "Domain slopes `slope_zmem0`, `slope_zexf0`, `slope_zlan0`, `slope_zvsp0`",
    "None located. The dataset paper reports memory 'slopes' (−0.15 in severely affected vs −0.11 in ADNC 3 donors) without a method in the accessible text",
    "Donor-constant; memory tracks the per-year MEM_E rate; executive/language shrunk toward the mean (table below)",
    "Model, time scale, the meaning of 'z' and '0' in the names, and uncertainty",
  "`slope_zcasi_irt0`",
    "None located",
    "Correlates 0.91 with the memory slope; no visit-level CASI-IRT released",
    "Model, units, uncertainty",
  "`CASI_bifactor_scores`",
    "–", "Absent from every local file", "Whether it exists in another release"
)
show_table(prov, caption = "Cognitive-variable provenance: what is documented, what is observed in these data, and what remains open. Slope definitions are not inferred from column names.")

se_tbl <- map_dfr(DOMAINS$visit, function(v) {
  se <- act_visits[[paste0(v, "_SE")]]; se <- se[!is.na(se)]
  tibble(Score = v, `Visits with SE` = length(se), `Median SE` = median(se),
         `Share SE > 0.6` = percent(mean(se > 0.6), 1))
})
show_table(se_tbl, caption = "Visit-level score standard errors in ACT. The SE > 0.6 column applies the exclusion rule of the pooled genetic study only descriptively; no visit is excluded here.")
```

**Why do 58% of visuospatial visits share one value?** The score distribution suggests an answer,
though no document states it:

```{r}
#| label: vsp-distribution

vsp <- act_visits %>% filter(!is.na(VSP_E)) %>% mutate(ybd = age - age_vis)
vsp_mode <- as.numeric(names(which.max(table(vsp$VSP_E))))
vsp_diag <- list(
  distinct = n_distinct(vsp$VSP_E), n = nrow(vsp),
  share_mode = mean(vsp$VSP_E == vsp_mode),
  se_at_mode = unique(vsp$VSP_E_SE[vsp$VSP_E == vsp_mode]),
  share_above = mean(vsp$VSP_E > vsp_mode),
  se_above = median(vsp$VSP_E_SE[vsp$VSP_E > vsp_mode]),
  max = max(vsp$VSP_E),
  donors_at_mode = n_distinct(vsp$donor_id[vsp$VSP_E == vsp_mode]),
  mem_distinct = n_distinct(act_visits$MEM_E, na.rm = TRUE)
)
stopifnot("every visit at the modal VSP_E value carries one identical SE" = length(vsp_diag$se_at_mode) == 1)
vsp %>% mutate(`Years before death` = cut(ybd, c(-1, 3, 7, 12, 30), labels = c("0–3", "4–7", "8–12", "> 12"))) %>%
  group_by(`Years before death`) %>%
  summarise(Visits = n(), `At modal value` = percent(mean(VSP_E == vsp_mode), 1),
            `Above modal value` = percent(mean(VSP_E > vsp_mode), 1), .groups = "drop") %>%
  show_table(caption = sprintf("VSP_E in ACT: share of visits at and above the modal value (%.3f) by time before death.", vsp_mode))
```

- **Few distinct values.** ACT `VSP_E` takes `r vsp_diag$distinct` distinct values over
  `r vsp_diag$n` visits, against `r vsp_diag$mem_distinct` for `MEM_E`.
- **An identical scored pattern.** The modal value `r sprintf("%.3f", vsp_mode)`
  holds `r percent(vsp_diag$share_mode, 1)` of visits and `r vsp_diag$donors_at_mode` of 69 donors
  reach it at least once. Every modal visit carries the same SE
  (`r sprintf("%.3f", vsp_diag$se_at_mode)`), which is what an identical item set with an
  identical response pattern produces.
- **Not the scale maximum.** Only `r percent(vsp_diag$share_above, 1)` of visits score above it
  (up to `r sprintf("%.2f", vsp_diag$max)`), and those carry lower SEs (median
  `r sprintf("%.2f", vsp_diag$se_above)`), as expected when more items were administered.
- **Empirical reading.** The pattern is **consistent with a ceiling on a small, routinely
  administered visuospatial item set**, with higher scores reachable only at visits with
  additional items. The share at the mode also varies non-monotonically with time before death,
  which suggests battery composition changed across ACT waves.
- **Status.** Neither the item content nor the battery history is documented in the accessible
  sources. This remains an empirical explanation, not a documented one. `VSP_E` and `slope_zvsp0`
  are kept as released, not rescaled or replaced.

**Released slopes against the visits.** Each released slope is compared with the ordinary
least-squares rate of the same donor's visits. That rate is a check only; it is never used as an
outcome.

```{r}
#| label: slope-verification

slope_check <- pmap_dfr(DOMAINS, function(domain, key, visit, slope) {
  d <- act_slopes %>% select(donor_id, released = all_of(slope)) %>%
    inner_join(dom_sum[[key]] %>% select(donor_id, ols_rate, n_visits), by = "donor_id") %>%
    filter(!is.na(ols_rate))
  fit <- lm(released ~ ols_rate, d)
  tibble(Domain = domain, Variable = slope, Donors = nrow(d),
         `r with OLS rate` = cor(d$released, d$ols_rate),
         `Slope on OLS rate` = coef(fit)[2],
         `Mean offset` = mean(d$released - d$ols_rate),
         `SD ratio` = sd(d$released) / sd(d$ols_rate))
})
show_table(slope_check, caption = "Released domain slopes against each donor's own OLS rate (score per year of age). No standard error is released for any slope.")
sc <- function(dom, col) slope_check[[col]][slope_check$Domain == dom]
stopifnot(
  "memory slope sits on the per-year MEM_E scale" = abs(sc("Memory", "Slope on OLS rate") - 1) < 0.15,
  "executive slope is shrunk relative to OLS" = sc("Executive", "SD ratio") < 0.8,
  "visuospatial agreement is poor" = sc("Visuospatial", "r with OLS rate") < 0.5
)
```

- **`slope_zmem0` (memory decline).** Behaves like a per-year `MEM_E` rate (coefficient
  `r sprintf("%.2f", sc("Memory", "Slope on OLS rate"))` on the donors' OLS rates, r
  `r sprintf("%.2f", sc("Memory", "r with OLS rate"))`), with a mean offset.
- **`slope_zexf0` and `slope_zlan0` (executive and language decline).** Shrunk toward the mean
  (SD ratios `r sprintf("%.2f", sc("Executive", "SD ratio"))` and
  `r sprintf("%.2f", sc("Language", "SD ratio"))`), consistent with model-based person-specific
  estimates.
- **`slope_zvsp0` (visuospatial decline).** Agrees poorly with the visits (r
  `r sprintf("%.2f", sc("Visuospatial", "r with OLS rate"))`).
- **`slope_zcasi_irt0` (global decline).** An IRT-scaled CASI slope that cannot be checked against
  visits, since no visit-level score is released. It is the only global longitudinal rate.

The exact models, units and uncertainty of all five slopes remain **unresolved**. They are used as
released, never replaced by the OLS rates, and never given assumed standard errors. Coefficients
on slopes are therefore reported in SD of the released variable, not in cognitive units per year.

### 4.2 Four-domain characterization: level and rate

**Level** is the mean domain score over visits within 10 years of death (≥ 2 visits). The last
visit is also shown, and the released slope is the **rate**.

```{r}
#| label: domain-characterization

vis_long <- map_dfr(DOMAINS$visit, ~ act_visits %>%
  transmute(donor_id, visit = .x, score = .data[[.x]], se = .data[[paste0(.x, "_SE")]]))
dom_tbl <- pmap_dfr(DOMAINS, function(domain, key, visit, slope) {
  v <- vis_long %>% filter(visit == !!visit, !is.na(score))
  modal <- as.numeric(names(which.max(table(v$score))))
  lvl <- act[[paste0(key, "_level")]]; rate <- act[[slope]]
  tibble(Domain = domain,
         `Donors (visits, median [IQR])` = sprintf("%d (%s)", n_distinct(v$donor_id),
                                                   qr(as.numeric(table(v$donor_id)))),
         `Visit SE / visit SD` = sprintf("%.2f / %.2f", median(v$se, na.rm = TRUE), sd(v$score)),
         `Visits at the modal value` = percent(mean(v$score == modal), 1),
         `Last visit, median [IQR]` = qr(act[[paste0(key, "_last")]], 2),
         `10-y level, median [IQR]` = qr(lvl, 2),
         `Released slope, median [IQR]` = qr(rate, 3),
         `ρ level vs rate` = scor(lvl, rate),
         `ρ rate vs last CASI` = scor(rate, act$casi))
})
show_table(dom_tbl, caption = "Four-domain characterization, 69 ACT donors. Visit SE and SD from all ACT visits. A large share of visits at one value indicates a ceiling or coarse scale.")

dm <- function(d, col) dom_tbl[[col]][dom_tbl$Domain == d]
vsp_modal <- dm("Visuospatial", "Visits at the modal value")
```

**Reading.**

- **Memory and language** both decline measurably, and in both the rate correlates with terminal
  CASI (ρ `r sprintf("%.2f", dm("Memory", "ρ rate vs last CASI"))` and
  `r sprintf("%.2f", dm("Language", "ρ rate vs last CASI"))`).
- **Executive function** declines more slowly. Its rate is related to terminal CASI more weakly
  (`r sprintf("%.2f", dm("Executive", "ρ rate vs last CASI"))`).
- **Visuospatial scores sit largely on one value** (`r vsp_modal` of visits). The domain has a
  coarse ceiling, and its rate is essentially unrelated to terminal CASI
  (`r sprintf("%.2f", dm("Visuospatial", "ρ rate vs last CASI"))`). Memory and language also show
  ceiling clusters (`r dm("Memory", "Visits at the modal value")` and
  `r dm("Language", "Visits at the modal value")` of visits at one value), so early decline from
  high performers is poorly resolved.
- **Level and rate** are moderately correlated within each domain, but they are not
  interchangeable.
- **A single visit is noisy.** Visit-level standard errors are comparable to the between-visit SD
  for memory and language, and larger than it for executive and visuospatial scores.

**Grouped-trajectory diagnostic (Figure 2C).** The binned trajectories pool whichever donors were
seen at each time before death. Two checks follow: how many donors feed each bin, and whether the
within-donor change over each donor's own final visits shows the same ordering.

```{r}
#| label: trajectory-diagnostic

traj_mem <- act_visits %>% filter(!is.na(MEM_E)) %>%
  inner_join(donor %>% select(donor_id, gp_tertile), by = "donor_id") %>%
  mutate(ybd = age - age_vis, bin = pmin(floor(ybd / 2) * 2 + 1, 19))
bin_counts <- traj_mem %>% filter(bin <= 11) %>%
  group_by(`Global pTau tertile` = gp_tertile, bin) %>%
  summarise(n = n_distinct(donor_id), .groups = "drop") %>%
  mutate(bin = sprintf("%d–%d y", bin - 1, bin + 1)) %>%
  pivot_wider(names_from = bin, values_from = n, values_fill = 0)
show_table(bin_counts, caption = "Donors contributing a memory visit to each 2-year bin before death (the bins of Figure 2C; bins with < 5 donors are not drawn).")

within_change <- traj_mem %>% group_by(donor_id, gp_tertile) %>% arrange(age_vis, .by_group = TRUE) %>%
  summarise(last_ybd = min(ybd),
            ref = which.min(abs((max(age_vis) - age_vis) - 4)),
            gap = max(age_vis) - age_vis[ref],
            change = last(MEM_E) - MEM_E[ref], .groups = "drop") %>%
  filter(gap >= 3)
within_tbl <- within_change %>% group_by(`Global pTau tertile` = gp_tertile) %>%
  summarise(Donors = n(), `Last visit, years before death (median)` = median(last_ybd),
            `Within-donor MEM_E change per year, mean [IQR]` =
              sprintf("%.3f [%.3f, %.3f]", mean(change / gap), quantile(change / gap, .25), quantile(change / gap, .75)),
            .groups = "drop")
show_table(within_tbl, caption = "Within-donor memory change between each donor's last visit and the visit closest to 4 years earlier (gap ≥ 3 years). Descriptive; not a trajectory model.")
wc <- function(t) mean(within_change$change[within_change$gp_tertile == t] / within_change$gap[within_change$gp_tertile == t])
bal_late <- traj_mem %>% filter(bin %in% c(1, 3, 5, 7)) %>% distinct(donor_id, bin) %>% count(donor_id) %>% filter(n == 4) %>% nrow()
stopifnot("within-donor late memory change is steepest in the high tertile" =
            wc(GT_LEVELS[3]) < wc(GT_LEVELS[2]) && wc(GT_LEVELS[2]) < wc(GT_LEVELS[1]))
```

**Reading of the diagnostic.**

- **Thin final bins.** The last 0–2 years before death hold at most
  `r max(bin_counts[["0–2 y"]])` donors per tertile and are not drawn. The 2–4-year bin holds
  `r min(bin_counts[["2–4 y"]])`–`r max(bin_counts[["2–4 y"]])`.
- **Different donors in different bins.** Only `r bal_late` donors contribute to every bin from 0
  to 8 years, so a fixed-donor version of the curves cannot be drawn.
- **The ordering is within-donor, not just composition.** Over each donor's own final ~4 years,
  memory declines faster in the higher tertiles
  (`r sprintf("%.3f", wc(GT_LEVELS[1]))`, `r sprintf("%.3f", wc(GT_LEVELS[2]))` and
  `r sprintf("%.3f", wc(GT_LEVELS[3]))` per year, low to high). The separation of the grouped
  curves reflects within-donor decline, not only who was seen when.
- **The late steepening is not an individual transition.** The apparent late steepening of the
  high-tertile curve mixes donors with different visit timing, and high-tertile donors' last
  visits fall earlier. It is not read as an individual-level acceleration or change point.

### 4.3 Cognitive phenotypes and global pTau progression

Each phenotype (standardized) is modelled on `CPS_Global_pTau` (per 0.1), age, sex, education and
assessment timing:

- the CASI-to-death interval for terminal CASI;
- years from the last harmonized visit to death for the levels and slopes.

That makes ten single-exposure models. The **co-pathology sensitivity** adds LATE-NC 2–3 and
severe arteriolosclerosis on identical donors. Braak and ADNC are shown descriptively only and
never enter these models alongside CPS.

```{r}
#| label: phenotype-global

PHENO <- tibble(
  phenotype = c("Terminal CASI", "Global slope (CASI-IRT)",
                paste(DOMAINS$domain, "rate"), paste(DOMAINS$domain, "level")),
  var = c("casi", SLOPE_GLOBAL, DOMAINS$slope, paste0(DOMAINS$key, "_level")),
  kind = c("Terminal", "Rate", rep("Rate", 4), rep("Level", 4)),
  domain = c("Global", "Global", DOMAINS$domain, DOMAINS$domain),
  timing = c("casi_interval", rep("last_ybd", 9))
)
fit_pheno <- function(var, timing, extra = NULL) {
  d <- act %>% filter(!is.na(.data[[var]])) %>% mutate(y = zstd(.data[[var]]))
  base <- lm(as.formula(paste("y ~ gp10 + age + sex + education +", timing)), d)
  full <- if (length(extra)) update(base, as.formula(paste(". ~ . +", paste(extra, collapse = " + ")))) else base
  stopifnot("co-pathology model on identical donors" = nobs(full) == nobs(base))
  ci <- confint(full)["gp10", ]
  tibble(estimate = coef(full)["gp10"], lo = ci[1], hi = ci[2], n = nobs(full), r2 = summary(full)$r.squared)
}
stopifnot("no LATE-unclassifiable ACT donor" = all(act$late != "Unclassifiable"))
pheno_fits <- PHENO %>%
  mutate(primary = map2(var, timing, ~ fit_pheno(.x, .y)),
         copath = map2(var, timing, ~ fit_pheno(.x, .y, c("late_2_3", "arteriolo_severe")))) %>%
  pivot_longer(c(primary, copath), names_to = "variant", values_to = "res") %>% unnest(res) %>%
  mutate(variant = recode(variant, primary = "Primary", copath = "+ LATE-NC 2–3 + severe arteriolosclerosis"))
pf <- function(p, v = "Primary") pheno_fits %>% filter(phenotype == p, variant == v)
pf_ci <- function(p, v = "Primary") { x <- pf(p, v); fmt_sci(x$estimate, x$lo, x$hi) }
stopifnot(
  "global pTau tracks terminal CASI, the global slope, memory and language rates" =
    all(map_lgl(c("Terminal CASI", "Global slope (CASI-IRT)", "Memory rate", "Language rate"), ~ pf(.x)$hi < 0)),
  "global pTau does not clearly track executive or visuospatial rate" =
    pf("Executive rate")$hi > 0 && pf("Visuospatial rate")$hi > 0
)

pheno_fits %>%
  mutate(cell = fmt_sci(estimate, lo, hi)) %>%
  select(Phenotype = phenotype, Kind = kind, variant, cell, n, r2) %>%
  pivot_wider(names_from = variant, values_from = c(cell, r2), names_glue = "{variant} {.value}") %>%
  select(Phenotype, Kind, Donors = n, `Primary cell`, `Primary r2`,
         `+ LATE-NC 2–3 + severe arteriolosclerosis cell`) %>%
  rename(`SD per 0.1 Global pTau [95% CI]` = `Primary cell`, `Model R²` = `Primary r2`,
         `+ co-pathology` = `+ LATE-NC 2–3 + severe arteriolosclerosis cell`) %>%
  show_table(caption = "Each phenotype (SD units) on CPS_Global_pTau per 0.1, adjusted for age, sex, education and assessment timing; OLS 95% CI; one row per ACT donor. Levels need ≥ 2 visits within 10 years of death.")

stage_cog <- act %>%
  group_by(`Braak stage` = braak_group) %>%
  summarise(Donors = n(), `Last CASI` = qr(casi),
            `Memory rate` = qr(slope_zmem0, 3), `Executive rate` = qr(slope_zexf0, 3),
            `Language rate` = qr(slope_zlan0, 3), `Visuospatial rate` = qr(slope_zvsp0, 3),
            .groups = "drop")
show_table(stage_cog, caption = "Descriptive context: cognitive phenotypes (median [IQR]) by Braak stage in ACT. No model.")
```

**Progression shape.** The linear representation was checked for the two prioritized outcomes,
terminal CASI and memory rate. Mean residuals were examined by quintile of global pTau, and one
nonlinear sensitivity was fitted (a quadratic term), with no spline or threshold search.

```{r}
#| label: progression-shape

shape <- map_dfr(c(`Terminal CASI` = "casi", `Memory rate` = "slope_zmem0"), function(y) {
  timing <- if (y == "casi") "casi_interval" else "last_ybd"
  d <- act %>% mutate(yz = zstd(.data[[y]]))
  f1 <- lm(as.formula(paste("yz ~ CPS_Global_pTau + age + sex + education +", timing)), d)
  f2 <- update(f1, . ~ . + I(CPS_Global_pTau^2))
  stopifnot("linear and quadratic fits on identical donors" = nobs(f1) == nobs(f2))
  q <- cut(d$CPS_Global_pTau, quantile(d$CPS_Global_pTau, 0:5 / 5), include.lowest = TRUE, labels = paste0("Q", 1:5))
  as_tibble(as.list(round(tapply(residuals(f1), q, mean), 2))) %>%
    mutate(dr2 = summary(f2)$r.squared - summary(f1)$r.squared,
           p = anova(f1, f2)$`Pr(>F)`[2], quad = coef(f2)["I(CPS_Global_pTau^2)"])
}, .id = "Outcome")
shape %>% transmute(Outcome, Q1, Q2, Q3, Q4, Q5,
                    `ΔR² quadratic` = sprintf("%.3f", dr2), `F-test p` = sprintf("%.3f", p)) %>%
  show_table(caption = "Mean residual (SD units) of the primary linear model by quintile of CPS_Global_pTau (Q1 lowest), and the single quadratic sensitivity. ACT, 69 donors.")
shp <- function(o, col) shape[[col]][shape$Outcome == o]
n_ceiling <- sum(act$casi >= 95)
stopifnot("terminal CASI shows concave curvature (negative quadratic)" = shp("Terminal CASI", "quad") < 0)
```

- **Terminal CASI departs from linearity.** The mid-stage residuals are positive and those at both
  extremes negative. A quadratic term adds ΔR² `r sprintf("%.3f", shp("Terminal CASI", "dr2"))`
  (p `r sprintf("%.3f", shp("Terminal CASI", "p"))`). CASI is flat at low global stage, where
  `r n_ceiling` donors score ≥ 95, and falls more steeply at high stage.
- **Memory rate shows the same shape, more weakly** (ΔR² `r sprintf("%.3f", shp("Memory rate", "dr2"))`,
  p `r sprintf("%.3f", shp("Memory rate", "p"))`).
- **What this means for the linear model.** The linear coefficient is best read as the average
  association over the observed range. It remains the primary representation, and the direction
  and ranking of domains below are unaffected.
- **Where it matters.** The curvature bears on Section 5, since any feature concentrated at high
  stage could absorb it, so that section adds the quadratic as a sensitivity.

**Reading.** Global pTau progression separates the domains:

- **Terminal CASI:** `r pf_ci("Terminal CASI")` SD per 0.1, model R²
  `r sprintf("%.2f", pf("Terminal CASI")$r2)`.
- **Global slope:** `r pf_ci("Global slope (CASI-IRT)")`.
- **Memory rate:** `r pf_ci("Memory rate")`.
- **Language rate:** `r pf_ci("Language rate")`.
- **Executive rate:** `r pf_ci("Executive rate")`, weak.
- **Visuospatial rate:** unrelated (`r pf_ci("Visuospatial rate")`).

Rates track global pTau more closely than levels do (memory level `r pf_ci("Memory level")`). All
estimates hold with co-pathology added.

The Braak table shows the same ordering descriptively, with wide within-stage spread in every
rate.

### 4.4 Global versus anatomically matched local pTau

Only one anatomically motivated region per domain is used (not a region × domain grid):

| Domain | Region | Rationale |
|---|---|---|
| Memory | MEC | entorhinal input to the hippocampal memory system |
| Executive | DFC | dorsolateral prefrontal cortex |
| Language | MTG | lateral temporal lexical-semantic cortex |
| Visuospatial | AnG | inferior parietal cortex (profiled only in the AD-spectrum donors) |

For each domain rate, two models are fitted on **identical donors**, one with global and one with
matched local pTau, each alone with the same covariates. They are compared by standardized
coefficient and ΔR² (local − global) with a donor bootstrap. The two are never entered together.

```{r}
#| label: global-vs-local

MATCH <- tibble(domain = DOMAINS$domain, slope = DOMAINS$slope, region = c("MEC", "DFC", "MTG", "AnG"))
gl_fun <- function(slope, region) function(d) {
  d <- d %>% mutate(y = zstd(.data[[slope]]), g = zstd(CPS_Global_pTau), l = zstd(.data[[paste0("lp_", region)]]))
  fg <- lm(y ~ g + age + sex + education + last_ybd, d)
  fl <- lm(y ~ l + age + sex + education + last_ybd, d)
  c(global = unname(coef(fg)["g"]), local = unname(coef(fl)["l"]),
    dr2 = summary(fl)$r.squared - summary(fg)$r.squared)
}
gl <- pmap_dfr(MATCH, function(domain, slope, region) {
  d <- act %>% filter(!is.na(.data[[paste0("lp_", region)]]))
  cluster_bootstrap(d, gl_fun(slope, region), B_BOOT, seed = BOOT_SEED) %>%
    mutate(domain = domain, region = region, n = nrow(d))
})
stopifnot("every bootstrap resample fitted" = all(gl$n_boot_ok == B_BOOT))
glv <- function(dom, t) gl %>% filter(domain == dom, term == t)
gl_ci <- function(dom, t, d = 2) { x <- glv(dom, t); fmt_sci(x$estimate, x$lo, x$hi, d) }

gl %>% mutate(cell = fmt_sci(estimate, lo, hi)) %>%
  select(Domain = domain, Region = region, Donors = n, term, cell) %>%
  pivot_wider(names_from = term, values_from = cell) %>%
  rename(`Global pTau (SD/SD)` = global, `Matched local pTau (SD/SD)` = local, `ΔR² local − global` = dr2) %>%
  show_table(caption = sprintf("Domain rate on global or matched local pTau CPS, separately, on identical donors (age, sex, education, last-visit timing); donor bootstrap 95%% CI (%d resamples).", B_BOOT))
stopifnot(
  "MEC local pTau does not out-predict global pTau for memory" = glv("Memory", "dr2")$estimate < 0,
  "DFC local pTau is at least as predictive as global for executive rate" = glv("Executive", "dr2")$estimate > 0
)
```

**Reading.** Global and local progression relate differently to the domains:

- **Memory decline** follows **global** pTau (`r gl_ci("Memory", "global")` SD/SD) at least as
  closely as MEC local pTau (`r gl_ci("Memory", "local")`; ΔR² `r gl_ci("Memory", "dr2", 3)`).
  MEC is heavily involved early in almost every donor (Section 2), so its variation between
  donors is compressed. Memory decline reads as a stage-wide phenotype.
- **Executive decline** leans toward **DFC** (`r gl_ci("Executive", "local")`) over global
  (`r gl_ci("Executive", "global")`; ΔR² `r gl_ci("Executive", "dr2", 3)`). This is the one
  domain where the matched region edges out global stage, as the first pass found with direct
  AT8, but the difference is imprecise.
- **Language** follows global (`r gl_ci("Language", "global")`) at least as closely as MTG
  (`r gl_ci("Language", "local")`).
- **Visuospatial** is unrelated to either (AnG, n = `r glv("Visuospatial", "global")$n`), which is
  unsurprising given its ceiling and poorly defined slope.

None of these comparisons shows that a region's pathological distribution measures circuit
function. They show which progression representation carries the domain's association.

```{r}
#| label: fig2-build

# --- A. measurement timing ---------------------------------------------------------------
timing_d <- act_visits %>% filter(!is.na(MEM_E)) %>% mutate(ybd = age - age_vis)
donor_order <- act %>% arrange(desc(last_ybd), desc(first_ybd)) %>% pull(donor_id)
timing_d <- timing_d %>% mutate(row = match(donor_id, donor_order))
casi_t <- act %>% transmute(donor_id, row = match(donor_id, donor_order), ybd = casi_interval / 12)
f2a <- ggplot(timing_d, aes(ybd, row)) +
  geom_segment(data = timing_d %>% group_by(row) %>% summarise(a = min(ybd), b = max(ybd)),
               aes(x = a, xend = b, y = row, yend = row), inherit.aes = FALSE,
               colour = "#d6d5d0", linewidth = 0.3) +
  geom_point(size = 0.5, colour = INK2) +
  geom_point(data = casi_t, aes(shape = "Last CASI"), size = 1.0, colour = INK, stroke = 0.4) +
  scale_shape_manual(values = c(`Last CASI` = 4), name = NULL) +
  scale_x_reverse(breaks = seq(0, 25, 5)) +
  scale_y_continuous(expand = expansion(add = 1), breaks = NULL) +
  labs(title = "When cognition was measured",
       subtitle = sprintf("%d ACT donors; dots = harmonized visits", nrow(act)),
       x = "Years before death", y = "Donors (by last visit)") +
  theme(panel.grid.major.y = element_blank(), axis.line.y = element_blank(), legend.position = "bottom")

# --- B. correlation among phenotypes -----------------------------------------------------------
COR_VARS <- c(`Terminal CASI` = "casi", `Global slope` = SLOPE_GLOBAL,
              setNames(DOMAINS$slope, paste(DOMAINS$domain, "rate")),
              setNames(paste0(DOMAINS$key, "_level"), paste(DOMAINS$domain, "level")),
              `Global pTau` = "CPS_Global_pTau")
cm <- cor(act[COR_VARS], method = "spearman", use = "pairwise.complete.obs")
dimnames(cm) <- list(names(COR_VARS), names(COR_VARS))
cm_long <- as_tibble(cm, rownames = "a") %>% pivot_longer(-a, names_to = "b", values_to = "rho") %>%
  mutate(ia = match(a, names(COR_VARS)), ib = match(b, names(COR_VARS))) %>% filter(ia > ib) %>%
  mutate(a = factor(a, levels = rev(names(COR_VARS))), b = factor(b, levels = names(COR_VARS)))
f2b <- ggplot(cm_long, aes(b, a, fill = rho)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%.2f", rho), colour = abs(rho) > 0.55), size = 1.9, family = FIG_FONT) +
  scale_fill_gradient2(low = "#c2410c", mid = "#f4f3ef", high = "#1c5cab", midpoint = 0, limits = c(-1, 1),
                       name = "Spearman ρ") +
  scale_colour_manual(values = c(`FALSE` = INK, `TRUE` = "white"), guide = "none") +
  labs(title = "Level and rate are distinct", x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1), axis.line = element_blank(),
        axis.ticks = element_blank(), panel.grid.major.y = element_blank(),
        legend.position = "bottom", legend.key.width = unit(7, "mm"), legend.key.height = unit(2.5, "mm"))

# --- C. four-domain trajectories by Global pTau tertile -----------------------------------------
traj <- act_visits %>%
  inner_join(donor %>% select(donor_id, gp_tertile), by = "donor_id") %>%
  mutate(ybd = age - age_vis) %>%
  select(donor_id, gp_tertile, ybd, all_of(DOMAINS$visit)) %>%
  pivot_longer(all_of(DOMAINS$visit), names_to = "visit", values_to = "score") %>%
  filter(!is.na(score)) %>%
  mutate(domain = factor(DOMAINS$domain[match(visit, DOMAINS$visit)], levels = DOMAINS$domain))
traj_bins <- traj %>%
  mutate(bin = pmin(floor(ybd / 2) * 2 + 1, 19)) %>%
  group_by(domain, gp_tertile, bin) %>%
  summarise(n = n_distinct(donor_id), mean = mean(score), se = sd(score) / sqrt(n), .groups = "drop") %>%
  filter(n >= 5)
dodge <- position_dodge(width = 0.8)
f2c <- ggplot(traj, aes(ybd, score)) +
  geom_line(aes(group = donor_id), colour = "#e6e5e1", linewidth = 0.18) +
  geom_linerange(data = traj_bins, aes(bin, ymin = mean - se, ymax = mean + se, colour = gp_tertile),
                 inherit.aes = FALSE, linewidth = 0.4, position = dodge) +
  geom_line(data = traj_bins, aes(bin, mean, colour = gp_tertile, linetype = gp_tertile), linewidth = 0.55,
            position = dodge) +
  geom_point(data = traj_bins, aes(bin, mean, fill = gp_tertile, shape = gp_tertile), colour = INK,
             size = 1.2, stroke = 0.3, position = dodge) +
  scale_colour_manual(values = PAL_GT, name = NULL) + scale_fill_manual(values = PAL_GT, name = NULL) +
  scale_shape_manual(values = c(21, 22, 24), name = NULL) +
  scale_linetype_manual(values = c("13", "22", "solid"), name = NULL) +
  scale_x_reverse(breaks = seq(0, 20, 5), limits = c(21, 0)) +
  coord_cartesian(ylim = c(-2, 2)) +
  facet_wrap(~ domain, nrow = 1) +
  labs(title = "Four-domain trajectories by Global pTau tertile",
       x = "Years before death", y = "Harmonized domain score") +
  theme(legend.position = "bottom", legend.key.width = unit(6, "mm"), panel.spacing.x = unit(3, "mm"))

# --- D. phenotype associations with global pTau ------------------------------------------------
f2d_data <- pheno_fits %>% filter(variant == "Primary") %>%
  mutate(term = factor(recode(domain, Global = "Global cognition"), levels = c("Global cognition", DOMAINS$domain)),
         variant = factor(kind, levels = c("Terminal", "Rate", "Level")))
f2d <- forest_panel(f2d_data %>% mutate(variant = as.character(variant)),
                    "SD per 0.1 CPS_Global_pTau (95% CI)", "Phenotype vs global progression",
                    variants = c("Terminal", "Rate", "Level")) +
  scale_shape_manual(values = c(Terminal = 23, Rate = 21, Level = 1), name = NULL) +
  guides(shape = guide_legend(nrow = 1)) + theme(legend.direction = "horizontal")

# --- E. global vs matched local -------------------------------------------------------------------
f2e_data <- gl %>% filter(term %in% c("global", "local")) %>%
  mutate(variant = recode(term, global = "Global pTau CPS", local = "Matched local pTau CPS"),
         term = factor(sprintf("%s rate\n(local: %s)", domain, region),
                       levels = sprintf("%s rate\n(local: %s)", MATCH$domain, MATCH$region)))
f2e <- forest_panel(f2e_data, "SD of rate per SD of progression (95% CI)", "Global vs matched local",
                    variants = c("Global pTau CPS", "Matched local pTau CPS")) +
  guides(shape = guide_legend(nrow = 1)) + theme(legend.direction = "horizontal")

fig2 <- plot_grid(
  compose(f2a, f2b, labels = c("A", "B"), align = "h", axis = "b", rel_widths = c(0.8, 1.2)),
  compose(f2c, labels = "C"),
  compose(f2d, f2e, labels = c("D", "E"), align = "h", axis = "b"),
  ncol = 1, rel_heights = c(1.1, 0.75, 1)
)
fig2_png <- save_figure(fig2, "fig2_longitudinal_phenotypes", height_mm = 230)
fig2_cap <- sprintf(paste0(
  "**Figure 2.** Longitudinal cognitive phenotypes in the %d ACT donors (all with ≥ 3 harmonized visits and all five released slopes; ADRC excluded). ",
  "**A** Timing of every harmonized visit (dots) and the last CASI (×) relative to death; whole-year ages add ±1 year of rounding. ",
  "**B** Spearman correlations among terminal CASI, the released global and domain slopes (rate), 10-year domain levels, and CPS_Global_pTau. ",
  "**C** Memory, executive, language and visuospatial trajectories; grey lines are donors, coloured lines 2-year-bin means ± SE within CPS_Global_pTau tertiles (all-donor cuts), shown where a bin holds ≥ 5 donors, so the final 0–2 years are not drawn; bins pool different donors and are not individual curves, and late steepening is not an individual transition (within-donor diagnostic in Section 4.2). Visuospatial scores sit mostly at one value (Section 4.1). ",
  "**D** Each phenotype (SD) on CPS_Global_pTau per 0.1, adjusted for age, sex, education and assessment timing; OLS 95%% CI. ",
  "**E** Each domain rate on global pTau or on one anatomically matched local pTau (memory–MEC, executive–DFC, language–MTG, visuospatial–AnG), fitted separately on identical donors (AnG n = %d); donor bootstrap 95%% CI. ",
  "Released slopes have no codebook or standard errors and behave like model-based estimates (Section 4.1). Associations are cross-sectional in tissue and not causal."),
  nrow(act), glv("Visuospatial", "global")$n)
```

```{r}
#| label: fig2
#| fig-cap: !expr fig2_cap
#| out-width: "100%"

include_graphics(fig2_png)
```

---

## 5. Heterogeneity at comparable global progression

This section asks whether donors at the same `CPS_Global_pTau` but with different cognition
differ in regional involvement or tissue response. The design is continuous and at donor level,
with no resilience threshold and no matched cutoffs.

**Model.** Each prespecified feature, standardized and one at a time, is added to

`phenotype ~ CPS_Global_pTau + age + sex + education + timing + feature`

**Outcomes:** terminal CASI and memory and executive rates, in SD units.

**Features:**

- local pTau in MEC and in DFC, the memory- and executive-matched regions;
- donor-wide direct AT8;
- donor-wide MG4;
- donor-wide NeuN;
- donor-wide astrocyte activation.

A local-pTau coefficient reads as *involvement of that region beyond the donor's global pTau
stage*. It does not treat local as a residual or as an independent exposure; its collinearity with
global progression is tabulated below. Sensitivity: add LATE-NC 2–3 and severe arteriolosclerosis
on identical donors.

```{r}
#| label: heterogeneity

feat_donor <- tissue %>% distinct(donor_id, at8_donor, neun_donor) %>%
  left_join(bal %>% distinct(donor_id, mg4_donor, astro_donor), by = "donor_id")
act_f <- act %>% left_join(feat_donor, by = "donor_id")
FEATURES <- c(lp_MEC = "Local pTau CPS, MEC", lp_DFC = "Local pTau CPS, DFC",
              at8_donor = "Donor-wide direct AT8", mg4_donor = "Donor-wide MG4",
              neun_donor = "Donor-wide NeuN", astro_donor = "Donor-wide astrocyte activation")
H_OUT <- tibble(outcome = c("Terminal CASI", "Memory rate", "Executive rate"),
                var = c("casi", "slope_zmem0", "slope_zexf0"),
                timing = c("casi_interval", "last_ybd", "last_ybd"))
fit_h <- function(feature, var, timing, extra = NULL, drop = character(0)) {
  d <- act_f %>% filter(!is.na(.data[[feature]]), !donor_id %in% drop) %>%
    mutate(x = zstd(.data[[feature]]), y = zstd(.data[[var]]))
  rhs <- paste("CPS_Global_pTau + age + sex + education +", timing,
               if (length(extra)) paste("+", paste(extra, collapse = " + ")))
  base <- lm(as.formula(paste("y ~", rhs)), d)
  full <- update(base, . ~ . + x)
  stopifnot("nested models use identical donors" = nobs(base) == nobs(full), nobs(full) == nrow(d))
  ci <- confint(full)["x", ]
  tibble(estimate = coef(full)["x"], lo = ci[1], hi = ci[2], n = nobs(full),
         d_r2 = summary(full)$r.squared - summary(base)$r.squared)
}
h_fits <- expand_grid(feature = names(FEATURES), H_OUT) %>%
  mutate(primary = pmap(list(feature, var, timing), fit_h),
         copath = pmap(list(feature, var, timing), ~ fit_h(..1, ..2, ..3, c("late_2_3", "arteriolo_severe"))),
         nofloor = pmap(list(feature, var, timing), ~ fit_h(..1, ..2, ..3, drop = NEUN_FLOOR_DONORS)),
         quad = pmap(list(feature, var, timing), ~ fit_h(..1, ..2, ..3, "I(CPS_Global_pTau^2)"))) %>%
  pivot_longer(c(primary, copath, nofloor, quad), names_to = "variant", values_to = "res") %>% unnest(res) %>%
  mutate(variant = recode(variant, primary = "Primary", copath = "+ LATE-NC 2–3 + severe arteriolosclerosis",
                          nofloor = "Without the 2 NeuN-floor donors", quad = "+ quadratic global pTau"),
         term = factor(FEATURES[feature], levels = FEATURES))
hv <- function(f, o, v = "Primary") h_fits %>% filter(feature == f, outcome == o, variant == v)
hv_ci <- function(f, o, v = "Primary") { x <- hv(f, o, v); fmt_sci(x$estimate, x$lo, x$hi) }
h_collin <- tibble(Feature = FEATURES,
                   `ACT donors` = map_int(names(FEATURES), ~ sum(!is.na(act_f[[.x]]))),
                   `ρ with CPS_Global_pTau` = map_dbl(names(FEATURES), ~ scor(act_f[[.x]], act_f$CPS_Global_pTau)))
stopifnot(
  "DFC involvement beyond global stage goes with lower terminal CASI" = hv("lp_DFC", "Terminal CASI")$hi < 0,
  "MEC involvement beyond global stage does not" = hv("lp_MEC", "Terminal CASI")$lo < 0,
  "donor-wide MG4 adds no clear terminal-CASI association beyond global pTau" =
    hv("mg4_donor", "Terminal CASI")$hi > 0,
  "donor-wide NeuN: positive for terminal CASI and memory rate" =
    hv("neun_donor", "Terminal CASI")$estimate > 0 && hv("neun_donor", "Memory rate")$lo > 0,
  "the NeuN memory-rate association survives removing the NeuN-floor donors" =
    hv("neun_donor", "Memory rate", "Without the 2 NeuN-floor donors")$lo > 0,
  "the DFC - terminal CASI interval includes zero once global pTau enters quadratically" =
    hv("lp_DFC", "Terminal CASI", "+ quadratic global pTau")$hi > 0,
  "the regional executive pattern holds under the quadratic" =
    hv("lp_DFC", "Executive rate", "+ quadratic global pTau")$hi < 0 &&
    hv("lp_MEC", "Executive rate", "+ quadratic global pTau")$lo > 0
)
# The NeuN association against the published severely-affected phenotype: an
# exclusion sensitivity on the same model, not an explanatory term.
neun_sa <- map_dfr(c(`Terminal CASI` = 1, `Memory rate` = 2), function(i) {
  fit_h("neun_donor", H_OUT$var[i], H_OUT$timing[i], drop = severe_ids)
}, .id = "outcome")
nsa_ci <- function(o) { x <- neun_sa[neun_sa$outcome == o, ]; fmt_sci(x$estimate, x$lo, x$hi) }
n_severe_act <- sum(act$donor_id %in% severe_ids)
stopifnot("without all severely affected donors, both NeuN intervals include zero" = all(neun_sa$lo < 0))

h_fits %>%
  mutate(cell = fmt_sci(estimate, lo, hi)) %>%
  select(Feature = term, variant, outcome, cell) %>%
  pivot_wider(names_from = outcome, values_from = cell) %>%
  arrange(Feature, variant) %>% rename(Model = variant) %>%
  show_table(caption = "Feature (per SD) added to phenotype ~ CPS_Global_pTau + age + sex + education + timing; outcomes in SD; OLS 95% CI; one row per ACT donor; nested models on identical donors within each variant. The NeuN-floor variant removes the two donors with near-zero NeuN immunoreactivity (Section 3.4). The quadratic variant allows curvature in global pTau (Section 4.3).")
show_table(h_collin, caption = "Collinearity of each feature with global pTau progression (ACT donors).")
```

**Reading.** At comparable global pTau progression:

- **DFC involvement and terminal CASI: not robust to curvature.** Donors with more DFC
  involvement have lower terminal CASI in the linear model (`r hv_ci("lp_DFC", "Terminal CASI")`
  SD per SD). Once global pTau enters quadratically it attenuates to
  `r hv_ci("lp_DFC", "Terminal CASI", "+ quadratic global pTau")`. DFC involvement rises mainly at
  high stage, where CASI falls fastest (Section 4.3), so this association is partly curvature in
  global progression, not regional involvement.
- **The executive pattern is regional and consistent in direction, but imprecise.** Its primary
  intervals just reach zero and clear it only once curvature is allowed. More DFC involvement goes
  with faster
  executive decline (`r hv_ci("lp_DFC", "Executive rate")`; with the quadratic
  `r hv_ci("lp_DFC", "Executive rate", "+ quadratic global pTau")`). Relatively MEC-predominant tau
  goes with slower executive decline (`r hv_ci("lp_MEC", "Executive rate")`; with the quadratic
  `r hv_ci("lp_MEC", "Executive rate", "+ quadratic global pTau")`). MEC involvement beyond global
  stage is unrelated to terminal CASI (`r hv_ci("lp_MEC", "Terminal CASI")`).
- **Consistency across representations.** This matches the first-pass MEC-vs-neocortex AT8
  gradient and the executive–DFC association of Section 4.4, on released rather than direct
  measures.
- **Direct donor-wide AT8** (`r hv_ci("at8_donor", "Terminal CASI")`) adds less beyond global pTau
  than it did beyond the Gabitto MTG CPS (Section 6). The released global pTau stage already
  captures most of the multiregional tau the first pass found missing from the MTG CPS.
- **Donor-wide NeuN: associated, but not separable from the severely affected phenotype.**
  - **Primary:** higher donor-wide NeuN goes with better terminal CASI
    (`r hv_ci("neun_donor", "Terminal CASI")`) and slower memory decline
    (`r hv_ci("neun_donor", "Memory rate")`). It is nearly independent of global pTau (ρ
    `r sprintf("%.2f", h_collin[[3]][h_collin$Feature == "Donor-wide NeuN"])`).
  - **Without the two NeuN-floor donors:** memory rate is unchanged
    (`r hv_ci("neun_donor", "Memory rate", "Without the 2 NeuN-floor donors")`).
  - **Without all `r n_severe_act` severely affected ACT donors:**
    `r nsa_ci("Terminal CASI")` and `r nsa_ci("Memory rate")`, both intervals including zero.
  - **With curvature in global pTau:**
    `r hv_ci("neun_donor", "Memory rate", "+ quadratic global pTau")`.

  The release describes these donors as having reduced NeuN labelling and steeper late memory
  decline (Section 3.4). This association therefore cannot be distinguished from that phenotype,
  and it is not carried forward as an independent tissue-integrity correlate of resilience.
- **Molecular response adds nothing clear.** Donor-wide MG4 (`r hv_ci("mg4_donor", "Terminal CASI")`)
  and astrocyte activation show no clear association beyond global stage.
- **Co-pathology** leaves the pattern intact.
- **Collinearity** is substantial (ρ with global pTau up to
  `r sprintf("%.2f", max(abs(h_collin[[3]])))`), so these are estimates of regional *pattern* at
  a given stage, with wide intervals.

```{r}
#| label: fig3-build

# --- A. cognition at comparable global progression ----------------------------------------
f3a <- ggplot(act, aes(CPS_Global_pTau, casi)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, colour = INK2, linewidth = 0.45, linetype = "22") +
  exec(geom_point, !!!point_args, mapping = aes(fill = adnc)) +
  adnc_fill() +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  scale_y_continuous(limits = c(60, 101), breaks = seq(60, 100, 10)) +
  labs(title = "Cognition vs global stage",
       x = "CPS_Global_pTau", y = "Last CASI (0–100)") +
  guides(fill = guide_legend(nrow = 2)) +
  theme(legend.position = "bottom")

# --- B. features beyond global progression -------------------------------------------------
f3b <- forest_panel(h_fits %>% mutate(outcome = factor(outcome, levels = H_OUT$outcome)),
                    "SD of phenotype per SD of feature, beyond CPS_Global_pTau (95% CI)",
                    "What differs at comparable global progression",
                    variants = c("Primary", "+ LATE-NC 2–3 + severe arteriolosclerosis",
                                 "Without the 2 NeuN-floor donors", "+ quadratic global pTau")) +
  facet_wrap(~ outcome, nrow = 1) +
  guides(shape = guide_legend(nrow = 2)) +
  theme(legend.direction = "horizontal", panel.spacing.x = unit(5, "mm"))

# --- C, D. added-variable plots for the two prespecified anatomical pairings --------------------
av_panel <- function(feature, var, timing, outcome_label, feature_label, outcome) {
  d <- act_f %>% filter(!is.na(.data[[feature]])) %>% mutate(x = zstd(.data[[feature]]), y = zstd(.data[[var]]))
  rhs <- paste("CPS_Global_pTau + age + sex + education +", timing)
  a <- tibble(x_res = residuals(lm(as.formula(paste("x ~", rhs)), d)),
              y_res = residuals(lm(as.formula(paste("y ~", rhs)), d)))
  e <- hv(feature, outcome)
  ggplot(a, aes(x_res, y_res)) +
    geom_hline(yintercept = 0, colour = MUTED, linewidth = 0.3) +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = INK2, linewidth = 0.45, linetype = "22") +
    exec(geom_point, !!!modifyList(point_args, list(size = 1.5)), fill = "#8a8984") +
    annotate("text", x = min(a$x_res), y = min(a$y_res), hjust = 0, vjust = 0, size = 2.3,
             family = FIG_FONT, colour = INK2,
             label = sprintf("n = %d ACT donors\n%s SD per SD", nrow(a), fmt_sci(e$estimate, e$lo, e$hi))) +
    labs(title = sprintf("%s vs %s", outcome_label, sub(" local pTau", "", feature_label)),
         x = sprintf("%s | global pTau, covariates", feature_label),
         y = sprintf("%s | global pTau, covariates (SD)", outcome_label))
}
f3c <- av_panel("lp_MEC", "slope_zmem0", "last_ybd", "Memory rate", "MEC local pTau", "Memory rate")
f3d <- av_panel("lp_DFC", "slope_zexf0", "last_ybd", "Executive rate", "DFC local pTau", "Executive rate")

fig3 <- plot_grid(
  compose(f3a, f3c, f3d, labels = c("A", "B", "C"), ncol = 3, align = "h", axis = "b"),
  compose(f3b, labels = "D"),
  ncol = 1, rel_heights = c(0.9, 1.15)
)
fig3_png <- save_figure(fig3, "fig3_heterogeneity_comparable_progression", height_mm = 175)
fig3_cap <- sprintf(paste0(
  "**Figure 3.** Regional involvement and tissue response among ACT donors at comparable global pTau progression. ",
  "**A** Last CASI against CPS_Global_pTau (fill = ADNC; dashed descriptive quadratic fit, reflecting the curvature found in Section 4.3) for the %d ACT donors: wide spread at every stage. ",
  "**B, C** Added-variable plots for the two prespecified anatomical pairings: memory rate against MEC local pTau, and executive rate against DFC local pTau, each residualized on CPS_Global_pTau, age, sex, education and last-visit timing. ",
  "**D** Each feature (per SD) added alone to phenotype ~ CPS_Global_pTau + age + sex + education + timing, for terminal CASI and the released memory and executive rates (SD units); filled = primary, open circle = + LATE-NC 2–3 and severe arteriolosclerosis, triangle = without the two donors with near-zero NeuN immunoreactivity, square = + quadratic global pTau (Section 4.3); OLS 95%% CI; %d–%d donors per estimate (MG4 and astrocyte require all three core ROIs). ",
  "A local-pTau coefficient describes regional involvement at a given global stage; global and local are collinear (ρ up to %.2f) and are not independent exposures. Cross-sectional associations; not causal; regional pathological distribution is not a measure of circuit function."),
  nrow(act), min(h_fits$n), max(h_fits$n), max(abs(h_collin[[3]])))
```

```{r}
#| label: fig3
#| fig-cap: !expr fig3_cap
#| out-width: "100%"

include_graphics(fig3_png)
```

---

## 6. First-pass analyses conditioned on the Gabitto MTG CPS (retained, not extended)

These results come from this notebook's first pass and are kept for continuity. All of them
condition on `SpecimenMetadata.csv`'s MTG-fitted CPS, and none is repeated with it here.

<details><summary>Terminal versus longitudinal resilience, and multiregional features beyond the MTG CPS</summary>

```{r}
#| label: first-pass

SLOPE_RHS_V1 <- "cps + age + sex + education + last_ybd"
v1_slopes <- map(c(memory = "slope_zmem0", executive = "slope_zexf0"),
                 ~ lm(as.formula(paste(.x, "~", SLOPE_RHS_V1)), data = act))
stopifnot("first-pass slope models use all 69 ACT donors" = all(map_int(v1_slopes, nobs) == 69L))
v1_adj <- tibble(terminal = act$adj_cognition, memory = residuals(v1_slopes$memory),
                 executive = residuals(v1_slopes$executive))
v1_rho <- c(mem = cor(v1_adj$terminal, v1_adj$memory, method = "spearman"),
            exf = cor(v1_adj$terminal, v1_adj$executive, method = "spearman"))

v1_feat <- act_f %>% mutate(at8_z = zstd(at8_donor))
v1_fit <- function(feature, extra = NULL) {
  d <- v1_feat %>% filter(!is.na(.data[[feature]])) %>% mutate(x = zstd(.data[[feature]]))
  base <- lm(as.formula(paste("casi ~ cps + age + sex + education + casi_interval",
                              if (length(extra)) paste("+", extra))), d)
  full <- update(base, . ~ . + x)
  stopifnot(nobs(base) == nobs(full))
  ci <- confint(full)["x", ]
  fmt_sci(coef(full)["x"], ci[1], ci[2], 1)
}
tibble(
  Result = c("Terminal residual vs pathology-adjusted memory decline (ρ)",
             "Terminal residual vs pathology-adjusted executive decline (ρ)",
             "Donor-wide direct AT8 beyond MTG CPS (CASI points per SD)",
             "Donor-wide MG4 beyond MTG CPS (CASI points per SD)",
             "Donor-wide MG4 beyond MTG CPS + donor-wide AT8"),
  Value = c(sprintf("%.2f", v1_rho["mem"]), sprintf("%.2f", v1_rho["exf"]),
            v1_fit("at8_donor"), v1_fit("mg4_donor"), v1_fit("mg4_donor", "at8_z"))
) %>% show_table(caption = "First-pass results, recomputed unchanged from notebook 04's CASI model and the Gabitto MTG CPS.")
```

First-pass conclusions, still standing:

- **Terminal vs longitudinal phenotypes.** Pathology-adjusted memory decline largely reproduces
  notebook 04's terminal residual, while executive decline is a partly separate axis.
- **Multiregional tau.** At fixed MTG CPS, donor-wide direct AT8 is associated with worse terminal
  CASI, and MG4's cognitive association weakens once that multiregional tau is included.
- **Matched pairs.** Matching on the three-region AT8 profile narrowed matched CASI differences,
  and the MG4/NeuN differences in CPS-matched pairs reversed under profile matching. Pairs were
  illustrative only.

Section 5 shows that the released global pTau stage absorbs much of the multiregional tau that
the MTG CPS missed.

</details>

---

## 7. Synthesis

```{r}
#| label: synthesis-numbers

syn <- list(
  b5 = sprintf("%d of %d", stg$b5_overlap, stg$b5_n),
  mid_iqr = sprintf("%.2f–%.2f", min(mid_iqr$iqr), max(mid_iqr$iqr)),
  mg4_global = mc_ci("global"), mg4_within = mc_ci("within"),
  neun_local = nf_ci("local_ptau"), mg4_neun = nf_ci("mg4_given_local"),
  mem_g = gl_ci("Memory", "global"), exf_l = gl_ci("Executive", "local"), exf_g = gl_ci("Executive", "global"),
  dfc_casi = hv_ci("lp_DFC", "Terminal CASI"), mec_exf = hv_ci("lp_MEC", "Executive rate")
)
```

**1. What global and regional progression explain about cognition.**

- **Global pTau stage** carries most of the pathology–cognition association:
  - terminal CASI `r pf_ci("Terminal CASI")` SD per 0.1;
  - global slope `r pf_ci("Global slope (CASI-IRT)")`;
  - memory rate `r pf_ci("Memory rate")`;
  - language rate `r pf_ci("Language rate")`.

  These are robust to co-pathology. Terminal CASI and memory rate are concave in global stage
  (flat with a CASI ceiling early, steeper late). The linear estimate is an average association.
- **Conventional staging is too coarse to stand in for progression.** Braak V and High ADNC each
  span most of the global range (`r syn$b5` Braak V donors overlap the Braak IV or VI range).
- **Matched regions add nothing for memory or language** beyond global stage (memory: global
  `r syn$mem_g`).

**2. What remains associated with cognition at comparable progression.**

- **Executive decline, the one regionally structured phenotype:**
  - faster with DFC involvement (`r hv_ci("lp_DFC", "Executive rate", "+ quadratic global pTau")`
    with curvature allowed);
  - slower with MEC-predominant tau (`r hv_ci("lp_MEC", "Executive rate", "+ quadratic global pTau")`).

  The direction holds across representations (direct AT8 in the first pass, released local pTau
  here) and under co-pathology and curvature. The primary linear intervals just reach zero, so the
  estimate is imprecise.
- **DFC–terminal CASI does not survive curvature** in global stage
  (`r hv_ci("lp_DFC", "Terminal CASI", "+ quadratic global pTau")`).
- **Donor-wide NeuN** is associated with slower memory decline, but it cannot be separated from the
  published severely affected phenotype (Section 5).
- **Donor-wide MG4 and astrocyte activation** add nothing clear.

**3. What MG4 and NeuN currently support biologically.**

- **MG4** is a donor-level microglial state that scales with global pTau stage
  (`r syn$mg4_global` SD/SD). It is not a regional one (within-donor `r syn$mg4_within`), and it
  carries no cognitive information beyond stage. It is a marker of disease stage, not an
  established protective or harmful factor.
- **NeuN immunoreactive density** is modestly lower where local pTau is higher within donors
  (`r syn$neun_local`, interval reaching zero). MG4 at comparable local pathology is at most weakly
  positive for it (`r syn$mg4_neun`).
- **The two donors with near-zero NeuN** show a severe neuronal phenotype: reduced NeuN
  immunoreactivity, and low neuronal representation in snRNA-seq from donors defined partly by
  poor nuclei QC. Absolute neuronal loss is neither established nor excluded. NeuN density measures
  NeuN-immunoreactive profiles, not neuron number or synaptic integrity.

**4. Which aspects of the cognitive measurements remain uncertain.**

- **Domain scores.** Documented as co-calibrated CFA/IRT scores with per-score SEs. Their
  reference mean/SD and the calibration release used here are not documented.
- **Slopes.** No accessible source documents the model, time scale, units or uncertainty of any
  released slope, so associations are reported in SD of the released variable.
- **Visuospatial.** Dominated by a single scored pattern (`r percent(vsp_diag$share_mode, 1)` of
  visits), empirically consistent with a ceiling on a small item set. The item content and battery
  history are undocumented.
- **Visit timing.** Whole-year ages, a thin final two years before death, and grouped curves that
  pool different donors limit any claim about terminal acceleration.

**5. Variables ready for morphology integration.**

| Block | Variable | Status |
|---|---|---|
| Global progression | `CPS_Global_pTau` (donor), with a quadratic term as sensitivity | ready; principal cognition covariate |
| Regional progression | `CPS_Local_pTau` in DFC and MEC, conditional on global stage; direct AT8 normal scores for tissue coupling | ready; DFC/MEC for executive decline |
| Staging context | Braak, ADNC | descriptive only |
| Cognition (ACT) | terminal CASI; `slope_zmem0`, `slope_zexf0`, `slope_zlan0`, `slope_zcasi_irt0` as released | ready, with provenance caveats; `slope_zvsp0` not recommended |
| Molecular | donor-wide MG4 competitive score (astrocyte activation complementary) | ready as a stage-linked covariate or modifier, not a regional exposure |
| Tissue | NeuN normal score, with the near-zero-immunoreactivity flag and the published severely affected flag for sensitivity | ready as a tissue reference; not an independent resilience correlate |
| Sensitivity covariates | LATE-NC 2–3, severe arteriolosclerosis | ready |

The first morphology tests these results support are regional morphometric deviation against
regional pTau deviation within donors, and DFC versus MEC morphology against executive decline at
fixed global stage.

<details><summary>Validation checks run during rendering</summary>

```{r}
#| label: validation

tibble::tribble(
  ~Check, ~Result,
  "Donor × ROI keys unique after QNP + molecular + released CPS join", "asserted",
  "Released CPS unique per donor × region; Global constant within donor; no missing value", "asserted",
  "Cognition set = 69 ACT donors; ADRC never modelled", "asserted",
  "Released slopes constant within donor; used as released", "asserted",
  "Nested models on identical donors (co-pathology; feature additions; global vs local comparisons)", "asserted",
  "Every bootstrap resample fitted", "asserted",
  sprintf("Donors without complete core QNP: %s", paste(dropped_qnp, collapse = ", ")), "reported, not imputed",
  "Severely affected flag used only as an exclusion sensitivity, never as a model term", "by construction",
  "Linear vs quadratic global pTau fits on identical donors", "asserted",
  "Every modal VSP_E visit carries one identical SE", "asserted"
) %>% show_table(caption = "Assertions. A failed assertion stops the render.")

input_files <- c(paths$qnp_2026, paths$donor_metadata, paths$specimen_index, paths$cognition,
                 analysis$analysis_table, omics$cps_by_region, omics$abundances, state_p$row_scores)
tibble(File = basename(input_files), md5 = unname(tools::md5sum(input_files))) %>%
  show_table(caption = "Inputs read by this notebook (read-only; nothing under data/ is written).")
```

</details>
